---
title: "Heat-Related Mortality and Lifespan Inequality"
subtitle: "Part 2: Life Tables and Inequality Analysis"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    keep_tex: false
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = FALSE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook (Part 2) constructs **cohort life tables** and analyzes **lifespan inequality** using mortality multipliers computed in Part 1.

**Prerequisites**: You must run **Part 1** first to generate:

- Mortality multipliers by age and year
- Configuration choices for city, scenario, and adaptation

**This notebook covers**:

1. Setting explicit analysis parameters
2. Validating age and year coverage before analysis
3. Loading baseline mortality projections (Eurostat EUROPOP2023)
4. Applying climate multipliers to baseline mortality rates
5. Constructing cohort life tables over the configured observation horizon
6. Computing life expectancy impacts
7. Analyzing lifespan inequality metrics (Gini, SD, IQR)
8. Exporting life table and inequality results
9. Outlining next steps

**Key concept**: Lifespan inequality measures how unequally distributed ages at death are. Climate change affects inequality through the **age pattern** of excess deaths, not just total mortality.

**Important approximation note**: Eurostat provides raw baseline mortality only through age 100 and year 2100, and Part 1 multipliers are raw through age 100 and year 2099. In this notebook, ages above 100 and years beyond the raw source horizon are extended using a **constant tail assumption** based on the last available age-specific values. The resulting life expectancy and lifespan-inequality metrics should therefore be interpreted as **approximation-based beyond the raw source coverage**.

---

# 1. Setup and Configuration

## 1.1 Load Required Libraries

```{r setup-libraries, include=FALSE}
library(data.table)
library(ggplot2)
library(scales)
library(eurostat)

cat("Libraries loaded successfully.\n")
```

---

## 1.2 Set Configuration Parameters

```{r set-config}
# Core analysis settings
city <- "Vienna"
city_code <- "AT001C"
scenario <- 3
scenario_label <- "SSP3-7.0"
adaptation_level <- 0.2  # 20% risk reduction by 2100

# Baseline mortality settings (Eurostat EUROPOP2023)
geo_code <- "AT"
sex_code <- "F"
sex_label <- if (sex_code == "F") "female" else if (sex_code == "M") "male" else stop("Unsupported sex_code: ", sex_code)
mortality_dataset_id <- "proj_23naasmr"
projection_code <- "BSL"
baseline_year_min <- 2022
baseline_year_max <- 2100
age_min <- 20
age_max_source <- 100           # maximum age available from the raw source
age_max <- 120                  # analysis tail age after constant age extension
extend_baseline_constant <- TRUE
extend_multiplier_constant <- TRUE

# Diagnostic summary settings
sample_years <- c(2022, 2050, 2080)
sample_ages <- c(30, 50, 70, 90)

# Cohort settings
cohort_birth_year <- 2002
cohort_start_age <- 20
cohort_terminal_age <- 120      # extended terminal age for approximate full death distribution
radix <- 100000
require_full_coverage <- TRUE

# File paths
mult_path <- sprintf(
  "results/mortality_multipliers_%s_ssp%d.csv",
  tolower(gsub(" ", "_", city)),
  scenario
)

mort_proj_path <- sprintf(
  "../data/mortality_projections_austria_%s.csv",
  sex_label
)

results_stub <- sprintf(
  "%s_ssp%d_%s_cohort%d",
  tolower(gsub(" ", "_", city)),
  scenario,
  sex_label,
  cohort_birth_year
)

cat("=== Configuration Parameters ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: %s\n", scenario_label))
cat(sprintf("  Adaptation by 2100: %.0f%%\n", adaptation_level * 100))
cat(sprintf("  Baseline geography: %s\n", geo_code))
cat(sprintf("  Baseline sex: %s (%s)\n", sex_code, sex_label))
cat(sprintf("  Eurostat dataset: %s | projection = %s\n", mortality_dataset_id, projection_code))
cat(sprintf("  Baseline source ages: %d–%d\n", age_min, age_max_source))
cat(sprintf("  Analysis ages after tail extension: %d–%d\n", age_min, age_max))
cat(sprintf("  Eurostat source years: %d–%d\n", baseline_year_min, baseline_year_max))
cat(sprintf("  Extend baseline constant outside source years: %s\n", ifelse(extend_baseline_constant, "TRUE", "FALSE")))
cat(sprintf("  Extend multipliers constant after last source year: %s\n", ifelse(extend_multiplier_constant, "TRUE", "FALSE")))
cat(sprintf("  Cohort birth year: %d\n", cohort_birth_year))
cat(sprintf("  Cohort observation ages: %d–%d\n", cohort_start_age, cohort_terminal_age))
cat(sprintf("  Cohort observation years: %d–%d\n",
            cohort_birth_year + cohort_start_age,
            cohort_birth_year + cohort_terminal_age))
cat(sprintf("  Initial cohort size (radix): %s\n", format(radix, big.mark = ",")))
cat(sprintf("  Require full coverage: %s\n", ifelse(require_full_coverage, "TRUE", "FALSE")))
```

---

## 1.3 Load Mortality Multipliers from Part 1

```{r load-multipliers}
extend_year_constant <- function(dt, target_year_min = NULL, target_year_max = NULL) {
  dt <- copy(dt)
  source_year_min <- min(dt$year)
  source_year_max <- max(dt$year)
  out <- copy(dt)

  if (!is.null(target_year_min) && target_year_min < source_year_min) {
    template_min <- dt[year == source_year_min]
    past_years <- target_year_min:(source_year_min - 1L)
    past_ext <- rbindlist(lapply(past_years, function(y) {
      tmp <- copy(template_min)
      tmp[, year := y]
      tmp
    }))
    out <- rbindlist(list(out, past_ext), use.names = TRUE)
  }

  if (!is.null(target_year_max) && target_year_max > source_year_max) {
    template_max <- dt[year == source_year_max]
    future_years <- (source_year_max + 1L):target_year_max
    future_ext <- rbindlist(lapply(future_years, function(y) {
      tmp <- copy(template_max)
      tmp[, year := y]
      tmp
    }))
    out <- rbindlist(list(out, future_ext), use.names = TRUE)
  }

  out <- unique(out, by = c("age", "year"))
  setorder(out, year, age)
  out
}

extend_age_constant <- function(dt, target_age_max, source_age_max = max(dt$age)) {
  dt <- copy(dt)
  out <- copy(dt)

  if (target_age_max > source_age_max) {
    template_max <- dt[age == source_age_max]
    future_ages <- (source_age_max + 1L):target_age_max
    age_ext <- rbindlist(lapply(future_ages, function(a) {
      tmp <- copy(template_max)
      tmp[, age := a]
      if ("agegroup" %in% names(tmp)) tmp[, agegroup := a]
      if ("decade" %in% names(tmp)) tmp[, decade := (a %/% 10L) * 10L]
      tmp
    }))
    out <- rbindlist(list(out, age_ext), use.names = TRUE, fill = TRUE)
  }

  out <- unique(out, by = c("age", "year"))
  setorder(out, year, age)
  out
}

if (!file.exists(mult_path)) {
  stop("Mortality multipliers file not found: ", mult_path,
       "\nRun Part 1 first using the same city/scenario configuration.")
}

multipliers <- fread(mult_path)

required_mult_cols <- c("age", "year", "multiplier_mean")
missing_mult_cols <- setdiff(required_mult_cols, names(multipliers))
if (length(missing_mult_cols) > 0L) {
  stop("Mortality multiplier file is missing required columns: ",
       paste(missing_mult_cols, collapse = ", "))
}

requested_year_min <- min(c(sample_years, cohort_birth_year + cohort_start_age))
requested_year_max <- max(c(sample_years, cohort_birth_year + cohort_terminal_age))
mult_source_year_min <- min(multipliers$year)
mult_source_year_max <- max(multipliers$year)
mult_source_age_max <- max(multipliers$age)

if (extend_multiplier_constant) {
  multipliers <- extend_year_constant(
    multipliers,
    target_year_min = requested_year_min,
    target_year_max = requested_year_max
  )
}

if (cohort_terminal_age > mult_source_age_max) {
  multipliers <- extend_age_constant(
    multipliers,
    target_age_max = cohort_terminal_age,
    source_age_max = mult_source_age_max
  )
}

cat(sprintf("\n=== Mortality Multipliers Loaded ===\n"))
cat(sprintf("  File: %s\n", mult_path))
cat(sprintf("  Records: %s\n", format(nrow(multipliers), big.mark = ",")))
cat(sprintf("  Source years: %d–%d\n", mult_source_year_min, mult_source_year_max))
cat(sprintf("  Analysis years after extension: %d–%d\n", min(multipliers$year), max(multipliers$year)))
cat(sprintf("  Source ages: %d–%d\n", min(multipliers$age[multipliers$year <= mult_source_year_max]), mult_source_age_max))
cat(sprintf("  Analysis ages after extension: %d–%d\n", min(multipliers$age), max(multipliers$age)))
cat(sprintf("  Mean multiplier: %.4f\n", mean(multipliers$multiplier_mean)))
cat(sprintf("  Multiplier range: [%.3f, %.3f]\n",
            min(multipliers$multiplier_mean), max(multipliers$multiplier_mean)))

head(multipliers, 10)
```

**Data structure**:

- `year`: Calendar year of the climate multiplier
- `age`: Single-year age
- `multiplier_mean`: Ensemble mean across GCMs
- `multiplier_sd`: Standard deviation across GCMs
- `multiplier_q05`, `multiplier_q50`, `multiplier_q95`: Quantiles for uncertainty

**Interpretation**: `multiplier > 1` indicates increased mortality risk relative to the historical reference used in Part 1.

---

## 1.4 Validate Analysis Horizon and Coverage

```{r validate-coverage}
mult_year_min <- min(multipliers$year)
mult_year_max <- max(multipliers$year)
mult_age_min <- min(multipliers$age)
mult_age_max <- max(multipliers$age)

cohort_start_year <- cohort_birth_year + cohort_start_age
cohort_terminal_year <- cohort_birth_year + cohort_terminal_age
requested_year_min <- min(c(sample_years, cohort_start_year))
requested_year_max <- max(c(sample_years, cohort_terminal_year))

max_terminal_age_baseline_raw <- min(age_max_source, baseline_year_max - cohort_birth_year)
max_terminal_age_multiplier_raw <- min(mult_source_age_max, mult_source_year_max - cohort_birth_year)
max_terminal_age_full <- if (extend_baseline_constant && extend_multiplier_constant) {
  age_max
} else if (extend_baseline_constant && !extend_multiplier_constant) {
  min(age_max, max_terminal_age_multiplier_raw)
} else if (!extend_baseline_constant && extend_multiplier_constant) {
  min(age_max_source, max_terminal_age_baseline_raw)
} else {
  min(max_terminal_age_baseline_raw, max_terminal_age_multiplier_raw)
}

if (cohort_start_age < age_min) {
  stop("cohort_start_age is below the baseline mortality age range.")
}
if (cohort_terminal_age > age_max) {
  stop("cohort_terminal_age exceeds the configured analysis age range.")
}
if (cohort_terminal_age < cohort_start_age) {
  stop("cohort_terminal_age must be greater than or equal to cohort_start_age.")
}
if (!extend_baseline_constant && cohort_start_year < baseline_year_min) {
  stop("The requested cohort starts in year ", cohort_start_year,
       ", but baseline mortality begins in ", baseline_year_min, ".")
}
if (!extend_baseline_constant && require_full_coverage && cohort_terminal_year > baseline_year_max) {
  stop("The requested cohort ends in year ", cohort_terminal_year,
       ", but baseline mortality ends in ", baseline_year_max,
       ". Lower cohort_terminal_age or change the cohort year.")
}
if (!extend_multiplier_constant && require_full_coverage && cohort_terminal_year > mult_source_year_max) {
  stop("The requested cohort ends in year ", cohort_terminal_year,
       ", but climate multipliers end in ", mult_source_year_max,
       ". Lower cohort_terminal_age or extend the multiplier series.")
}
if (cohort_start_age < mult_age_min || cohort_terminal_age > mult_age_max) {
  stop("The requested cohort age range lies outside the multiplier age range: ",
       mult_age_min, "–", mult_age_max)
}
if (any(sample_ages < age_min | sample_ages > age_max_source)) {
  stop("sample_ages must lie within the raw baseline mortality age range: ",
       age_min, "–", age_max_source)
}
if (cohort_terminal_age > max_terminal_age_full) {
  stop(
    "Requested cohort_terminal_age=", cohort_terminal_age,
    " exceeds full configured coverage for birth year ", cohort_birth_year,
    ". Maximum allowable age is ", max_terminal_age_full,
    " (raw baseline allows ", max_terminal_age_baseline_raw,
    ", raw multipliers allow ", max_terminal_age_multiplier_raw, ")."
  )
}

analysis_year_min <- requested_year_min
analysis_year_max <- requested_year_max

cat(sprintf("\n=== Coverage Validation Passed ===\n"))
cat(sprintf("  Multiplier source years: %d–%d\n", mult_source_year_min, mult_source_year_max))
cat(sprintf("  Multiplier analysis years: %d–%d\n", mult_year_min, mult_year_max))
cat(sprintf("  Multiplier ages after extension: %d–%d\n", mult_age_min, mult_age_max))
cat(sprintf("  Requested analysis years: %d–%d\n", analysis_year_min, analysis_year_max))
if (extend_baseline_constant && requested_year_max > baseline_year_max) {
  cat(sprintf("  Baseline mortality will be extended forward from %d to %d using the age-specific rates from %d.\n",
              baseline_year_max, requested_year_max, baseline_year_max))
}
if (cohort_terminal_age > age_max_source) {
  cat(sprintf("  Baseline mortality age tail will be extended from %d to %d using the age-specific rates at age %d.\n",
              age_max_source, cohort_terminal_age, age_max_source))
}
if (extend_multiplier_constant && requested_year_max > mult_source_year_max) {
  cat(sprintf("  Climate multipliers will be extended forward from %d to %d using the age-specific multipliers from %d.\n",
              mult_source_year_max, requested_year_max, mult_source_year_max))
}
if (cohort_terminal_age > mult_source_age_max) {
  cat(sprintf("  Climate multiplier age tail will be extended from %d to %d using the multipliers at age %d.\n",
              mult_source_age_max, cohort_terminal_age, mult_source_age_max))
}
cat(sprintf("  Cohort observation years: %d–%d\n", cohort_start_year, cohort_terminal_year))
cat(sprintf("  Cohort observation ages: %d–%d\n", cohort_start_age, cohort_terminal_age))
```

---

# 2. Baseline Mortality Data

To construct life tables, we need **baseline mortality rates** by age and year. These represent expected mortality in the absence of climate change impacts.

We use **Eurostat mortality projections** (EUROPOP2023), which provide age-specific mortality rates for European countries through 2100.

## 2.1 Load Eurostat Mortality Projections

```{r load-baseline-mortality}
if (file.exists(mort_proj_path)) {
  mort_proj <- fread(mort_proj_path)
} else {
  raw <- as.data.table(eurostat::get_eurostat(
    id = mortality_dataset_id,
    filters = list(
      geo = geo_code,
      projection = projection_code,
      sex = sex_code,
      sinceTimePeriod = baseline_year_min,
      untilTimePeriod = baseline_year_max
    ),
    time_format = "num",
    type = "code",
    stringsAsFactors = FALSE
  ))

  if (nrow(raw) == 0L) {
    stop("Eurostat returned zero rows for ", mortality_dataset_id,
         " / ", geo_code, " / ", projection_code, " / sex=", sex_code)
  }
  if (!identical(sort(unique(raw$projection)), projection_code)) {
    stop("Unexpected projection values returned: ", paste(sort(unique(raw$projection)), collapse = ", "))
  }
  if (!identical(sort(unique(raw$sex)), sex_code)) {
    stop("Unexpected sex values returned: ", paste(sort(unique(raw$sex)), collapse = ", "))
  }
  if (data.table::uniqueN(raw$unit) != 1L) {
    stop("Eurostat returned multiple units: ", paste(sort(unique(raw$unit)), collapse = ", "))
  }
  if (nrow(raw[, .N, by = .(age, time)][N > 1L]) > 0L) {
    stop("Eurostat returned duplicate age-year rows after filtering.")
  }

  raw[, age_num := fifelse(
    age == "Y_LT1", 0L,
    fifelse(age == "Y_GE100", 100L, as.integer(sub("^Y", "", age)))
  )]

  mort_proj <- raw[
    age_num >= age_min & age_num <= age_max_source,
    .(
      age = age_num,
      year = as.integer(time),
      mx = as.numeric(values)
    )
  ]

  if (nrow(mort_proj) == 0L) {
    stop("No mortality rows left after converting ages and filtering to ages ", age_min, "-", age_max_source, ".")
  }
  if (anyNA(mort_proj$age) || anyNA(mort_proj$year) || anyNA(mort_proj$mx)) {
    stop("Missing values detected in converted mortality data.")
  }
  if (nrow(mort_proj[, .N, by = .(age, year)][N > 1L]) > 0L) {
    stop("Converted mortality data still contain duplicate age-year rows.")
  }

  dir.create(dirname(mort_proj_path), recursive = TRUE, showWarnings = FALSE)
  fwrite(mort_proj, mort_proj_path)
}

if (extend_baseline_constant) {
  mort_proj <- extend_year_constant(
    mort_proj,
    target_year_min = requested_year_min,
    target_year_max = requested_year_max
  )
}

if (cohort_terminal_age > age_max_source) {
  mort_proj <- extend_age_constant(
    mort_proj,
    target_age_max = cohort_terminal_age,
    source_age_max = age_max_source
  )
}

mort_proj <- mort_proj[age >= age_min & age <= cohort_terminal_age]
mort_proj <- mort_proj[year >= requested_year_min & year <= requested_year_max]

if (nrow(mort_proj) == 0L) {
  stop("Baseline mortality table is empty after final filtering.")
}
if (nrow(mort_proj[, .N, by = .(age, year)][N > 1L]) > 0L) {
  stop("Baseline mortality table contains duplicate age-year rows.")
}
if (anyNA(mort_proj$mx)) {
  stop("Baseline mortality table contains missing mx values.")
}

setkey(mort_proj, year, age)

cat(sprintf("\n=== Baseline Mortality Data ===\n"))
cat(sprintf("  Source: Eurostat %s | projection = %s | sex = %s\n",
            mortality_dataset_id, projection_code, sex_code))
cat(sprintf("  Analysis ages: %d–%d\n", min(mort_proj$age), max(mort_proj$age)))
cat(sprintf("  Analysis years: %d–%d\n", min(mort_proj$year), max(mort_proj$year)))
cat(sprintf("  Records: %s\n", format(nrow(mort_proj), big.mark = ",")))

head(mort_proj, 10)
```

**Data structure**:

- `age`: Single-year ages in the analysis table after any tail extension
- `year`: Calendar years in the analysis table after any time extension
- `mx`: Central death rate

**Interpretation**: `mx = 0.001` means approximately 1 death per 1,000 person-years.

---

## 2.2 Baseline Mortality Summary

```{r baseline-mort-summary}
mort_sample <- mort_proj[year %in% sample_years & age %in% sample_ages]
mort_wide <- dcast(mort_sample, age ~ year, value.var = "mx")

cat("\n=== Baseline Mortality Rates (mx) ===\n")
print(mort_wide)

ggplot(mort_proj[year %in% sample_years], aes(x = age, y = mx, color = factor(year))) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_viridis_d(name = "Year") +
  labs(
    title = "Baseline Mortality Rates by Age",
    subtitle = "Eurostat EUROPOP2023 projections with analysis-tail extension where configured",
    x = "Age",
    y = "Central Death Rate (mx)",
    caption = "Log scale"
  ) +
  theme_minimal(base_size = 12)
```

---

# 3. Apply Climate Multipliers to Baseline Mortality

We now combine:

1. **Baseline mortality**: `mx_base[age, year]` from Eurostat
2. **Climate multipliers**: `M[age, year]` from Part 1

To obtain **climate-adjusted mortality**:

$$
mx_{\text{climate}}(a, y) = mx_{\text{baseline}}(a, y) \times M(a, y)
$$

where $M(a, y) > 1$ indicates increased mortality due to climate change.

---

## 3.1 Merge Multipliers with Baseline Mortality

```{r merge-multipliers}
mort_proj_overlap <- mort_proj[year >= analysis_year_min & year <= analysis_year_max]
mult_subset <- multipliers[
  age >= age_min & age <= age_max & year >= analysis_year_min & year <= analysis_year_max,
  .(age, year, multiplier = multiplier_mean)
]

mort_climate <- merge(
  mort_proj_overlap,
  mult_subset,
  by = c("age", "year"),
  all.x = TRUE
)

missing_mult <- mort_climate[is.na(multiplier)]
if (nrow(missing_mult) > 0L) {
  print(head(missing_mult, 10))
  stop("Missing climate multipliers remain after merge. Review age-year coverage before proceeding.")
}

mort_climate[, mx_climate := mx * multiplier]
setkey(mort_climate, year, age)

cat(sprintf("\n=== Climate-Adjusted Mortality Computed ===\n"))
cat(sprintf("  Records: %s\n", format(nrow(mort_climate), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(mort_climate$year), max(mort_climate$year)))
cat(sprintf("  Mean multiplier: %.4f\n", mean(mort_climate$multiplier)))
cat(sprintf("  Mean baseline mx: %.6f\n", mean(mort_climate$mx)))
cat(sprintf("  Mean climate mx: %.6f\n", mean(mort_climate$mx_climate)))

head(mort_climate, 15)
```

---

## 3.2 Comparison: Baseline vs Climate Mortality

```{r compare-mortality-summary}
mort_sample <- mort_proj[year %in% sample_years & age %in% sample_ages,
                         .(age, year, mx)]
mort_sample[, scenario := "Baseline"]

mort_sample_climate <- mort_climate[year %in% sample_years & age %in% sample_ages,
                                    .(age, year, mx = mx_climate)]
mort_sample_climate[, scenario := "Climate"]

# Compare both scenarios as subplots (side by side)
plot_data <- rbind(mort_sample, mort_sample_climate)

ggplot(plot_data, aes(x = age, y = mx, color = factor(year), group = year)) +
  geom_line(linewidth = 1) +
  facet_wrap(~scenario, ncol = 2) +
  scale_y_log10(labels = label_comma()) +
  scale_color_viridis_d(name = "Year") +
  labs(
    title = "Mortality Rates by Age",
    subtitle = "Eurostat EUROPOP2023 projections with analysis-tail extension where configured",
    x = "Age",
    y = "Central Death Rate (mx)",
    caption = "Log scale"
  ) +
  theme_minimal(base_size = 12)
```

We examine the **relative impact** of climate change on mortality by computing percentage change.

**Formula:**

$$
\text{Percentage change} = \frac{mx_{\text{climate}} - mx_{\text{baseline}}}{mx_{\text{baseline}}} \times 100\%
$$

**Interpretation:**

- Positive values: Climate **increases** mortality risk (worse outcomes)
- Negative values: Climate **decreases** mortality risk (e.g., reduced cold-related deaths)

```{r compare-mortality}
mort_climate[, pct_change := ((mx_climate - mx) / mx) * 100]

comparison <- mort_climate[year >= 2050, .(
  mean_mx_base    = mean(mx),
  mean_mx_climate = mean(mx_climate),
  mean_multiplier = mean(multiplier),
  mean_pct_change = mean(pct_change),
  max_pct_change  = max(pct_change),
  min_pct_change  = min(pct_change)
), by = year]

cat("\n=== Mortality Comparison: Baseline vs Climate (2050+) ===\n")
print(comparison[year %in% intersect(c(2050, 2070, 2090, 2099), comparison$year)])

ggplot(mort_climate[year %in% intersect(c(2050, 2070, 2099), unique(mort_climate$year))],
       aes(x = age, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_line(linewidth = 1.2, color = "#d62728") +
  geom_ribbon(aes(ymin = 0, ymax = pct_change, fill = pct_change > 0), alpha = 0.3) +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_manual(
    values = c("TRUE" = "#d62728", "FALSE" = "#2166ac"),
    labels = c("TRUE" = "Increased risk", "FALSE" = "Decreased risk"),
    name = ""
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = sprintf("Relative Mortality Impact of Climate Change — %s (%s)",
                    city, scenario_label),
    subtitle = sprintf("Adaptation: %.0f%% by 2100 | Percentage change in death rates",
                       adaptation_level * 100),
    x = "Age",
    y = "Change in Mortality Rate (%)",
    caption = "Positive values: increased mortality | Negative values: decreased mortality"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11)
  )
```

---

# 4. Cohort Life Table Construction

A **cohort life table** tracks a hypothetical birth cohort through time, applying age-specific mortality rates for each calendar year they live through.

For a cohort born in year $Y_0$:

- At age 20 (year $Y_0 + 20$), use mortality for year $Y_0 + 20$, age 20
- At age 21 (year $Y_0 + 21$), use mortality for year $Y_0 + 21$, age 21
- ... and so on through the **configured terminal age**

This differs from a **period life table**, which uses mortality rates from a single year.

---

## 4.1 Cohort Parameters

```{r cohort-setup}
cat(sprintf("\n=== Cohort Parameters ===\n"))
cat(sprintf("  Birth year: %d\n", cohort_birth_year))
cat(sprintf("  Start tracking at age %d (year %d)\n",
            cohort_start_age, cohort_start_year))
cat(sprintf("  Stop tracking at age %d (year %d)\n",
            cohort_terminal_age, cohort_terminal_year))
cat(sprintf("  Initial cohort size (radix): %s\n", format(radix, big.mark = ",")))
```

---

## 4.2 Build Cohort Life Table Function

```{r lifetable-function}
build_cohort_lifetable <- function(mort_data, birth_year, start_age, terminal_age, radix = 100000) {
  mort_data <- copy(mort_data)
  mort_data[, cohort_year := year - age]

  cohort_data <- mort_data[
    cohort_year == birth_year & age >= start_age & age <= terminal_age
  ]
  setorder(cohort_data, age)

  if (nrow(cohort_data) == 0L) {
    stop("No data found for this cohort")
  }

  expected_ages <- as.integer(start_age:terminal_age)
  expected_years <- as.integer(birth_year + expected_ages)

  if (nrow(cohort_data) != length(expected_ages) ||
      !all(cohort_data$age == expected_ages) ||
      !all(cohort_data$year == expected_years)) {

    expected_path <- data.table(age = expected_ages, year = expected_years)
    observed_path <- unique(cohort_data[, .(age, year)])
    missing_path <- expected_path[!observed_path, on = .(age, year)]
    extra_path <- observed_path[!expected_path, on = .(age, year)]

    if (nrow(missing_path) > 0L) {
      cat("Missing cohort age-year cells:\n")
      print(missing_path)
    }
    if (nrow(extra_path) > 0L) {
      cat("Unexpected extra cohort age-year cells:\n")
      print(extra_path)
    }

    stop(
      "Cohort year trajectory is incomplete for birth year ", birth_year,
      ". Requested ages ", start_age, "-", terminal_age,
      " require years ", min(expected_years), "-", max(expected_years), "."
    )
  }

  n <- nrow(cohort_data)

  cohort_data[, qx_base := mx / (1 + 0.5 * mx)]
  cohort_data[, qx_climate := mx_climate / (1 + 0.5 * mx_climate)]

  cohort_data[qx_base > 1, qx_base := 1]
  cohort_data[qx_climate > 1, qx_climate := 1]

  # Close the final extended single-year interval consistently.
  cohort_data[n, qx_base := 1]
  cohort_data[n, qx_climate := 1]

  cohort_data[, lx_base := as.numeric(NA)]
  cohort_data[, lx_climate := as.numeric(NA)]
  cohort_data[, dx_base := as.numeric(NA)]
  cohort_data[, dx_climate := as.numeric(NA)]

  cohort_data[1, lx_base := radix]
  cohort_data[1, lx_climate := radix]

  for (i in seq_len(n)) {
    cohort_data[i, dx_base := lx_base * qx_base]
    cohort_data[i, dx_climate := lx_climate * qx_climate]

    if (i < n) {
      cohort_data[i + 1, lx_base := cohort_data[i, lx_base - dx_base]]
      cohort_data[i + 1, lx_climate := cohort_data[i, lx_climate - dx_climate]]
    }
  }

  cohort_data[, lx_next_base := c(lx_base[-1], 0)]
  cohort_data[, lx_next_climate := c(lx_climate[-1], 0)]

  cohort_data[, Lx_base := (lx_base + lx_next_base) / 2]
  cohort_data[, Lx_climate := (lx_climate + lx_next_climate) / 2]

  # Final age is a closed one-year interval, not an open age interval.
  cohort_data[n, Lx_base := lx_base / 2]
  cohort_data[n, Lx_climate := lx_climate / 2]

  cohort_data[, Tx_base := rev(cumsum(rev(Lx_base)))]
  cohort_data[, Tx_climate := rev(cumsum(rev(Lx_climate)))]

  cohort_data[, ex_base := Tx_base / lx_base]
  cohort_data[, ex_climate := Tx_climate / lx_climate]

  cohort_data
}
```

---

## 4.3 Construct Baseline and Climate Life Tables

```{r build-lifetable}
lifetable <- build_cohort_lifetable(
  mort_climate,
  birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  terminal_age = cohort_terminal_age,
  radix = radix
)

cat(sprintf("\n=== Cohort Life Table Constructed ===\n"))
cat(sprintf("  Ages: %d–%d\n", min(lifetable$age), max(lifetable$age)))
cat(sprintf("  Years: %d–%d\n", min(lifetable$year), max(lifetable$year)))

lifetable_display <- lifetable[, .(
  age, year,
  mx_base = round(mx, 6),
  mx_climate = round(mx_climate, 6),
  multiplier = round(multiplier, 3),
  qx_base = round(qx_base, 5),
  qx_climate = round(qx_climate, 5),
  lx_base = round(lx_base, 0),
  lx_climate = round(lx_climate, 0),
  ex_base = round(ex_base, 2),
  ex_climate = round(ex_climate, 2)
)]

cat("\nLife Table (first 20 rows):\n")
print(head(lifetable_display, 20))

cat("\nLife Table (last 10 rows):\n")
print(tail(lifetable_display, 10))
```

---

# 5. Life Expectancy Analysis

## 5.1 Life Expectancy at Starting Age

```{r life-expectancy-start}
e_start_base <- lifetable[age == cohort_start_age, ex_base]
e_start_climate <- lifetable[age == cohort_start_age, ex_climate]

delta_e <- e_start_climate - e_start_base
pct_delta_e <- (delta_e / e_start_base) * 100

cat(sprintf("\n=== Life Expectancy at Age %d ===\n", cohort_start_age))
cat(sprintf("  Baseline (no climate impact):   %.2f years\n", e_start_base))
cat(sprintf("  Climate-adjusted (%s):          %.2f years\n", scenario_label, e_start_climate))
cat(sprintf("  Difference:                     %.2f years (%.2f%%)\n",
            delta_e, pct_delta_e))

if (delta_e < 0) {
  cat(sprintf("\n  → Climate change reduces life expectancy by %.1f months\n",
              abs(delta_e) * 12))
} else {
  cat(sprintf("\n  → Climate change increases life expectancy by %.1f months\n",
              delta_e * 12))
  cat("     (Note: This may occur if cold-related mortality reduction exceeds heat increase)\n")
}
```

---

## 5.2 Life Expectancy Difference by Age

```{r life-expectancy-by-age}
lifetable[, delta_ex := ex_climate - ex_base]

ggplot(lifetable, aes(x = age, y = delta_ex)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.8) +
  geom_line(linewidth = 1.2, color = "#d62728") +
  geom_ribbon(aes(ymin = 0, ymax = delta_ex, fill = delta_ex > 0), alpha = 0.3) +
  scale_fill_manual(
    values = c("TRUE" = "#2166ac", "FALSE" = "#d62728"),
    labels = c("TRUE" = "LE gain", "FALSE" = "LE loss"),
    name = "Climate impact"
  ) +
  labs(
    title = sprintf("Climate Impact on Remaining Life Expectancy — Cohort Born %d",
                    cohort_birth_year),
    subtitle = sprintf("%s, %s, %s, Adaptation %.0f%%",
                       city, scenario_label, sex_label, adaptation_level * 100),
    x = "Age",
    y = "Change in Remaining Life Expectancy (years)",
    caption = sprintf("Approximation includes constant tail extension to age %d", cohort_terminal_age)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

---

## 5.3 Survivorship Curves

```{r survivorship-curves}
surv_data <- lifetable[, .(
  age,
  Baseline = lx_base,
  Climate = lx_climate
)]

surv_long <- melt(surv_data, id.vars = "age",
                  variable.name = "Scenario", value.name = "lx")

ggplot(surv_long, aes(x = age, y = lx / radix, color = Scenario)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = sprintf("Survivorship Curve — Cohort Born %d", cohort_birth_year),
    subtitle = sprintf("Starting with %s people at age %d | extended through age %d",
                       format(radix, big.mark = ","), cohort_start_age, cohort_terminal_age),
    x = "Age",
    y = "Proportion Surviving",
    caption = sprintf("%s, %s, %s, Adaptation %.0f%%",
                      city, scenario_label, sex_label, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Interpretation**: The gap between curves represents excess mortality due to climate change under the configured tail-extension assumptions.

---

# 6. Export Life Table Results

```{r export-lifetable}
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

output_path <- sprintf("results/lifetable_%s.csv", results_stub)

lifetable_export <- lifetable[, .(
  age, year,
  mx, mx_climate, multiplier,
  qx_base, qx_climate,
  lx_base, lx_climate,
  dx_base, dx_climate,
  ex_base, ex_climate
)]

fwrite(lifetable_export, output_path)
cat(sprintf("\nSaved life table: %s\n", output_path))

summary_stats <- data.table(
  city = city,
  city_code = city_code,
  geo_code = geo_code,
  sex = sex_code,
  sex_label = sex_label,
  scenario = sprintf("SSP%d", scenario),
  scenario_label = scenario_label,
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  terminal_age = cohort_terminal_age,
  radix = radix,
  e_start_baseline = e_start_base,
  e_start_climate = e_start_climate,
  delta_e_years = delta_e,
  delta_e_months = delta_e * 12,
  pct_change_e = pct_delta_e,
  tail_extension_used = TRUE,
  age_max_source = age_max_source,
  baseline_year_max = baseline_year_max,
  multiplier_year_max_raw = mult_source_year_max
)

summary_path <- sprintf("results/lifetable_summary_%s.csv", results_stub)

fwrite(summary_stats, summary_path)
cat(sprintf("Saved life table summary: %s\n", summary_path))
```

---

# 7. Lifespan Inequality Analysis

Lifespan inequality (LI) measures the **dispersion** or **spread** of ages at death within a population. Unlike life expectancy (which is a mean), LI captures how equally or unequally distributed lifespans are.

**Key insight**: Two populations can have identical life expectancy but very different inequality:

- **Low LI**: Deaths concentrated around a narrow age range (e.g., most people die 75–85)
- **High LI**: Deaths spread across a wide age range (e.g., deaths from 60–100)

Climate change affects LI through the **age pattern** of excess deaths, not just their total number.

**Important interpretation note**: The full death distribution used below is approximate beyond the raw source horizon because ages above 100 and years beyond 2100/2099 are extended using constant tail assumptions.

---

## 7.1 Age-at-Death Distribution

The life table column `dx` represents the **distribution of deaths by age**. This distribution is the basis for the inequality metrics that follow, so it is useful to inspect it before computing Gini, SD, and IQR.

```{r age-at-death-distribution}
deaths_base <- lifetable[, .(age, dx = dx_base)]
deaths_climate <- lifetable[, .(age, dx = dx_climate)]

deaths_base[, prop := dx / sum(dx)]
deaths_climate[, prop := dx / sum(dx)]

cat(sprintf("\n=== Age-at-Death Distribution ===\n"))
cat(sprintf("Baseline deaths: %s\n", format(round(sum(deaths_base$dx)), big.mark = ",")))
cat(sprintf("Climate deaths: %s\n", format(round(sum(deaths_climate$dx)), big.mark = ",")))
cat("Total deaths are equal by construction in a closed cohort life table with the same radix.\n")

dist_combined <- rbind(
  deaths_base[, .(age, deaths = dx, scenario = "Baseline")],
  deaths_climate[, .(age, deaths = dx, scenario = "Climate")]
)

dist_combined[, prop := deaths / sum(deaths), by = scenario]

ggplot(dist_combined, aes(x = age, y = prop, fill = scenario)) +
  geom_area(alpha = 0.6, position = "identity") +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728"),
    name = "Scenario"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Age-at-Death Distribution",
    subtitle = sprintf("Cohort born %d — %s, %s, %s, Adaptation %.0f%%",
                       cohort_birth_year, city, scenario_label, sex_label, adaptation_level * 100),
    x = "Age at Death",
    y = "Proportion of Deaths",
    caption = sprintf("Includes constant tail extension beyond raw source coverage through age %d", cohort_terminal_age)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 7.2 Lifespan Inequality Metrics

### 7.2.1 Gini Coefficient

The **Gini coefficient** measures inequality on a 0–1 scale:

- **0**: Perfect equality (everyone dies at the same age)
- **1**: Maximum inequality (all deaths at extremes)

```{r compute-gini}
compute_gini <- function(ages, deaths) {
  valid <- deaths > 0
  ages <- ages[valid]
  deaths <- deaths[valid]

  n <- length(ages)
  total_deaths <- sum(deaths)
  mean_age <- sum(ages * deaths) / total_deaths

  numerator <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      numerator <- numerator + abs(ages[i] - ages[j]) * deaths[i] * deaths[j]
    }
  }

  denominator <- 2 * mean_age * total_deaths^2
  numerator / denominator
}

gini_base <- compute_gini(deaths_base$age, deaths_base$dx)
gini_climate <- compute_gini(deaths_climate$age, deaths_climate$dx)

delta_gini <- gini_climate - gini_base
pct_change_gini <- (delta_gini / gini_base) * 100

cat(sprintf("\n=== Gini Coefficient ===\n"))
cat(sprintf("Baseline:         %.4f\n", gini_base))
cat(sprintf("Climate-adjusted: %.4f\n", gini_climate))
cat(sprintf("Change:           %.4f (%.2f%%)\n", delta_gini, pct_change_gini))

if (delta_gini > 0) {
  cat("\n→ Climate change INCREASES lifespan inequality\n")
  cat("  (deaths become more unequally distributed)\n")
} else {
  cat("\n→ Climate change DECREASES lifespan inequality\n")
  cat("  (deaths become more equally distributed)\n")
}
```

---

### 7.2.2 Standard Deviation of Age at Death

The **standard deviation** (SD) measures dispersion in years.

```{r compute-sd}
compute_sd_age_at_death <- function(ages, deaths) {
  total_deaths <- sum(deaths)
  mean_age <- sum(ages * deaths) / total_deaths
  variance <- sum((ages - mean_age)^2 * deaths) / total_deaths
  sqrt(variance)
}

sd_base <- compute_sd_age_at_death(deaths_base$age, deaths_base$dx)
sd_climate <- compute_sd_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_sd <- sd_climate - sd_base
pct_change_sd <- (delta_sd / sd_base) * 100

cat(sprintf("\n=== Standard Deviation of Age at Death ===\n"))
cat(sprintf("Baseline:         %.2f years\n", sd_base))
cat(sprintf("Climate-adjusted: %.2f years\n", sd_climate))
cat(sprintf("Change:           %.2f years (%.2f%%)\n", delta_sd, pct_change_sd))
```

**Interpretation**: SD quantifies how "spread out" deaths are. Higher SD means ages at death are more dispersed.

---

### 7.2.3 Interquartile Range (IQR)

The **IQR** is the difference between the 75th and 25th percentile of age at death.

```{r compute-iqr}
compute_iqr_age_at_death <- function(ages, deaths) {
  cumsum_deaths <- cumsum(deaths)
  total_deaths <- sum(deaths)

  q25_idx <- which(cumsum_deaths >= 0.25 * total_deaths)[1]
  q25 <- ages[q25_idx]

  q75_idx <- which(cumsum_deaths >= 0.75 * total_deaths)[1]
  q75 <- ages[q75_idx]

  list(q25 = q25, q75 = q75, iqr = q75 - q25)
}

iqr_base <- compute_iqr_age_at_death(deaths_base$age, deaths_base$dx)
iqr_climate <- compute_iqr_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_iqr <- iqr_climate$iqr - iqr_base$iqr

cat(sprintf("\n=== Interquartile Range (IQR) ===\n"))
cat(sprintf("Baseline:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_base$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_base$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_base$iqr))

cat(sprintf("\nClimate-adjusted:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_climate$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_climate$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_climate$iqr))

cat(sprintf("\nChange in IQR: %.2f years\n", delta_iqr))
```

---

## 7.3 Age-Specific Climate Displacement

The plot below shows the **age pattern of death displacement** under climate change. This is useful descriptively for understanding where the distribution shifts, but it should not be interpreted using the invalid threshold heuristic that appeared in earlier drafts.

```{r plot-excess-deaths-by-age, fig.width=10, fig.height=6}
lifetable[, excess_deaths_age := dx_climate - dx_base]

ggplot(lifetable, aes(x = age, y = excess_deaths_age)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  geom_line(color = "#d62728", linewidth = 1) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Age-Specific Death Displacement Due to Climate Change",
    subtitle = sprintf("Cohort born %d — %s, %s, %s, Adaptation %.0f%%",
                       cohort_birth_year, city, scenario_label, sex_label, adaptation_level * 100),
    x = "Age",
    y = "Difference in Deaths (Climate - Baseline)",
    caption = sprintf("Approximation includes constant tail extension through age %d", cohort_terminal_age)
  ) +
  theme_minimal(base_size = 12)
```

**Interpretation**: Positive values indicate ages where the climate-adjusted cohort experiences more deaths than the baseline cohort; negative values indicate ages where deaths are displaced away from that age.

---

## 7.4 Lifespan Inequality Metrics Summary

```{r plot-li-metrics, fig.width=8, fig.height=6}
li_metrics <- data.table(
  Metric = c("Gini", "Gini", "SD", "SD", "IQR", "IQR"),
  Scenario = rep(c("Baseline", "Climate"), 3),
  Value = c(gini_base, gini_climate,
            sd_base, sd_climate,
            iqr_base$iqr, iqr_climate$iqr)
)

ggplot(li_metrics, aes(x = Scenario, y = Value, fill = Scenario)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = sprintf("%.3f", Value)),
            vjust = -0.5, size = 3.5) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = "Lifespan Inequality Metrics: Baseline vs Climate",
    subtitle = sprintf("%s, %s, %s, Adaptation %.0f%%",
                       city, scenario_label, sex_label, adaptation_level * 100),
    y = "Metric Value",
    x = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 7.5 Summary Table

```{r li-summary-table}
li_summary <- data.table(
  Metric = c("Life Expectancy at Start", "Gini Coefficient",
             "SD of Age at Death", "IQR of Age at Death"),
  Baseline = c(
    sprintf("%.2f years", e_start_base),
    sprintf("%.4f", gini_base),
    sprintf("%.2f years", sd_base),
    sprintf("%.1f years", iqr_base$iqr)
  ),
  Climate = c(
    sprintf("%.2f years", e_start_climate),
    sprintf("%.4f", gini_climate),
    sprintf("%.2f years", sd_climate),
    sprintf("%.1f years", iqr_climate$iqr)
  ),
  Change = c(
    sprintf("%.2f years (%.2f%%)", delta_e, pct_delta_e),
    sprintf("%.4f (%.2f%%)", delta_gini, pct_change_gini),
    sprintf("%.2f years (%.2f%%)", delta_sd, pct_change_sd),
    sprintf("%.1f years", delta_iqr)
  )
)

cat("\n=== Lifespan Inequality Summary ===\n")
print(li_summary)
```

---

# 8. Export Lifespan Inequality Results

```{r export-li-results}
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

li_results <- data.table(
  city = city,
  city_code = city_code,
  geo_code = geo_code,
  sex = sex_code,
  sex_label = sex_label,
  scenario = sprintf("SSP%d", scenario),
  scenario_label = scenario_label,
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  terminal_age = cohort_terminal_age,
  tail_extension_used = TRUE,
  age_max_source = age_max_source,
  baseline_year_max = baseline_year_max,
  multiplier_year_max_raw = mult_source_year_max,
  e_start_baseline = e_start_base,
  e_start_climate = e_start_climate,
  delta_e = delta_e,
  pct_change_e = pct_delta_e,
  gini_baseline = gini_base,
  gini_climate = gini_climate,
  delta_gini = delta_gini,
  pct_change_gini = pct_change_gini,
  sd_baseline = sd_base,
  sd_climate = sd_climate,
  delta_sd = delta_sd,
  pct_change_sd = pct_change_sd,
  iqr_baseline = iqr_base$iqr,
  iqr_climate = iqr_climate$iqr,
  delta_iqr = delta_iqr
)

li_path <- sprintf("results/lifespan_inequality_%s.csv", results_stub)
fwrite(li_results, li_path)
cat(sprintf("\nSaved: %s\n", li_path))

excess_by_age <- lifetable[, .(
  age, year,
  dx_baseline = dx_base,
  dx_climate = dx_climate,
  excess_deaths = excess_deaths_age,
  multiplier
)]

excess_path <- sprintf("results/excess_deaths_by_age_%s.csv", results_stub)
fwrite(excess_by_age, excess_path)
cat(sprintf("Saved: %s\n", excess_path))
```

---

## 8.1 Key Findings Interpretation

```{r interpretation}
cat("\n=== KEY FINDINGS ===\n\n")

cat(sprintf("For the %s cohort born in %d in %s under %s:\n\n",
            sex_label, cohort_birth_year, city, scenario_label))

cat("1. LIFE EXPECTANCY\n")
if (delta_e < 0) {
  cat(sprintf("   → Climate change REDUCES life expectancy by %.2f years (%.2f%%)\n",
              abs(delta_e), abs(pct_delta_e)))
  cat(sprintf("   → Equivalent to %.1f months lost\n", abs(delta_e) * 12))
} else {
  cat(sprintf("   → Climate change INCREASES life expectancy by %.2f years (%.2f%%)\n",
              delta_e, pct_delta_e))
  cat("   → (Likely due to reduced cold-related mortality)\n")
}

cat("\n2. LIFESPAN INEQUALITY\n")
if (delta_gini > 0) {
  cat(sprintf("   → Climate change INCREASES inequality (Gini: +%.4f, +%.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE unequally distributed\n")
} else {
  cat(sprintf("   → Climate change DECREASES inequality (Gini: %.4f, %.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE equally distributed\n")
}

cat(sprintf("   → Standard deviation changes by %.2f years\n", delta_sd))
cat(sprintf("   → IQR changes by %.1f years\n", delta_iqr))

cat("\n3. INTERPRETATION\n")
cat("   → The inequality metrics above are approximation-based because ages above 100 and years beyond the raw source horizon are extended using constant tail assumptions.\n")
cat("   → The age-specific displacement plot is descriptive and shows where deaths shift across ages under climate change.\n")
cat("   → Total deaths in the full closed cohort are equal by construction, so the main substantive quantities are changes in life expectancy and changes in the shape of the age-at-death distribution.\n")

cat("\n4. POLICY IMPLICATION\n")
if (delta_gini > 0) {
  cat("   → Climate change may slightly widen dispersion in ages at death under this approximation.\n")
  cat("   → Policies that reduce mortality at more vulnerable ages could limit both longevity losses and inequality increases.\n")
} else {
  cat("   → Climate change appears to slightly compress the age-at-death distribution under this approximation.\n")
  cat("   → Even with this compression, mean life expectancy still declines, so absolute mortality reduction remains a priority.\n")
}

cat(sprintf("\n   → With %.0f%% adaptation by 2100, these impacts are partially mitigated\n",
            adaptation_level * 100))
cat(sprintf("   → This notebook follows the cohort through age %d under the configured tail-extension assumptions\n",
            cohort_terminal_age))
```

---

# 9. Next Steps

This analysis provides the methodological foundation for comprehensive multi-scenario studies:

1. **Multiple scenarios**: Run for all SSPs (1, 2, 3, 5) and adaptation levels (0%, 25%, 50%, 75%, 100%)
2. **Multiple cities**: Extend to 10–20 European cities (Vienna, Madrid, Stockholm, Athens, etc.)
3. **Multiple cohorts**: Analyze cohorts born in different decades (2000, 2020, 2040, etc.)
4. **Geographic decomposition**: Compute within-city vs between-city inequality components
5. **Temporal trends**: Track how LI evolves year-by-year
6. **Heat vs cold decomposition**: Separate LI impacts of heat-related and cold-related mortality
7. **Tail sensitivity analysis**: Compare constant tail extension with alternative old-age extrapolation assumptions

**To run this analysis for a different configuration**:
1. Modify the explicit parameters near the top of the notebook
2. Run Part 1 to generate new multipliers if needed
3. Re-run Part 2 with the same city, scenario, sex, and age/year settings

---

# Session Information

```{r session-info}
sessionInfo()
```