---
title: "Climate Change Impact on Mortality and Lifespan Inequality"
subtitle: "Single-City Deep Dive Analysis"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# 1. Configuration

**Analysis Parameters:** Select the target city for single-city deep dive analysis.

```{r config}
# ============================================================================
# USER CONFIGURATION: Select Target City
# ============================================================================

target_city_code <- "AT001C"  # Vienna, Austria
target_city_name <- "Vienna"  # Display name

# Analysis parameters (match pipeline settings)
cohort_birth_year <- 2020
cohort_start_age <- 20
radix <- 100000
```

---

# 2. Data Loading

This analysis uses **pre-computed results** from the full pipeline run. No heavy computation is performed in this notebook.

```{r load-packages}
library(data.table)
library(ggplot2)
library(scales)
library(viridis)
library(gridExtra)
```

```{r load-data}
# Load main results summary
results <- fread("../results/lifespan_inequality_summary.csv")

cat(sprintf("Total scenarios loaded: %s\n", format(nrow(results), big.mark = ",")))
cat(sprintf("Cities: %d\n", length(unique(results$city_code))))
cat(sprintf("SSPs: %s\n", paste(unique(results$ssp), collapse = ", ")))
cat(sprintf("Adaptation levels: %d\n", length(unique(results$adaptation))))

# Filter to target city
city_results <- results[city_code == target_city_code]

if (nrow(city_results) == 0) {
  stop(sprintf("No results found for city: %s", target_city_code))
}

cat(sprintf("\n%s: %d scenarios available\n", target_city_name, nrow(city_results)))
```

---

# 3. Methodological Context

## 3.1 Overview of the Analytical Pipeline

This analysis is based on a multi-stage computational pipeline:

**Stage 1: Exposure-Response Functions (ERFs)**
- City-specific temperature-mortality relationships from Masselot et al.
- Age-stratified coefficients (5 age groups interpolated to single-year ages)
- Distributed lag non-linear model (DLNM) with B-spline basis

**Stage 2: Climate Projections**
- Daily temperature projections from multiple General Circulation Models (GCMs)
- Shared Socioeconomic Pathways (SSP2-4.5, SSP3-7.0)
- Historical baseline period for comparison

**Stage 3: Mortality Multipliers**
- Age-specific mortality risk ratios computed for each year (2015-2100)
- Relative to historical baseline (typically 1991-2020)
- Adaptation scenarios: linear reduction of excess risk from 2030-2100

**Stage 4: Cohort Life Tables**
- Cohort born in 2020, tracked from age 20 onwards
- Mortality rates adjusted by climate multipliers
- Standard life table construction (mx → qx → lx → dx → ex)

**Stage 5: Lifespan Inequality Metrics**
- Gini coefficient, standard deviation, interquartile range
- Threshold age analysis (premature vs late deaths)
- Comparison of baseline vs climate-adjusted scenarios

---

## 3.2 Adaptation Modeling

Adaptation is modeled as a **linear reduction in excess heat/cold-related mortality risk**:

$$
\text{Adaptation}(y) = \begin{cases}
0 & \text{if } y < 2030 \\
\alpha \cdot \frac{y - 2030}{2100 - 2030} & \text{if } 2030 \le y \le 2100 \\
\alpha & \text{if } y > 2100
\end{cases}
$$

where $\alpha \in [0,1]$ is the adaptation level achieved by 2100.

The **adapted relative risk** is:

$$
RR_{\text{adapted}}(T,y) = 1 + [RR(T) - 1] \times [1 - \text{Adaptation}(y)]
$$

This formulation:
- Preserves baseline mortality at minimum mortality temperature (MMT)
- Reduces excess risk proportionally to adaptation level
- Implements gradual adaptation (not instant)
- Allows comparison across scenarios (0%, 50%, 100% adaptation by 2100)

**Key assumption:** Adaptation affects all age groups equally. In reality, adaptation effectiveness may vary by age, socioeconomic status, and geographic location.

---

# 4. Life Expectancy Analysis

Life expectancy at age 20 (e₂₀) is the **expected remaining years of life** for a 20-year-old in the cohort. It is computed from the life table as:

$$
e_{20} = \frac{\sum_{x=20}^{\omega} L_x}{l_{20}}
$$

where $L_x$ is person-years lived at age $x$, and $l_{20}$ is the number of survivors at age 20.

**Climate impact on LE** is measured as:

$$
\Delta e_{20} = e_{20}^{\text{climate}} - e_{20}^{\text{baseline}}
$$

**Interpretation:**
- $\Delta e_{20} < 0$: Climate change **reduces** life expectancy (mortality burden exceeds benefits)
- $\Delta e_{20} > 0$: Climate change **increases** life expectancy (cold-related mortality reduction exceeds heat increase)
- Magnitude: 1 year ≈ 12 months of expected life gained/lost

---

## 4.1 Summary Statistics

```{r summary-table}
# Overview table by SSP and adaptation
summary_stats <- city_results[, .(
  e20_baseline = mean(e20_base),
  e20_climate = mean(e20_climate),
  delta_e20_years = mean(delta_e20),
  delta_e20_months = mean(delta_e20) * 12,
  pct_change_e20 = mean(pct_delta_e20)
), by = .(ssp, adaptation)]

cat("\n=== Life Expectancy Summary ===\n")
print(summary_stats[, .(
  ssp,
  adaptation = sprintf("%.0f%%", adaptation * 100),
  e20_climate = sprintf("%.2f", e20_climate),
  delta_months = sprintf("%.1f", delta_e20_months),
  pct_change = sprintf("%.2f%%", pct_change_e20)
)])
```

---

## 4.2 Life Expectancy Response to Adaptation

This plot shows how life expectancy changes under different adaptation scenarios. The **zero line** marks the threshold where climate becomes neutral (neither beneficial nor harmful).

**What to look for:**
- **Slope**: Steeper = greater adaptation effectiveness
- **Zero-crossing**: Adaptation level where climate becomes neutral
- **SSP divergence**: How much more adaptation is needed under higher emissions (SSP3 vs SSP2)
- **Curvature**: Linear relationship suggests proportional returns; non-linear suggests diminishing or accelerating returns

**Possible interpretations:**
- If curves cross zero, some adaptation level completely offsets climate mortality
- If curves remain negative, even full adaptation cannot eliminate climate mortality burden
- If curves are positive, cold-related mortality reduction exceeds heat-related mortality increase

```{r plot-le-adaptation, fig.height=6}
# Prepare data
le_data <- city_results[, .(
  adaptation = adaptation * 100,
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  delta_e20 = delta_e20,
  delta_e20_months = delta_e20 * 12
)]

ggplot(le_data, aes(x = adaptation, y = delta_e20_months, color = ssp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  scale_color_manual(
    values = c("SSP2-4.5" = "#2166ac", "SSP3-7.0" = "#b2182b"),
    name = "Scenario"
  ) +
  labs(
    title = sprintf("Life Expectancy Response to Adaptation — %s", target_city_name),
    subtitle = sprintf("Cohort born %d, tracked from age %d", 
                       cohort_birth_year, cohort_start_age),
    x = "Adaptation Level by 2100 (%)",
    y = "Change in Life Expectancy (months)",
    caption = "Positive values indicate longevity gains; negative values indicate losses.\nZero line marks adaptation threshold where climate becomes neutral."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

---

## 4.3 Reversal Threshold Analysis

The **reversal threshold** is the minimum adaptation level required to neutralize climate's impact on life expectancy (i.e., where $\Delta e_{20} = 0$).

**Method:** Linear interpolation between the two adaptation levels where the sign of $\Delta e_{20}$ changes.

**Policy relevance:** This threshold represents the **minimum viable adaptation target** to maintain current longevity trajectories under climate change.

```{r reversal-threshold}
# Find reversal threshold for each SSP
reversal_thresholds <- city_results[, {
  if (any(delta_e20 < 0) & any(delta_e20 > 0)) {
    # Find sign change
    idx_neg <- which(delta_e20 < 0)
    idx_pos <- which(delta_e20 > 0)
    
    if (length(idx_neg) > 0 & length(idx_pos) > 0) {
      # Get closest points around zero
      below <- .SD[delta_e20 < 0][which.max(delta_e20)]
      above <- .SD[delta_e20 > 0][which.min(delta_e20)]
      
      # Linear interpolation
      x1 <- below$adaptation
      x2 <- above$adaptation
      y1 <- below$delta_e20
      y2 <- above$delta_e20
      
      threshold <- x1 + (0 - y1) * (x2 - x1) / (y2 - y1)
      
      .(reversal_threshold = threshold * 100,
        below_adapt = x1 * 100,
        below_delta = y1 * 12,
        above_adapt = x2 * 100,
        above_delta = y2 * 12)
    } else {
      .(reversal_threshold = NA_real_)
    }
  } else if (all(delta_e20 >= 0)) {
    .(reversal_threshold = 0, note = "Always beneficial")
  } else {
    .(reversal_threshold = 100, note = "Never neutralized")
  }
}, by = ssp]

cat("\n=== Reversal Threshold ===\n")
print(reversal_thresholds)

if (nrow(reversal_thresholds[!is.na(reversal_threshold)]) > 0) {
  cat("\nInterpretation:\n")
  for (i in 1:nrow(reversal_thresholds)) {
    s <- reversal_thresholds[i, ssp]
    t <- reversal_thresholds[i, reversal_threshold]
    if (!is.na(t)) {
      cat(sprintf("  SSP%d requires %.1f%% adaptation by 2100 to neutralize climate mortality\n", s, t))
    }
  }
}
```

**Discussion prompts:**
- Is this threshold achievable within realistic policy timelines?
- How sensitive is this threshold to uncertainty in climate projections?
- Does the threshold differ substantially between SSP2 and SSP3?

---

# 5. Lifespan Inequality Analysis

## 5.1 Conceptual Framework: What is Lifespan Inequality?

**Lifespan inequality (LI)** measures the **dispersion** or **heterogeneity** in ages at death within a population. Unlike life expectancy (a measure of central tendency), LI quantifies how equally or unequally lifespans are distributed.

**Key insight:** Two populations can have identical life expectancy but very different inequality:

- **Low LI:** Deaths concentrated around a narrow age range → "rectangularization" of survival curve
- **High LI:** Deaths spread across wide age range → high unpredictability of lifespan

**Why it matters for climate change:**
Climate affects mortality across different ages. The **pattern of excess deaths** (which ages are most affected) determines whether inequality increases or decreases, independently of the total mortality burden.

---

## 5.2 The Gini Coefficient for Ages at Death

The **Gini coefficient** ($G$) is the most widely used measure of lifespan inequality in demography. Originally developed by Corrado Gini (1912) for income inequality, it was adapted for mortality analysis in the 1980s.

**Definition:** For ages at death $x_i$ with associated deaths $d_i$:

$$
G = \frac{\sum_{i=1}^{n} \sum_{j=1}^{n} |x_i - x_j| \cdot d_i \cdot d_j}{2 \cdot \bar{x} \cdot \left(\sum_{i=1}^{n} d_i\right)^2}
$$

where $\bar{x}$ is the mean age at death.

**Interpretation:**
- $G = 0$: Perfect equality (everyone dies at exactly the same age)
- $G = 1$: Maximum inequality (all deaths at extremes)
- Typical values for high-income countries: $G \approx 0.06 - 0.12$

**Advantages over coefficient of variation (CV):**
- Scale-invariant (unaffected by shifts in mean age)
- Bounded [0,1] (easier interpretation)
- Captures entire distribution (not just variance)
- Widely used in demographic literature (comparability)

**Climate change and Gini:**
- $\Delta G > 0$: Climate **increases** lifespan inequality (deaths become more dispersed)
- $\Delta G < 0$: Climate **decreases** lifespan inequality (deaths become more concentrated)

**Reference:** Vaupel, J. W., Zhang, Z., & van Raalte, A. A. (2011). "Life expectancy and disparity: an international comparison of life table data." *BMJ Open*, 1(1), e000128.

---

## 5.3 Additional Inequality Metrics

**Standard Deviation (SD):**
$$
SD = \sqrt{\frac{\sum_{i=1}^{n} (x_i - \bar{x})^2 \cdot d_i}{\sum_{i=1}^{n} d_i}}
$$

- **Interpretation:** Average deviation from mean age at death (in years)
- **Advantage:** Intuitive scale (years)
- **Disadvantage:** Sensitive to outliers, not bounded

**Interquartile Range (IQR):**
$$
IQR = Q_{75} - Q_{25}
$$

where $Q_p$ is the $p$-th percentile of the age-at-death distribution.

- **Interpretation:** Spread of the "middle 50%" of deaths
- **Advantage:** Robust to outliers
- **Disadvantage:** Ignores tails of distribution

**Why use multiple metrics?**
Different metrics capture different aspects of inequality. Gini is comprehensive, SD is intuitive, IQR is robust. Agreement across metrics strengthens conclusions; disagreement highlights nuanced distributional changes.

---

# 6. The Life Expectancy-Inequality Paradox

A critical demographic insight: **life expectancy and lifespan inequality do not always move together**.

## 6.1 Theoretical Framework

**Historical pattern (pre-climate):** As life expectancy increased over the 20th century, lifespan inequality generally decreased (mortality compression). This suggests a "natural" negative correlation.

**Climate change disrupts this pattern:** Depending on the age-distribution of excess deaths, climate can:

1. **Reduce LE AND increase LI** (worst case): Extra deaths concentrated at younger-old ages
2. **Reduce LE but decrease LI** (compression): Extra deaths concentrated at very old ages
3. **Increase LE but increase LI** (paradox): Fewer cold deaths at old ages, but more heat deaths at young-old ages
4. **Increase LE AND decrease LI** (best case): Reduced cold mortality compresses distribution

**The threshold age (see Section 7) determines which scenario occurs.**

---

## 6.2 Scatter Plot: LE vs LI Changes

This plot visualizes the **joint distribution** of life expectancy and inequality changes across scenarios.

**Quadrant interpretation:**
- **Top-left** (ΔLE < 0, ΔGini > 0): "Worst case" — shorter lives AND more unequal
- **Top-right** (ΔLE > 0, ΔGini > 0): "Paradox" — longer lives but MORE unequal
- **Bottom-left** (ΔLE < 0, ΔGini < 0): "Compression" — shorter lives but more equal
- **Bottom-right** (ΔLE > 0, ΔGini < 0): "Win-win" — longer lives AND more equal

**What to look for:**
- Which quadrant dominates?
- Does SSP3 push scenarios into worse quadrants than SSP2?
- Is there a linear LE-LI relationship, or substantial scatter?

```{r plot-le-inequality-paradox, fig.height=7}
ggplot(city_results, 
       aes(x = delta_e20, y = delta_gini, color = factor(ssp))) +
  geom_point(aes(size = adaptation), alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(
    values = c("2" = "#2166ac", "3" = "#b2182b"),
    labels = c("SSP2-4.5", "SSP3-7.0"),
    name = "Scenario"
  ) +
  scale_size_continuous(
    name = "Adaptation",
    labels = function(x) sprintf("%.0f%%", x * 100),
    range = c(2, 8)
  ) +
  annotate("text", x = min(city_results$delta_e20) * 0.8, 
           y = max(city_results$delta_gini) * 0.8,
           label = "Worse LE\nMore inequality", 
           color = "gray30", hjust = 0, vjust = 1, size = 3.5) +
  annotate("text", x = max(city_results$delta_e20) * 0.8, 
           y = max(city_results$delta_gini) * 0.8,
           label = "Better LE\nMore inequality\n(PARADOX)", 
           color = "gray30", hjust = 1, vjust = 1, size = 3.5) +
  annotate("text", x = min(city_results$delta_e20) * 0.8, 
           y = min(city_results$delta_gini) * 0.8,
           label = "Worse LE\nLess inequality\n(compression)", 
           color = "gray30", hjust = 0, vjust = 0, size = 3.5) +
  annotate("text", x = max(city_results$delta_e20) * 0.8, 
           y = min(city_results$delta_gini) * 0.8,
           label = "Better LE\nLess inequality", 
           color = "gray30", hjust = 1, vjust = 0, size = 3.5) +
  labs(
    title = sprintf("Life Expectancy-Inequality Paradox — %s", target_city_name),
    subtitle = "Each point represents one SSP × adaptation scenario",
    x = "Change in Life Expectancy at Age 20 (years)",
    y = "Change in Gini Coefficient",
    caption = "Point size indicates adaptation level. Quadrants show different LE-LI regimes."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

**Discussion prompts:**
- Does this city show the "paradox" (top-right quadrant)?
- How does adaptation shift the LE-LI relationship?
- What might explain scenarios in the top-right quadrant (better LE, worse LI)?

---

## 6.3 Gini Coefficient Trajectories

Direct comparison of Gini changes between SSP scenarios.

**What to look for:**
- **Absolute magnitude:** Is ΔGini large or small? (Context: ±0.001 is ~1-2% change)
- **Direction:** Positive (increasing inequality) or negative (decreasing)?
- **SSP divergence:** How much worse is SSP3 than SSP2?
- **Adaptation gradient:** Does higher adaptation consistently reduce inequality impacts?

```{r plot-gini-trajectories, fig.height=6}
gini_data <- city_results[, .(
  adaptation = adaptation * 100,
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  delta_gini = delta_gini,
  pct_change_gini = pct_change_gini
)]

ggplot(gini_data, aes(x = adaptation, y = delta_gini, color = ssp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  scale_color_manual(
    values = c("SSP2-4.5" = "#2166ac", "SSP3-7.0" = "#b2182b"),
    name = "Scenario"
  ) +
  labs(
    title = sprintf("Lifespan Inequality Response to Adaptation — %s", target_city_name),
    subtitle = "Change in Gini coefficient across adaptation scenarios",
    x = "Adaptation Level by 2100 (%)",
    y = "Change in Gini Coefficient",
    caption = "Positive values indicate increasing inequality; negative values indicate compression.\nZero line marks no change in inequality."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Interpretation notes:**
- If curves are positive and increasing: climate worsens inequality, and adaptation helps but doesn't eliminate it
- If curves cross zero: some adaptation level neutralizes inequality impacts
- If curves are negative: climate compresses lifespan distribution (counterintuitive but possible if cold mortality dominates)

---

# 7. Threshold Age Analysis

## 7.1 Theoretical Foundation

**Key demographic insight (Vaupel et al., 2011):** There exists a **critical age** — the **threshold age** — that determines whether additional deaths increase or decrease lifespan inequality.

**Mathematical intuition:**
- Deaths **below** the threshold pull the distribution leftward → **increase** inequality
- Deaths **above** the threshold compress the right tail → **decrease** inequality

**Approximation:** The threshold age is roughly where **remaining life expectancy equals the population's life expectancy at the starting age**:

$$
\text{Threshold age} \approx \text{age } x \text{ where } e_x \approx e_{20}
$$

**For our cohorts starting at age 20:**
- If threshold ≈ 80, deaths at age 70 increase inequality (premature)
- Deaths at age 90 decrease inequality (late, expected)

**Climate change relevance:**
If climate predominantly causes deaths **below the threshold**, it will increase lifespan inequality, regardless of whether it increases or decreases life expectancy.

**Reference:** Vaupel, J. W., Zhang, Z., & van Raalte, A. A. (2011). "Life expectancy and disparity: an international comparison of life table data." *BMJ Open*, 1(1), e000128.

---

## 7.2 Excess Deaths Classification

This analysis classifies excess deaths (climate vs baseline) by their position relative to the threshold age.

**Data interpretation:**
- **`pct_below`**: % of excess deaths occurring **before** threshold age (increases inequality)
- **`pct_above`**: % of excess deaths occurring **after** threshold age (decreases inequality)

**Note on signs:** 
- If total excess deaths < 0 (climate reduces deaths overall), the percentages show which ages benefit most
- If total excess deaths > 0 (climate increases deaths), percentages show which ages are harmed most

```{r threshold-analysis}
# Summary of threshold classification
threshold_summary <- city_results[, .(
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  adaptation = adaptation * 100,
  threshold_age = threshold_age,
  total_excess = total_excess_deaths,
  below_threshold = excess_below_threshold,
  above_threshold = excess_above_threshold,
  pct_below = pct_below,
  pct_above = pct_above
)]

cat("\n=== Threshold Classification Summary ===\n")
print(threshold_summary[adaptation %in% c(0, 50, 100)])

# Plot classification
ggplot(city_results, 
       aes(x = adaptation * 100, y = pct_below, color = factor(ssp))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray30") +
  scale_color_manual(
    values = c("2" = "#2166ac", "3" = "#b2182b"),
    labels = c("SSP2-4.5", "SSP3-7.0"),
    name = "Scenario"
  ) +
  labs(
    title = sprintf("Excess Deaths Classification — %s", target_city_name),
    subtitle = sprintf("Deaths relative to threshold age (~%.0f years)", 
                       mean(city_results$threshold_age)),
    x = "Adaptation Level by 2100 (%)",
    y = "% of Excess Deaths Below Threshold",
    caption = "Deaths below threshold increase inequality; deaths above threshold decrease inequality.\n50% line marks equal split."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

**Key questions:**
- Are most excess deaths below or above threshold?
- Does the classification shift with adaptation?
- Is there a difference between SSP2 and SSP3?

**Interpretation guidance:**
- If `pct_below` > 70%: Climate predominantly causes **premature deaths** → expect ↑ inequality
- If `pct_below` < 30%: Climate predominantly affects **late deaths** → expect ↓ inequality (compression)
- If `pct_below` ≈ 50%: Mixed effects, net inequality change depends on magnitude

---

## 7.3 Threshold Stability Check

The threshold age itself can shift slightly between scenarios. This plot checks whether the threshold is **stable** across scenarios.

**What to expect:**
- Threshold should be relatively constant (±2-5 years variation)
- Large shifts would indicate fundamental changes in mortality patterns
- Threshold typically increases slightly with adaptation (as LE improves)

```{r threshold-stability, fig.height=5}
ggplot(city_results, 
       aes(x = adaptation * 100, y = threshold_age, color = factor(ssp))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("2" = "#2166ac", "3" = "#b2182b"),
    labels = c("SSP2-4.5", "SSP3-7.0"),
    name = "Scenario"
  ) +
  labs(
    title = sprintf("Threshold Age Stability — %s", target_city_name),
    subtitle = "Age threshold across scenarios (should be relatively stable)",
    x = "Adaptation Level by 2100 (%)",
    y = "Threshold Age (years)",
    caption = "Threshold age approximates where remaining LE ≈ LE at age 20"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

# 8. Multi-Metric Inequality Comparison

## 8.1 Gini vs Standard Deviation

Different inequality metrics can reveal different aspects of distributional change. Here we compare **Gini coefficient** and **standard deviation** of age at death.

**Why compare?**
- **Gini**: Scale-invariant, captures entire distribution, bounded [0,1]
- **SD**: Intuitive (years), sensitive to outliers, unbounded

**Expected pattern:** 
- If Gini and SD move in the same direction → robust signal of inequality change
- If they diverge → nuanced distributional changes (e.g., shifts vs spreads)

**What to look for:**
- Strong positive correlation (both metrics agree)
- Any scenarios where metrics disagree
- Whether adaptation affects both metrics similarly

```{r plot-gini-vs-sd, fig.height=6}
# Prepare data for correlation plot
metrics_data <- city_results[, .(
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  adaptation = adaptation * 100,
  delta_gini = delta_gini,
  delta_sd = delta_sd,
  pct_change_gini = pct_change_gini,
  pct_change_sd = pct_change_sd
)]

ggplot(metrics_data, 
       aes(x = delta_gini, y = delta_sd, color = ssp, size = adaptation)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(
    values = c("SSP2-4.5" = "#2166ac", "SSP3-7.0" = "#b2182b"),
    name = "Scenario"
  ) +
  scale_size_continuous(name = "Adaptation (%)", range = c(2, 8)) +
  labs(
    title = sprintf("Multi-Metric Inequality Comparison — %s", target_city_name),
    subtitle = "Do different inequality metrics agree?",
    x = "Change in Gini Coefficient",
    y = "Change in Standard Deviation (years)",
    caption = "If points cluster along a line, metrics agree. Scatter suggests different distributional changes."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

**Correlation check:**

```{r correlation-check}
cor_gini_sd <- cor(city_results$delta_gini, city_results$delta_sd)
cat(sprintf("\nCorrelation (Gini vs SD): %.3f\n", cor_gini_sd))

if (abs(cor_gini_sd) > 0.9) {
  cat("→ Strong agreement: Both metrics tell the same story\n")
} else if (abs(cor_gini_sd) > 0.7) {
  cat("→ Moderate agreement: Generally consistent but some nuance\n")
} else {
  cat("→ Weak agreement: Metrics capture different aspects of inequality\n")
}
```

---

## 8.2 Interquartile Range (IQR)

The **IQR** measures the spread of the "middle 50%" of deaths:

$$
IQR = Q_{75} - Q_{25}
$$

where $Q_p$ is the $p$-th percentile of age at death.

**Advantages:**
- Robust to outliers (focuses on central distribution)
- Easy interpretation (years between 25th and 75th percentile)

**Disadvantage:**
- Ignores tails (doesn't capture extreme early/late deaths)

```{r plot-iqr, fig.height=6}
iqr_data <- city_results[, .(
  adaptation = adaptation * 100,
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  delta_iqr = delta_iqr
)]

ggplot(iqr_data, aes(x = adaptation, y = delta_iqr, color = ssp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  scale_color_manual(
    values = c("SSP2-4.5" = "#2166ac", "SSP3-7.0" = "#b2182b"),
    name = "Scenario"
  ) +
  labs(
    title = sprintf("Interquartile Range of Age at Death — %s", target_city_name),
    subtitle = "Change in spread of middle 50% of deaths",
    x = "Adaptation Level by 2100 (%)",
    y = "Change in IQR (years)",
    caption = "IQR = Q75 - Q25. Positive values indicate wider spread of central deaths."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

**Interpretation:**
- ΔI QR > 0: The middle 50% of deaths become more spread out
- ΔIQR < 0: The middle 50% of deaths become more concentrated (compression)
- Compare to Gini/SD: If IQR shows different pattern, distributional changes may be in tails

---

# 9. Adaptation Effectiveness

## 9.1 Effectiveness Curves

This analysis assesses **adaptation as a policy intervention**: How much benefit (in terms of LE and LI) does each % of adaptation provide?

**Questions to explore:**
- Is adaptation effectiveness **linear** (constant returns) or **non-linear** (diminishing/accelerating returns)?
- What is the **marginal benefit** of increasing adaptation from 50% to 100%?
- Is there an **optimal adaptation level** (maximum benefit per unit effort)?

```{r adaptation-effectiveness, fig.height=8}
# Panel 1: Life expectancy gains from adaptation
p1 <- ggplot(city_results[ssp == 2], 
             aes(x = adaptation * 100, y = delta_e20 * 12)) +
  geom_line(linewidth = 1.2, color = "#2166ac") +
  geom_point(size = 3, color = "#2166ac") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  labs(
    title = "Life Expectancy Gain from Adaptation",
    subtitle = "SSP2-4.5 scenario",
    x = "Adaptation Level (%)",
    y = "ΔLE (months)"
  ) +
  theme_minimal(base_size = 11)

# Panel 2: Inequality reduction from adaptation
p2 <- ggplot(city_results[ssp == 2], 
             aes(x = adaptation * 100, y = delta_gini)) +
  geom_line(linewidth = 1.2, color = "#2166ac") +
  geom_point(size = 3, color = "#2166ac") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  labs(
    title = "Inequality Change with Adaptation",
    subtitle = "SSP2-4.5 scenario",
    x = "Adaptation Level (%)",
    y = "ΔGini"
  ) +
  theme_minimal(base_size = 11)

# Panel 3: Marginal effectiveness (derivative)
city_results[, marginal_le := c(NA, diff(delta_e20) / diff(adaptation)), by = ssp]

p3 <- ggplot(city_results[ssp == 2 & !is.na(marginal_le)],
             aes(x = adaptation * 100, y = marginal_le * 12)) +
  geom_line(linewidth = 1.2, color = "#d95f02") +
  geom_point(size = 3, color = "#d95f02") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  labs(
    title = "Marginal Benefit of Adaptation",
    subtitle = "Additional LE gain per 10% adaptation increase",
    x = "Adaptation Level (%)",
    y = "Marginal ΔLE (months per 10%)"
  ) +
  theme_minimal(base_size = 11)

# Combine
grid.arrange(p1, p2, p3, ncol = 1,
             top = sprintf("Adaptation Effectiveness — %s", target_city_name))
```

**Policy interpretation:**
- **Linear marginal benefit**: Each % of adaptation provides equal benefit → no priority for early vs late adaptation
- **Diminishing returns**: Early adaptation provides more benefit → prioritize achieving 50% quickly
- **Accelerating returns**: Later adaptation provides more benefit → may justify phased implementation

---

## 9.2 Adaptation Benefit-Cost Framework

While this analysis does not include cost data, we can frame the **effectiveness ratio**:

$$
\text{Effectiveness} = \frac{\Delta e_{20} \text{ (years gained)}}{\text{Adaptation level } (\%)}
$$

```{r effectiveness-ratio}
city_results[, effectiveness_ratio := delta_e20 / (adaptation + 0.001)]  # Avoid division by zero

effectiveness_table <- city_results[adaptation %in% c(0.5, 1.0), .(
  ssp = factor(ssp, labels = c("SSP2-4.5", "SSP3-7.0")),
  adaptation = sprintf("%.0f%%", adaptation * 100),
  delta_e20_months = sprintf("%.1f", delta_e20 * 12),
  effectiveness = sprintf("%.3f months/%% adapt", effectiveness_ratio * 12)
)]

cat("\n=== Adaptation Effectiveness Ratios ===\n")
print(effectiveness_table)
```

**Interpretation:** This ratio shows "months of life expectancy gained per % of adaptation achieved." Higher values indicate more efficient adaptation strategies.

**Caveat:** This does not account for:
- Implementation costs
- Diminishing returns at high adaptation levels
- Feasibility constraints
- Heterogeneous effectiveness across age groups

---

# 10. Summary Visualization

## 10.1 Comprehensive Metrics Heatmap

This heatmap shows **all key metrics simultaneously** across scenarios, allowing pattern recognition.

```{r summary-heatmap, fig.width=12, fig.height=6}
# Prepare data for heatmap
heatmap_data <- city_results[, .(
  scenario = sprintf("SSP%d-%.0f%%", ssp, adaptation * 100),
  `ΔLE (months)` = delta_e20 * 12,
  `ΔGini (%)` = pct_change_gini,
  `ΔSD (%)` = pct_change_sd,
  `% Below Threshold` = pct_below,
  `Excess Deaths` = total_excess_deaths
)]

# Reshape for ggplot
heatmap_long <- melt(heatmap_data, id.vars = "scenario",
                     variable.name = "Metric", value.name = "Value")

# Normalize each metric to [-1, 1] for color scale
heatmap_long[, Value_norm := (Value - mean(Value)) / (max(abs(Value)) + 1e-6), by = Metric]

ggplot(heatmap_long, aes(x = scenario, y = Metric, fill = Value_norm)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", Value)), size = 2.5) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0,
    name = "Normalized\nValue"
  ) +
  labs(
    title = sprintf("Comprehensive Summary — %s", target_city_name),
    subtitle = "All key metrics across SSP × adaptation scenarios",
    x = "Scenario (SSP-Adaptation)",
    y = "Metric"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

**How to read:**
- **Red cells**: Negative impacts (higher mortality, more inequality)
- **Blue cells**: Positive impacts (lower mortality, less inequality)
- **White cells**: Neutral (close to baseline)

**Pattern recognition:**
- Do all metrics agree (all red or all blue)?
- Are there contradictions (e.g., blue LE but red Gini)?
- How does adaptation shift the color palette?

---

## 10.2 Key Takeaways Table

```{r takeaways-table}
# Generate key statistics for presentation
takeaways <- data.table(
  Metric = c(
    "Baseline LE at age 20",
    "LE change (0% adapt, SSP2)",
    "LE change (50% adapt, SSP2)",
    "LE change (100% adapt, SSP2)",
    "Baseline Gini coefficient",
    "Gini change (0% adapt, SSP2)",
    "Gini change (50% adapt, SSP2)",
    "Threshold age",
    "% excess deaths below threshold (SSP2, 50% adapt)"
  ),
  Value = c(
    sprintf("%.2f years", city_results[1, e20_base]),
    sprintf("%.1f months", city_results[ssp == 2 & adaptation == 0.0, delta_e20 * 12]),
    sprintf("%.1f months", city_results[ssp == 2 & adaptation == 0.5, delta_e20 * 12]),
    sprintf("%.1f months", city_results[ssp == 2 & adaptation == 1.0, delta_e20 * 12]),
    sprintf("%.4f", city_results[1, gini_base]),
    sprintf("%.4f (%.1f%%)", 
            city_results[ssp == 2 & adaptation == 0.0, delta_gini],
            city_results[ssp == 2 & adaptation == 0.0, pct_change_gini]),
    sprintf("%.4f (%.1f%%)", 
            city_results[ssp == 2 & adaptation == 0.5, delta_gini],
            city_results[ssp == 2 & adaptation == 0.5, pct_change_gini]),
    sprintf("%.0f years", mean(city_results$threshold_age)),
    sprintf("%.1f%%", city_results[ssp == 2 & adaptation == 0.5, pct_below])
  )
)

cat("\n=== Key Takeaways ===\n")
print(takeaways, row.names = FALSE)
```

---

# 11. Discussion and Next Steps

## 11.1 Summary of Findings

This analysis examined the impact of climate change on life expectancy and lifespan inequality for **`r target_city_name`** under different SSP scenarios and adaptation levels.

**Key observations to consider:**

1. **Life Expectancy:**
   - Baseline LE at age 20: **`r sprintf("%.2f years", city_results[1, e20_base])`**
   - Climate impact ranges from **`r sprintf("%.1f", min(city_results$delta_e20) * 12)` to `r sprintf("%.1f", max(city_results$delta_e20) * 12)` months** depending on scenario
   - Adaptation effectiveness: [Reader should assess from figures]

2. **Lifespan Inequality:**
   - Baseline Gini coefficient: **`r sprintf("%.4f", city_results[1, gini_base])`**
   - Climate impact on Gini: **`r sprintf("%.1f%% to %.1f%%", min(city_results$pct_change_gini), max(city_results$pct_change_gini))`** change
   - Direction: [Reader should determine if predominantly increasing/decreasing]

3. **Threshold Analysis:**
   - Threshold age: approximately **`r sprintf("%.0f years", mean(city_results$threshold_age))`**
   - Excess deaths classification: [Reader should assess from Section 7]

4. **LE-LI Relationship:**
   - [Reader should determine from scatter plot]: Paradox present? Quadrant dominance?

---

## 11.2 Open Questions for Discussion

1. **Magnitude Assessment:**
   - Are the observed changes (months of LE, % change in Gini) large or small in practical terms?
   - How do these changes compare to other mortality interventions (e.g., smoking cessation, medical advances)?

2. **Mechanistic Understanding:**
   - Why does this city show this particular pattern of LE and LI changes?
   - What role does the baseline climate play?
   - Are there city-specific factors (latitude, heat island effect, healthcare capacity) that explain the results?

3. **Adaptation Feasibility:**
   - Is the adaptation effectiveness observed here realistic?
   - What specific interventions might achieve these adaptation levels?
   - Are there equity considerations in adaptation access?

4. **Comparative Context:**
   - How does this city compare to others in the same region?
   - Is there a north-south gradient in Europe?
   - Do Mediterranean cities show different patterns than Nordic cities?

5. **Policy Implications:**
   - What adaptation level should be targeted by 2100?
   - Is there a cost-effective adaptation strategy suggested by the diminishing/accelerating returns analysis?
   - Should policy prioritize LE gains or LI reduction?

---

## 11.3 Limitations and Caveats

**Methodological:**
- Adaptation model assumes linear risk reduction (may oversimplify)
- ERFs based on historical data (may not fully capture future adaptation)
- Single cohort tracked (cohort born 2020, starting at age 20)
- No consideration of migration or demographic change

**Data:**
- GCM uncertainty (reflected in ensemble spread)
- Mortality projections assume baseline improvements continue
- City-specific ERFs may have wide confidence intervals

**Scope:**
- Heat and cold mortality only (excludes other climate health impacts)
- No consideration of air quality, vector-borne diseases, extreme events
- Age 20+ only (excludes infant and child mortality)

---

## 11.4 Extensions and Future Work

**With current data:**
- Compare this city to others (multi-city analysis)
- Examine geographic patterns (north-south, coastal vs inland)
- Decompose inequality into within- vs between-city components (Paper 1, Output 3)

**With additional data (detailed life tables):**
- Age-specific excess deaths (which ages drive the patterns?)
- Age-at-death distribution histograms
- Decomposition of Gini into age-group contributions

**Methodological extensions:**
- Uncertainty quantification (confidence intervals from GCM ensemble)
- Sensitivity analysis (how robust are results to ERF uncertainty?)
- Alternative adaptation scenarios (non-linear, age-specific)

**Policy analysis:**
- Cost-benefit analysis (if cost data available)
- Equity analysis (differential adaptation access by socioeconomic status)
- Health-in-All-Policies framing (co-benefits of adaptation)

---

# 12. References

**Key Papers:**

1. Vaupel, J. W., Zhang, Z., & van Raalte, A. A. (2011). "Life expectancy and disparity: an international comparison of life table data." *BMJ Open*, 1(1), e000128.
   - Foundational paper on threshold age and lifespan inequality

2. Edwards, R. D., & Tuljapurkar, S. (2005). "Inequality in life spans and a new perspective on mortality convergence across industrialized countries." *Population and Development Review*, 31(4), 645-674.
   - Gini coefficient for mortality analysis

3. van Raalte, A. A., Sasson, I., & Martikainen, P. (2018). "The case for monitoring life-span inequality." *Science*, 362(6418), 1002-1004.
   - Why lifespan inequality matters for policy

4. Masselot, P., et al. (2023). "Excess mortality attributed to heat and cold: a health impact assessment study in 854 cities in Europe." *The Lancet Planetary Health*, 7(4), e271-e281.
   - Source of exposure-response functions used in this analysis

5. Colonna, N. (2025). "Climate Change, Mortality, and Lifespan Inequality: A Cohort Approach for European Cities." *Working Paper*.
   - Methods and full pipeline description

---

# Session Information

```{r session-info}
sessionInfo()
```

---

**End of Analysis**

**Contact:** Simon Project Team  
**Last Updated:** `r Sys.Date()`  
**Data Version:** Results from full 854-city pipeline
