---
title: "Heat-Related Mortality and Lifespan Inequality"
subtitle: "Part 2: Life Tables and Inequality Analysis"
author: "Daniel Precioso"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    keep_tex: false
---

# Overview

This notebook (Part 2) constructs **cohort life tables** and analyzes **lifespan inequality** using mortality multipliers computed in Part 1.

**Prerequisites**: You must run **Part 1** first to generate:
- Mortality multipliers by age and year
- Configuration file with city and scenario parameters

**This notebook covers**:
1. Loading baseline mortality projections (Eurostat EUROPOP2023)
2. Applying climate multipliers to baseline mortality rates
3. Constructing cohort life tables (baseline vs climate-adjusted)
4. Computing life expectancy impacts
5. Analyzing lifespan inequality metrics (Gini, SD, IQR)
6. Threshold analysis to understand inequality mechanisms

**Key concept**: Lifespan inequality measures how unequally distributed ages at death are. Climate change affects inequality through the **age pattern** of excess deaths, not just total mortality.

---

# 1. Setup and Configuration

## 1.1 Load Required Libraries

```{r setup-libraries, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(scales)

cat("Libraries loaded successfully.\n")
```

---

## 1.2 Load Configuration from Part 1

```{r load-config}
# Default configuration (modify if needed)
city <- "Vienna"
city_code <- "AT001C"
scenario <- 2
adaptation_level <- 0.5  # 50% adaptation by 2100


cat("=== Configuration Loaded from Part 1 ===\n")
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: SSP%d\n", scenario))
cat(sprintf("  Adaptation by 2100: %.0f%%\n", adaptation_level * 100))
```

---

## 1.3 Load Mortality Multipliers from Part 1

```{r load-multipliers}
# Construct multipliers file path
mult_path <- sprintf("results/mortality_multipliers_%s_ssp%d.csv",
                     tolower(gsub(" ", "_", city)),
                     scenario)

# Load multipliers
multipliers <- fread(mult_path)

cat(sprintf("\n=== Mortality Multipliers Loaded ===\n"))
cat(sprintf("  File: %s\n", mult_path))
cat(sprintf("  Records: %s\n", format(nrow(multipliers), big.mark = ",")))
cat(sprintf("  Years: %d–%d\n", min(multipliers$year), max(multipliers$year)))
cat(sprintf("  Ages: %d–%d\n", min(multipliers$age), max(multipliers$age)))
cat(sprintf("  Mean multiplier: %.4f\n", mean(multipliers$multiplier_mean)))
cat(sprintf("  Multiplier range: [%.3f, %.3f]\n", 
            min(multipliers$multiplier_mean), max(multipliers$multiplier_mean)))

head(multipliers, 10)
```

**Data structure**:
- `year`: Calendar year (2015–2100)
- `age`: Single-year age (20–100)
- `multiplier_mean`: Ensemble mean across GCMs
- `multiplier_sd`: Standard deviation across GCMs
- `multiplier_q05`, `multiplier_q50`, `multiplier_q95`: Quantiles for uncertainty

**Interpretation**: `multiplier > 1` indicates increased mortality risk compared to historical baseline.

---

# 2. Baseline Mortality Data

To construct life tables, we need **baseline mortality rates** by age and year. These represent expected mortality in the absence of climate change impacts.

We use **Eurostat mortality projections** (EUROPOP2023), which provide age-specific death rates for European countries through 2100.

## 2.1 Load Eurostat Mortality Projections

```{r load-baseline-mortality}
# Eurostat mortality projections for Austria
# Data source: Eurostat demo_mprojmorr (EUROPOP2023)
# 
# If you don't have this file, you can download it from:
# https://ec.europa.eu/eurostat/databrowser/view/demo_mprojmorr/

# For this demo, we'll load pre-processed data
mort_proj_path <- "../data/mortality_projections_austria.csv"

if (!file.exists(mort_proj_path)) {
  cat("Warning: Baseline mortality file not found.\n")
  cat("Please download Eurostat EUROPOP2023 projections.\n")
  cat("Using synthetic data for demonstration...\n")
  
  # Create synthetic baseline mortality (Gompertz model)
  # mx = a * exp(b * age)
  # Typical values for high-income countries
  ages_synth <- 20:100
  years_synth <- 2015:2100
  
  mort_proj <- CJ(age = ages_synth, year = years_synth)
  
  # Gompertz parameters (calibrated to approximate Austria)
  a <- 0.0001
  b <- 0.08
  
  # Add slight mortality improvement over time (1% per decade)
  mort_proj[, mx := a * exp(b * age) * (0.99 ^ ((year - 2015) / 10))]
  
} else {
  mort_proj <- fread(mort_proj_path)
  
  # Filter to relevant ages and years
  mort_proj <- mort_proj[age >= 20 & age <= 100]
  mort_proj <- mort_proj[year >= 2015 & year <= 2100]
}

# Ensure proper ordering
setkey(mort_proj, year, age)

cat(sprintf("\n=== Baseline Mortality Data ===\n"))
cat(sprintf("  Ages: %d–%d\n", min(mort_proj$age), max(mort_proj$age)))
cat(sprintf("  Years: %d–%d\n", min(mort_proj$year), max(mort_proj$year)))
cat(sprintf("  Records: %s\n", format(nrow(mort_proj), big.mark = ",")))

head(mort_proj, 10)
```

**Data structure**:
- `age`: Single-year ages (20–100)
- `year`: Calendar year (2015–2100)
- `mx`: Central death rate (annual probability of death)

**Interpretation**: `mx = 0.001` means 1 death per 1,000 person-years (0.1% annual mortality)

---

## 2.2 Baseline Mortality Summary

```{r baseline-mort-summary}
# Sample mortality rates for selected ages and years
sample_years <- c(2020, 2050, 2080)
sample_ages <- c(30, 50, 70, 90)

mort_sample <- mort_proj[year %in% sample_years & age %in% sample_ages]
mort_wide <- dcast(mort_sample, age ~ year, value.var = "mx")

cat("\n=== Baseline Mortality Rates (mx) ===\n")
print(mort_wide)

# Visualize baseline mortality by age (for selected years)
ggplot(mort_proj[year %in% sample_years], aes(x = age, y = mx, color = factor(year))) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_viridis_d(name = "Year") +
  labs(
    title = "Baseline Mortality Rates by Age",
    subtitle = "Eurostat EUROPOP2023 projections (no climate impacts)",
    x = "Age",
    y = "Central Death Rate (mx)",
    caption = "Log scale"
  ) +
  theme_minimal(base_size = 12)
```

---

# 3. Apply Climate Multipliers to Baseline Mortality

We now combine:

1. **Baseline mortality**: `mx_base[age, year]` from Eurostat
2. **Climate multipliers**: `M[age, year]` from Part 1

To obtain **climate-adjusted mortality**:

$$
mx_{\text{climate}}(a, y) = mx_{\text{baseline}}(a, y) \times M(a, y)
$$

where $M(a, y) > 1$ indicates increased mortality due to climate change.

---

## 3.1 Merge Multipliers with Baseline Mortality

```{r merge-multipliers}
# Use ensemble mean multiplier (already averaged across GCMs)
mult_subset <- multipliers[, .(age, year, multiplier = multiplier_mean)]

# Merge with baseline mortality
mort_climate <- merge(
  mort_proj,
  mult_subset,
  by = c("age", "year"),
  all.x = TRUE
)

# For years/ages without multipliers, assume no climate impact
mort_climate[is.na(multiplier), multiplier := 1.0]

# Compute climate-adjusted mortality
mort_climate[, mx_climate := mx * multiplier]

cat(sprintf("\n=== Climate-Adjusted Mortality Computed ===\n"))
cat(sprintf("  Records: %s\n", format(nrow(mort_climate), big.mark = ",")))
cat(sprintf("  Mean multiplier: %.4f\n", mean(mort_climate$multiplier, na.rm = TRUE)))
cat(sprintf("  Mean baseline mx: %.6f\n", mean(mort_climate$mx)))
cat(sprintf("  Mean climate mx: %.6f\n", mean(mort_climate$mx_climate)))

head(mort_climate, 15)
```

---

## 3.2 Comparison: Baseline vs Climate Mortality

We examine the **relative impact** of climate change on mortality by computing percentage change.

**Formula:** 
$$
\text{Percentage change} = \frac{mx_{\text{climate}} - mx_{\text{baseline}}}{mx_{\text{baseline}}} \times 100\%
$$

**Interpretation:**
- Positive values: Climate **increases** mortality risk (worse outcomes)
- Negative values: Climate **decreases** mortality risk (e.g., reduced cold-related deaths)

```{r compare-mortality}
# Compute percentage change
mort_climate[, pct_change := ((mx_climate - mx) / mx) * 100]

# Summary statistics
comparison <- mort_climate[year >= 2050, .(
  mean_mx_base    = mean(mx),
  mean_mx_climate = mean(mx_climate),
  mean_multiplier = mean(multiplier),
  mean_pct_change = mean(pct_change),
  max_pct_change  = max(pct_change),
  min_pct_change  = min(pct_change)
), by = year]

cat("\n=== Mortality Comparison: Baseline vs Climate (2050+) ===\n")
print(comparison[year %in% c(2050, 2070, 2090, 2099)])

# Visualize percentage change
ggplot(mort_climate[year %in% c(2050, 2070, 2099)], 
       aes(x = age, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_line(linewidth = 1.2, color = "#d62728") +
  geom_ribbon(aes(ymin = 0, ymax = pct_change, fill = pct_change > 0), alpha = 0.3) +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_manual(
    values = c("TRUE" = "#d62728", "FALSE" = "#2166ac"),
    labels = c("TRUE" = "Increased risk", "FALSE" = "Decreased risk"),
    name = ""
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = sprintf("Relative Mortality Impact of Climate Change — %s (SSP%d)", 
                    city, scenario),
    subtitle = sprintf("Adaptation: %.0f%% by 2100 | Percentage change in death rates",
                       adaptation_level * 100),
    x = "Age",
    y = "Change in Mortality Rate (%)",
    caption = "Positive values: increased mortality | Negative values: decreased mortality"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11)
  )
```

---

# 4. Cohort Life Table Construction

A **cohort life table** tracks a hypothetical birth cohort through time, applying age-specific mortality rates for each calendar year they live through.

For a cohort born in year $Y_0$:
- At age 20 (year $Y_0 + 20$), use mortality for year $Y_0 + 20$, age 20
- At age 21 (year $Y_0 + 21$), use mortality for year $Y_0 + 21$, age 21
- ... and so on until the cohort dies out

This differs from a **period life table**, which uses mortality rates from a single year.

---

## 4.1 Define Cohort Parameters

```{r cohort-setup}
cohort_birth_year <- 2020  # Birth year of cohort
cohort_start_age  <- 20    # Start tracking at age 20 (year 2040)
radix             <- 100000  # Initial cohort size (lx at start age)

cat(sprintf("\n=== Cohort Parameters ===\n"))
cat(sprintf("  Birth year: %d\n", cohort_birth_year))
cat(sprintf("  Start tracking at age %d (year %d)\n", 
            cohort_start_age, cohort_birth_year + cohort_start_age))
cat(sprintf("  Initial cohort size (radix): %s\n", format(radix, big.mark = ",")))
```

---

## 4.2 Build Cohort Life Table Function

```{r lifetable-function}
build_cohort_lifetable <- function(mort_data, birth_year, start_age, radix = 100000) {
  # Filter to cohort's age-year trajectory
  # For someone born in 'birth_year', at age 'a' they are in year 'birth_year + a'
  
  mort_data <- copy(mort_data)
  mort_data[, cohort_year := year - age]
  
  cohort_data <- mort_data[cohort_year == birth_year & age >= start_age]
  setorder(cohort_data, age)
  
  if (nrow(cohort_data) == 0) {
    stop("No data found for this cohort")
  }
  
  # Build life table columns
  n <- nrow(cohort_data)
  
  # Convert central death rate (mx) to probability of death (qx)
  # Standard approximation: qx ≈ mx / (1 + 0.5 * mx)
  cohort_data[, qx_base := mx / (1 + 0.5 * mx)]
  cohort_data[, qx_climate := mx_climate / (1 + 0.5 * mx_climate)]
  
  # Ensure qx does not exceed 1
  cohort_data[qx_base > 1, qx_base := 1]
  cohort_data[qx_climate > 1, qx_climate := 1]
  
  # Initialize survivors
  cohort_data[, lx_base := as.numeric(NA)]
  cohort_data[, lx_climate := as.numeric(NA)]
  
  cohort_data[1, lx_base := radix]
  cohort_data[1, lx_climate := radix]
  
  # Iteratively compute survivors and deaths
  for (i in 1:n) {
    if (i < n) {
      # Baseline
      cohort_data[i, dx_base := lx_base * qx_base]
      cohort_data[i + 1, lx_base := cohort_data[i, lx_base - dx_base]]
      
      # Climate
      cohort_data[i, dx_climate := lx_climate * qx_climate]
      cohort_data[i + 1, lx_climate := cohort_data[i, lx_climate - dx_climate]]
    } else {
      # Last age: all remaining die
      cohort_data[i, dx_base := lx_base]
      cohort_data[i, dx_climate := lx_climate]
    }
  }
  
  # Person-years lived (Lx): approximation using average of lx and lx+1
  cohort_data[, Lx_base := (lx_base + c(lx_base[-1], 0)) / 2]
  cohort_data[, Lx_climate := (lx_climate + c(lx_climate[-1], 0)) / 2]
  
  # For last age, use lx / mx (if mx > 0)
  cohort_data[n, Lx_base := ifelse(mx > 0, lx_base / mx, lx_base)]
  cohort_data[n, Lx_climate := ifelse(mx_climate > 0, lx_climate / mx_climate, lx_climate)]
  
  # Total person-years from age x onwards (Tx)
  cohort_data[, Tx_base := rev(cumsum(rev(Lx_base)))]
  cohort_data[, Tx_climate := rev(cumsum(rev(Lx_climate)))]
  
  # Life expectancy at age x (ex)
  cohort_data[, ex_base := Tx_base / lx_base]
  cohort_data[, ex_climate := Tx_climate / lx_climate]
  
  return(cohort_data)
}
```

---

## 4.3 Construct Baseline and Climate Life Tables

```{r build-lifetable}
lifetable <- build_cohort_lifetable(
  mort_climate,
  birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  radix = radix
)

cat(sprintf("\n=== Cohort Life Table Constructed ===\n"))
cat(sprintf("  Ages: %d–%d\n", min(lifetable$age), max(lifetable$age)))
cat(sprintf("  Years: %d–%d\n", min(lifetable$year), max(lifetable$year)))

# Display key columns
lifetable_display <- lifetable[, .(
  age, year,
  mx_base = round(mx, 6),
  mx_climate = round(mx_climate, 6),
  multiplier = round(multiplier, 3),
  qx_base = round(qx_base, 5),
  qx_climate = round(qx_climate, 5),
  lx_base = round(lx_base, 0),
  lx_climate = round(lx_climate, 0),
  ex_base = round(ex_base, 2),
  ex_climate = round(ex_climate, 2)
)]

cat("\nLife Table (first 20 rows):\n")
print(head(lifetable_display, 20))

cat("\nLife Table (last 10 rows):\n")
print(tail(lifetable_display, 10))
```

---

# 5. Life Expectancy Analysis

## 5.1 Life Expectancy at Starting Age

```{r life-expectancy-start}
e_start_base <- lifetable[age == cohort_start_age, ex_base]
e_start_climate <- lifetable[age == cohort_start_age, ex_climate]

delta_e <- e_start_climate - e_start_base
pct_delta_e <- (delta_e / e_start_base) * 100

cat(sprintf("\n=== Life Expectancy at Age %d ===\n", cohort_start_age))
cat(sprintf("  Baseline (no climate impact):   %.2f years\n", e_start_base))
cat(sprintf("  Climate-adjusted (SSP%d):        %.2f years\n", scenario, e_start_climate))
cat(sprintf("  Difference:                     %.2f years (%.2f%%)\n", 
            delta_e, pct_delta_e))

if (delta_e < 0) {
  cat(sprintf("\n  → Climate change reduces life expectancy by %.1f months\n", 
              abs(delta_e) * 12))
} else {
  cat(sprintf("\n  → Climate change increases life expectancy by %.1f months\n", 
              delta_e * 12))
  cat("     (Note: This may occur if cold-related mortality reduction exceeds heat increase)\n")
}
```

---

## 5.2 Life Expectancy Difference by Age

```{r life-expectancy-by-age}
# Compute difference in life expectancy
lifetable[, delta_ex := ex_climate - ex_base]

ggplot(lifetable, aes(x = age, y = delta_ex)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.8) +
  geom_line(linewidth = 1.2, color = "#d62728") +
  geom_ribbon(aes(ymin = 0, ymax = delta_ex, fill = delta_ex > 0), alpha = 0.3) +
  scale_fill_manual(
    values = c("TRUE" = "#2166ac", "FALSE" = "#d62728"),
    labels = c("TRUE" = "LE gain", "FALSE" = "LE loss"),
    name = "Climate impact"
  ) +
  labs(
    title = sprintf("Climate Impact on Remaining Life Expectancy — Cohort Born %d",
                    cohort_birth_year),
    subtitle = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                       city, scenario, adaptation_level * 100),
    x = "Age",
    y = "Change in Remaining Life Expectancy (years)",
    caption = "Positive values: climate increases LE | Negative values: climate decreases LE"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

---

## 5.3 Survivorship Curves

```{r survivorship-curves}
surv_data <- lifetable[, .(
  age,
  Baseline = lx_base,
  Climate = lx_climate
)]

surv_long <- melt(surv_data, id.vars = "age",
                  variable.name = "Scenario", value.name = "lx")

ggplot(surv_long, aes(x = age, y = lx / radix, color = Scenario)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = sprintf("Survivorship Curve — Cohort Born %d", cohort_birth_year),
    subtitle = sprintf("Starting with %s people at age %d",
                       format(radix, big.mark = ","), cohort_start_age),
    x = "Age",
    y = "Proportion Surviving",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Interpretation**: The gap between curves represents excess deaths due to climate change.

---

## 5.4 Cumulative Deaths

```{r cumulative-deaths}
# Calculate cumulative deaths by age
lifetable[, cumul_deaths_base := cumsum(dx_base)]
lifetable[, cumul_deaths_climate := cumsum(dx_climate)]
lifetable[, excess_deaths := cumul_deaths_climate - cumul_deaths_base]

# Total excess deaths over lifetime
total_excess <- lifetable[age == max(age), excess_deaths]
pct_excess <- (total_excess / radix) * 100

cat(sprintf("\n=== Cumulative Excess Deaths ===\n"))
cat(sprintf("  Initial cohort size: %s\n", format(radix, big.mark = ",")))
cat(sprintf("  Total excess deaths: %.0f (%.2f%% of cohort)\n", 
            total_excess, pct_excess))

# Plot excess deaths over age
ggplot(lifetable, aes(x = age, y = excess_deaths)) +
  geom_area(fill = "#d62728", alpha = 0.6) +
  geom_line(color = "#d62728", linewidth = 1) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Cumulative Excess Deaths Due to Climate Change",
    subtitle = sprintf("Cohort born %d, starting with %s at age %d",
                       cohort_birth_year, format(radix, big.mark = ","), 
                       cohort_start_age),
    x = "Age",
    y = "Cumulative Excess Deaths",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12)
```

---

# 6. Export Life Table Results

```{r export-lifetable}
# Create output directory if needed
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

# Save full life table
output_path <- sprintf("results/lifetable_%s_ssp%d_cohort%d.csv",
                       tolower(gsub(" ", "_", city)),
                       scenario,
                       cohort_birth_year)

lifetable_export <- lifetable[, .(
  age, year,
  mx, mx_climate, multiplier,
  qx_base, qx_climate,
  lx_base, lx_climate,
  dx_base, dx_climate,
  ex_base, ex_climate
)]

fwrite(lifetable_export, output_path)
cat(sprintf("\nSaved life table: %s\n", output_path))

# Save summary statistics
summary_stats <- data.table(
  city = city,
  scenario = sprintf("SSP%d", scenario),
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  radix = radix,
  e_start_baseline = e_start_base,
  e_start_climate = e_start_climate,
  delta_e_years = delta_e,
  delta_e_months = delta_e * 12,
  pct_change_e = pct_delta_e,
  total_excess_deaths = total_excess,
  pct_excess_deaths = pct_excess
)

summary_path <- sprintf("results/lifetable_summary_%s_ssp%d_cohort%d.csv",
                        tolower(gsub(" ", "_", city)),
                        scenario,
                        cohort_birth_year)

fwrite(summary_stats, summary_path)
cat(sprintf("Saved life table summary: %s\n", summary_path))
```

---

# 7. Lifespan Inequality Analysis

Lifespan inequality (LI) measures the **dispersion** or **spread** of ages at death within a population. Unlike life expectancy (which is a mean), LI captures how equally or unequally distributed lifespans are.

**Key insight**: Two populations can have identical life expectancy but very different inequality:
- **Low LI**: Deaths concentrated around a narrow age range (e.g., most people die 75-85)
- **High LI**: Deaths spread across a wide age range (e.g., deaths from 60-100)

Climate change affects LI through the **age pattern** of excess deaths, not just their total number.

---

## 7.1 Age-at-Death Distribution

The life table column `dx` represents the **distribution of deaths by age**. From this, we can compute various inequality metrics.

```{r age-at-death-distribution}
# Extract age-at-death distributions
deaths_base <- lifetable[, .(age, dx = dx_base)]
deaths_climate <- lifetable[, .(age, dx = dx_climate)]

# Normalize to proportions (for visualization)
deaths_base[, prop := dx / sum(dx)]
deaths_climate[, prop := dx / sum(dx)]

cat(sprintf("\n=== Age-at-Death Distribution ===\n"))
cat(sprintf("Baseline deaths: %s\n", format(round(sum(deaths_base$dx)), big.mark = ",")))
cat(sprintf("Climate deaths: %s\n", format(round(sum(deaths_climate$dx)), big.mark = ",")))
cat(sprintf("Total excess deaths: %s\n", format(round(sum(deaths_climate$dx) - sum(deaths_base$dx)), big.mark = ",")))
```

---

## 7.2 Lifespan Inequality Metrics

### 7.2.1 Gini Coefficient

The **Gini coefficient** measures inequality on a 0-1 scale:
- **0**: Perfect equality (everyone dies at the same age)
- **1**: Maximum inequality (all deaths at extremes)

```{r compute-gini}
# Gini coefficient function
compute_gini <- function(ages, deaths) {
  # Remove zero-death ages
  valid <- deaths > 0
  ages <- ages[valid]
  deaths <- deaths[valid]
  
  n <- length(ages)
  total_deaths <- sum(deaths)
  
  # Weighted mean age at death
  mean_age <- sum(ages * deaths) / total_deaths
  
  # Compute pairwise absolute differences
  numerator <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      numerator <- numerator + abs(ages[i] - ages[j]) * deaths[i] * deaths[j]
    }
  }
  
  denominator <- 2 * mean_age * total_deaths^2
  
  gini <- numerator / denominator
  return(gini)
}

# Compute Gini for both scenarios
gini_base <- compute_gini(deaths_base$age, deaths_base$dx)
gini_climate <- compute_gini(deaths_climate$age, deaths_climate$dx)

delta_gini <- gini_climate - gini_base
pct_change_gini <- (delta_gini / gini_base) * 100

cat(sprintf("\n=== Gini Coefficient ===\n"))
cat(sprintf("Baseline:         %.4f\n", gini_base))
cat(sprintf("Climate-adjusted: %.4f\n", gini_climate))
cat(sprintf("Change:           %.4f (%.2f%%)\n", delta_gini, pct_change_gini))

if (delta_gini > 0) {
  cat("\n→ Climate change INCREASES lifespan inequality\n")
  cat("  (deaths become more unequally distributed)\n")
} else {
  cat("\n→ Climate change DECREASES lifespan inequality\n")
  cat("  (deaths become more equally distributed)\n")
}
```

---

### 7.2.2 Standard Deviation of Age at Death

The **standard deviation** (SD) measures dispersion in years.

```{r compute-sd}
# Standard deviation function
compute_sd_age_at_death <- function(ages, deaths) {
  total_deaths <- sum(deaths)
  mean_age <- sum(ages * deaths) / total_deaths
  variance <- sum((ages - mean_age)^2 * deaths) / total_deaths
  sd <- sqrt(variance)
  return(sd)
}

# Compute SD for both scenarios
sd_base <- compute_sd_age_at_death(deaths_base$age, deaths_base$dx)
sd_climate <- compute_sd_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_sd <- sd_climate - sd_base
pct_change_sd <- (delta_sd / sd_base) * 100

cat(sprintf("\n=== Standard Deviation of Age at Death ===\n"))
cat(sprintf("Baseline:         %.2f years\n", sd_base))
cat(sprintf("Climate-adjusted: %.2f years\n", sd_climate))
cat(sprintf("Change:           %.2f years (%.2f%%)\n", delta_sd, pct_change_sd))
```

**Interpretation**: SD quantifies how "spread out" deaths are. Higher SD means ages at death are more dispersed.

---

### 7.2.3 Interquartile Range (IQR)

The **IQR** is the difference between the 75th and 25th percentile of age at death.

```{r compute-iqr}
# IQR function
compute_iqr_age_at_death <- function(ages, deaths) {
  # Create cumulative distribution
  cumsum_deaths <- cumsum(deaths)
  total_deaths <- sum(deaths)
  
  # Find 25th percentile
  q25_idx <- which(cumsum_deaths >= 0.25 * total_deaths)[1]
  q25 <- ages[q25_idx]
  
  # Find 75th percentile
  q75_idx <- which(cumsum_deaths >= 0.75 * total_deaths)[1]
  q75 <- ages[q75_idx]
  
  iqr <- q75 - q25
  return(list(q25 = q25, q75 = q75, iqr = iqr))
}

# Compute IQR for both scenarios
iqr_base <- compute_iqr_age_at_death(deaths_base$age, deaths_base$dx)
iqr_climate <- compute_iqr_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_iqr <- iqr_climate$iqr - iqr_base$iqr

cat(sprintf("\n=== Interquartile Range (IQR) ===\n"))
cat(sprintf("Baseline:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_base$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_base$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_base$iqr))

cat(sprintf("\nClimate-adjusted:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_climate$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_climate$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_climate$iqr))

cat(sprintf("\nChange in IQR: %.2f years\n", delta_iqr))
```

---

## 7.3 Threshold Analysis

A critical insight from demography: there is a **threshold age** that determines whether additional deaths increase or decrease lifespan inequality.

**Rule of thumb**: The threshold is approximately where **remaining life expectancy** equals the population's **life expectancy at the starting age**.

- **Deaths below threshold**: Increase LI (premature deaths)
- **Deaths above threshold**: Decrease LI (late deaths compress distribution)

### 7.3.1 Identify Threshold Age

```{r identify-threshold}
# Threshold: age where remaining LE ≈ LE at starting age
e_start_val <- lifetable[age == cohort_start_age, ex_base]

# Find age closest to this remaining LE
lifetable[, diff_from_e_start := abs(ex_base - e_start_val)]
threshold_age <- lifetable[which.min(diff_from_e_start), age]

cat(sprintf("\n=== Threshold Age Analysis ===\n"))
cat(sprintf("Life expectancy at age %d: %.2f years\n", cohort_start_age, e_start_val))
cat(sprintf("Threshold age: %d years\n", threshold_age))
cat(sprintf("Remaining LE at threshold: %.2f years\n", 
            lifetable[age == threshold_age, ex_base]))
```

---

### 7.3.2 Classify Excess Deaths by Threshold

```{r classify-excess-deaths}
# Add threshold classification to life table
lifetable[, below_threshold := age < threshold_age]
lifetable[, excess_deaths_age := dx_climate - dx_base]

# Summarize excess deaths by position relative to threshold
excess_summary <- lifetable[, .(
  total_excess = sum(excess_deaths_age),
  excess_below = sum(excess_deaths_age[below_threshold == TRUE]),
  excess_above = sum(excess_deaths_age[below_threshold == FALSE])
)]

pct_below <- (excess_summary$excess_below / excess_summary$total_excess) * 100
pct_above <- (excess_summary$excess_above / excess_summary$total_excess) * 100

cat(sprintf("\n=== Excess Deaths Classification ===\n"))
cat(sprintf("Total excess deaths: %.0f\n", excess_summary$total_excess))
cat(sprintf("\nBelow threshold (age < %d):\n", threshold_age))
cat(sprintf("  Excess deaths: %.0f (%.1f%%)\n", excess_summary$excess_below, pct_below))
cat(sprintf("  Effect: INCREASES lifespan inequality\n"))

cat(sprintf("\nAbove threshold (age ≥ %d):\n", threshold_age))
cat(sprintf("  Excess deaths: %.0f (%.1f%%)\n", excess_summary$excess_above, pct_above))
cat(sprintf("  Effect: DECREASES lifespan inequality\n"))

# Net effect interpretation
if (excess_summary$excess_below > abs(excess_summary$excess_above)) {
  cat(sprintf("\n→ NET EFFECT: Climate change increases lifespan inequality\n"))
  cat(sprintf("   (more excess deaths below threshold)\n"))
} else {
  cat(sprintf("\n→ NET EFFECT: Climate change decreases lifespan inequality\n"))
  cat(sprintf("   (more excess deaths above threshold)\n"))
}
```

**Interpretation**: This classification reveals **where** climate impacts mortality distribution, not just **how much**.

---

## 7.4 Visualizations

### 7.4.1 Age-at-Death Distribution Comparison

```{r plot-age-at-death, fig.width=10, fig.height=6}
# Combine distributions for plotting
dist_combined <- rbind(
  deaths_base[, .(age, deaths = dx, scenario = "Baseline")],
  deaths_climate[, .(age, deaths = dx, scenario = "Climate")]
)

# Calculate proportions for clean visualization
dist_combined[, prop := deaths / sum(deaths), by = scenario]

ggplot(dist_combined, aes(x = age, y = prop, fill = scenario)) +
  geom_area(alpha = 0.6, position = "identity") +
  geom_vline(xintercept = threshold_age, linetype = "dashed", 
             color = "black", linewidth = 1) +
  annotate("text", x = threshold_age + 2, y = max(dist_combined$prop) * 0.9,
           label = sprintf("Threshold\n(age %d)", threshold_age),
           hjust = 0, size = 3.5) +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728"),
    name = "Scenario"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Age-at-Death Distribution",
    subtitle = sprintf("Cohort born %d — %s, SSP%d, Adaptation %.0f%%",
                       cohort_birth_year, city, scenario, adaptation_level * 100),
    x = "Age at Death",
    y = "Proportion of Deaths",
    caption = "Dashed line: threshold age (deaths below increase LI; above decrease LI)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

### 7.4.2 Excess Deaths by Age

```{r plot-excess-deaths-by-age, fig.width=10, fig.height=6}
# Highlight threshold region
ggplot(lifetable, aes(x = age, y = excess_deaths_age)) +
  geom_rect(aes(xmin = -Inf, xmax = threshold_age, ymin = -Inf, ymax = Inf),
            fill = "#fee5d9", alpha = 0.3) +
  geom_rect(aes(xmin = threshold_age, xmax = Inf, ymin = -Inf, ymax = Inf),
            fill = "#deebf7", alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  geom_line(color = "#d62728", linewidth = 1) +
  geom_vline(xintercept = threshold_age, linetype = "dashed", 
             color = "black", linewidth = 1) +
  annotate("text", x = threshold_age - 5, y = max(lifetable$excess_deaths_age) * 0.8,
           label = "Below threshold\n(increases LI)", hjust = 1, size = 3) +
  annotate("text", x = threshold_age + 5, y = max(lifetable$excess_deaths_age) * 0.8,
           label = "Above threshold\n(decreases LI)", hjust = 0, size = 3) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Excess Deaths by Age Due to Climate Change",
    subtitle = sprintf("Threshold age: %d (where remaining LE ≈ LE at age %d)",
                       threshold_age, cohort_start_age),
    x = "Age",
    y = "Excess Deaths (Climate - Baseline)",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12)
```

**Interpretation**: This plot shows the **age pattern** of climate impacts, revealing whether climate predominantly affects "younger old" or "older old" populations.

---

### 7.4.3 Lifespan Inequality Metrics Summary

```{r plot-li-metrics, fig.width=8, fig.height=6}
# Create summary data frame
li_metrics <- data.table(
  Metric = c("Gini", "Gini", "SD", "SD", "IQR", "IQR"),
  Scenario = rep(c("Baseline", "Climate"), 3),
  Value = c(gini_base, gini_climate, 
            sd_base, sd_climate,
            iqr_base$iqr, iqr_climate$iqr)
)

ggplot(li_metrics, aes(x = Scenario, y = Value, fill = Scenario)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = sprintf("%.3f", Value)), 
            vjust = -0.5, size = 3.5) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = "Lifespan Inequality Metrics: Baseline vs Climate",
    subtitle = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                       city, scenario, adaptation_level * 100),
    y = "Metric Value",
    x = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 7.5 Summary Table

```{r li-summary-table}
# Create comprehensive summary
li_summary <- data.table(
  Metric = c("Life Expectancy at Start", "Gini Coefficient", 
             "SD of Age at Death", "IQR of Age at Death",
             "Threshold Age", "Excess Deaths Below Threshold",
             "Excess Deaths Above Threshold"),
  Baseline = c(
    sprintf("%.2f years", e_start_base),
    sprintf("%.4f", gini_base),
    sprintf("%.2f years", sd_base),
    sprintf("%.1f years", iqr_base$iqr),
    sprintf("%d years", threshold_age),
    "—",
    "—"
  ),
  Climate = c(
    sprintf("%.2f years", e_start_climate),
    sprintf("%.4f", gini_climate),
    sprintf("%.2f years", sd_climate),
    sprintf("%.1f years", iqr_climate$iqr),
    "—",
    sprintf("%.0f (%.1f%%)", excess_summary$excess_below, pct_below),
    sprintf("%.0f (%.1f%%)", excess_summary$excess_above, pct_above)
  ),
  Change = c(
    sprintf("%.2f years (%.2f%%)", delta_e, pct_delta_e),
    sprintf("%.4f (%.2f%%)", delta_gini, pct_change_gini),
    sprintf("%.2f years (%.2f%%)", delta_sd, pct_change_sd),
    sprintf("%.1f years", delta_iqr),
    "—",
    "—",
    "—"
  )
)

cat("\n=== Lifespan Inequality Summary ===\n")
print(li_summary)
```

---

# 8. Export Lifespan Inequality Results

```{r export-li-results}
# Create output directory if needed
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

# Save detailed metrics
li_results <- data.table(
  city = city,
  scenario = sprintf("SSP%d", scenario),
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  # Life expectancy
  e_start_baseline = e_start_base,
  e_start_climate = e_start_climate,
  delta_e = delta_e,
  pct_change_e = pct_delta_e,
  # Gini
  gini_baseline = gini_base,
  gini_climate = gini_climate,
  delta_gini = delta_gini,
  pct_change_gini = pct_change_gini,
  # Standard deviation
  sd_baseline = sd_base,
  sd_climate = sd_climate,
  delta_sd = delta_sd,
  pct_change_sd = pct_change_sd,
  # IQR
  iqr_baseline = iqr_base$iqr,
  iqr_climate = iqr_climate$iqr,
  delta_iqr = delta_iqr,
  # Threshold
  threshold_age = threshold_age,
  # Excess deaths classification
  total_excess_deaths = excess_summary$total_excess,
  excess_deaths_below_threshold = excess_summary$excess_below,
  excess_deaths_above_threshold = excess_summary$excess_above,
  pct_excess_below = pct_below,
  pct_excess_above = pct_above
)

# Save
li_path <- sprintf("results/lifespan_inequality_%s_ssp%d_cohort%d.csv",
                   tolower(gsub(" ", "_", city)),
                   scenario,
                   cohort_birth_year)

fwrite(li_results, li_path)
cat(sprintf("\nSaved: %s\n", li_path))

# Also save age-specific excess deaths for detailed analysis
excess_by_age <- lifetable[, .(
  age, year,
  dx_baseline = dx_base,
  dx_climate = dx_climate,
  excess_deaths = excess_deaths_age,
  below_threshold,
  multiplier
)]

excess_path <- sprintf("results/excess_deaths_by_age_%s_ssp%d_cohort%d.csv",
                       tolower(gsub(" ", "_", city)),
                       scenario,
                       cohort_birth_year)

fwrite(excess_by_age, excess_path)
cat(sprintf("Saved: %s\n", excess_path))
```

---

## 8.1 Key Findings Interpretation

```{r interpretation}
cat("\n=== KEY FINDINGS ===\n\n")

cat(sprintf("For the cohort born in %d in %s under SSP%d:\n\n", 
            cohort_birth_year, city, scenario))

# Life expectancy
cat("1. LIFE EXPECTANCY\n")
if (delta_e < 0) {
  cat(sprintf("   → Climate change REDUCES life expectancy by %.2f years (%.2f%%)\n",
              abs(delta_e), abs(pct_delta_e)))
  cat(sprintf("   → Equivalent to %.1f months lost\n", abs(delta_e) * 12))
} else {
  cat(sprintf("   → Climate change INCREASES life expectancy by %.2f years (%.2f%%)\n",
              delta_e, pct_delta_e))
  cat("   → (Likely due to reduced cold-related mortality)\n")
}

# Lifespan inequality
cat("\n2. LIFESPAN INEQUALITY\n")
if (delta_gini > 0) {
  cat(sprintf("   → Climate change INCREASES inequality (Gini: +%.4f, +%.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE unequally distributed\n")
} else {
  cat(sprintf("   → Climate change DECREASES inequality (Gini: %.4f, %.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE equally distributed\n")
}

cat(sprintf("   → Standard deviation changes by %.2f years\n", delta_sd))
cat(sprintf("   → IQR changes by %.1f years\n", delta_iqr))

# Mechanism
cat("\n3. MECHANISM\n")
cat(sprintf("   → Threshold age: %d years\n", threshold_age))
cat(sprintf("   → %.1f%% of excess deaths occur BELOW threshold\n", pct_below))
cat(sprintf("   → %.1f%% of excess deaths occur ABOVE threshold\n", pct_above))

if (pct_below > 60) {
  cat("   → Climate predominantly kills 'younger old' people\n")
  cat("   → This is the PRIMARY driver of increased inequality\n")
} else if (pct_above > 60) {
  cat("   → Climate predominantly kills 'older old' people\n")
  cat("   → This compresses the age-at-death distribution\n")
} else {
  cat("   → Climate impacts are distributed across ages\n")
  cat("   → Net inequality effect depends on relative magnitudes\n")
}

# Policy implication
cat("\n4. POLICY IMPLICATION\n")
if (delta_gini > 0) {
  cat("   → Without intervention, climate change increases health inequity\n")
  cat("   → Priority: protect vulnerable 'younger old' populations\n")
} else {
  cat("   → Climate has a compression effect on lifespans\n")
  cat("   → But mean LE still affected—focus on absolute longevity\n")
}

cat(sprintf("\n   → With %.0f%% adaptation by 2100, these impacts are partially mitigated\n",
            adaptation_level * 100))
```

---

# 9. Next Steps

This analysis provides the methodological foundation for comprehensive multi-scenario studies:

1. **Multiple scenarios**: Run for all SSPs (1, 2, 3, 5) and adaptation levels (0%, 25%, 50%, 75%, 100%)
2. **Multiple cities**: Extend to 10-20 European cities (Vienna, Madrid, Stockholm, Athens, etc.)
3. **Multiple cohorts**: Analyze cohorts born in different decades (2000, 2020, 2040, etc.)
4. **Geographic decomposition**: Compute within-city vs between-city inequality components
5. **Temporal trends**: Track how LI evolves year-by-year
6. **Heat vs cold decomposition**: Separate LI impacts of heat vs cold mortality

**To run this analysis for a different scenario**:
1. Modify configuration in Part 1 (city, scenario, adaptation)
2. Run Part 1 to generate new multipliers
3. Re-run Part 2 with the same configuration

---

# Session Information

```{r session-info}
sessionInfo()
```
