---
title: "Heat-Related Mortality and Lifespan Inequality"
subtitle: "Climate Projection Analysis for European Cities"
author: "Simon Project"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This notebook estimates the impact of future temperature on mortality and lifespan inequality for a selected European city under a chosen climate scenario. The analysis integrates:

1. **Exposure-Response Functions (ERFs)** from Masselot et al., which quantify the relationship between temperature and mortality risk by age group
2. **Climate projections** from multiple General Circulation Models (GCMs) under a selected Shared Socioeconomic Pathway (SSP)
3. **Adaptation scenarios** that model potential reductions in heat vulnerability over time

The core output is age-specific **mortality multipliers** for each year (2015–2100) and age (20–100), which quantify changes in mortality risk relative to a historical baseline.

---

# 1. Configuration

We begin by selecting the analysis parameters:

- **City**: The urban area of interest
- **Scenario**: SSP pathway (1 = SSP1-2.6, 2 = SSP2-4.5, 3 = SSP3-7.0, 5 = SSP5-8.5)
- **Adaptation level**: The fraction of heat-related excess risk eliminated by 2100 (0 = no adaptation, 1 = complete adaptation)

```{r setup}
city <- "Vienna"
city_code <- "AT001C"
scenario <- 2  # SSP2-4.5
adaptation_level <- 0.5  # 50% risk reduction by 2100

cat(sprintf("Analysis Configuration:\n"))
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: SSP%d\n", scenario))
cat(sprintf("  Adaptation by 2100: %.0f%%\n", adaptation_level * 100))
cat(sprintf("  Years: 2015–2100 (all years)\n"))
```

---

# 2. Data Loading

## 2.1 Exposure-Response Functions (ERFs)

We load city-specific coefficients from Masselot et al. that define the temperature-mortality relationship. These coefficients parameterize a distributed lag non-linear model (DLNM) with a B-spline basis.

The ERF model is:

$$
\log RR(T) = \sum_{j=1}^{5} b_j \cdot s_j(T)
$$

where $s_j(T)$ are B-spline basis functions and $b_j$ are fitted coefficients.

```{r load-coefficients, message=FALSE}
source("../R/load_coefficients.R")

coefs <- load_city_coefficients(city_code, coefs_path = "../data/coefs.csv")

# Display city-specific coefficients by age group
head(coefs$city)
```

**Interpretation**: Each row represents one age group with 5 spline coefficients (b1–b5) that define the shape of the temperature-mortality curve for that age.

---

## 2.2 Single-Age Coefficient Interpolation

The original ERF coefficients are grouped into broad age categories:

- 20–44 years
- 45–64 years
- 65–74 years
- 75–84 years
- 85+ years

For fine-grained life table analysis, we interpolate these to **single-year ages** (20–100).

```{r interpolate-ages}
agelabs       <- c("20-44", "45-64", "65-74", "75-84", "85+")
age_midpoints <- c(32.5, 55, 70, 80, 92.5)
age_range     <- 20:100

coefs_single_age <- interpolate_coefs_to_single_age(
  coefs$city, coefs$all,
  agelabs, age_midpoints,
  age_range = age_range,
  city_code = city_code
)

head(coefs_single_age)
```

We now have ERF coefficients for **81 single-year ages** (20–100).

---

## 2.3 Temperature Projections

We load daily temperature projections from multiple GCMs for the selected city. The data covers:

- **Historical period** (`ssp == "hist"`): Used for baseline mortality risk
- **Future projections** (`ssp == scenario`): 2015–2100 under the selected SSP

```{r load-temperature, message=FALSE, warning=FALSE}
source("../R/load_data.R")

parquet_path <- "../data/tmeanproj.gz.parquet"

# Exclude problematic GCMs
gcmexcl <- c("CMCC_CM2_SR5", "TaiESM1")

temp_data <- load_projected_temperatures(
  city_code,
  parquet_path = parquet_path,
  gcmexcl = gcmexcl
)

proj_data <- temp_data$proj_data
gcm_cols  <- temp_data$gcm_cols

cat(sprintf("\nLoaded %d days of temperature data from %d GCMs\n",
            nrow(proj_data), length(gcm_cols)))
cat(sprintf("SSP scenarios available: %s\n", paste(unique(proj_data$ssp), collapse = ", ")))

head(proj_data)
```

**Data structure**: Each row is one day; columns `tas_*` contain daily mean temperature (°C) from different GCMs.

---

# 3. Data Preparation

## 3.1 Reshape Temperature Data

We convert the wide-format GCM columns into a long format for the selected scenario:

```{r reshape-temperature}
library(data.table)
setDT(proj_data)

# Extract selected scenario only
temp_long <- proj_data[ssp == scenario]

temp_long <- melt(
  temp_long,
  id.vars = c("URAU_CODE", "date", "year", "doy", "ssp"),
  measure.vars = gcm_cols,
  variable.name = "gcm",
  value.name = "temp"
)

cat(sprintf("Reshaped data: %d observations (days × GCMs)\n", nrow(temp_long)))
cat(sprintf("Years: %d–%d\n", min(temp_long$year), max(temp_long$year)))
cat(sprintf("Memory size: %.1f MB\n", object.size(temp_long) / 1024^2))

head(temp_long)
```

---

## 3.2 Cross-Join with Age Groups

To compute age-specific RR for each day, we create a full panel dataset by cross-joining:

- **Days × GCMs** (temperature observations)
- **Ages** (ERF coefficients)

```{r cross-join-ages}
setDT(coefs_single_age)

# Extract coefficient columns
age_coefs <- coefs_single_age[, .(agegroup, b1, b2, b3, b4, b5)]

# Cartesian join
temp_long[, key := 1]
age_coefs[, key := 1]

dt <- merge(temp_long, age_coefs, by = "key", allow.cartesian = TRUE)
dt[, key := NULL]

cat(sprintf("Full panel: %d observations (days × GCMs × ages)\n", nrow(dt)))
cat(sprintf("Memory size: %.1f MB\n", object.size(dt) / 1024^2))

head(dt)
```

**Warning**: This creates a large dataset (~50 million rows). We will aggregate to annual resolution after computing daily RR.

---

# 4. Temperature-Mortality Model Construction

## 4.1 Define B-Spline Basis

We reconstruct the temperature basis function using the same specification as Masselot et al.:

- **Basis**: B-spline (degree 2)
- **Knots**: Placed at the 10th, 75th, and 90th percentiles of historical temperatures
- **Boundary**: Full historical temperature range

This ensures consistency with the original ERF estimation.

```{r define-basis}
library(dlnm)
library(splines)

varfun    <- "bs"
vardegree <- 2
varper    <- c(10, 75, 90)

# Extract historical temperatures for basis calibration
hist_temps <- unlist(
  proj_data[ssp == "hist", ..gcm_cols],
  use.names = FALSE
)
hist_temps <- hist_temps[!is.na(hist_temps)]

# Define knots and boundaries
varknots <- quantile(hist_temps, varper / 100, na.rm = TRUE)
varbound <- range(hist_temps, na.rm = TRUE)

cat(sprintf("Temperature basis:\n"))
cat(sprintf("  Knots: %.1f°C, %.1f°C, %.1f°C\n", varknots[1], varknots[2], varknots[3]))
cat(sprintf("  Boundary: [%.1f°C, %.1f°C]\n", varbound[1], varbound[2]))

# Store basis arguments
argvar <- list(
  fun = varfun,
  degree = vardegree,
  knots = varknots,
  Bound = varbound
)
```

---

## 4.2 Compute Temperature Basis

We evaluate the B-spline basis at each observed temperature:

$$
\mathbf{B}(T) = [s_1(T), s_2(T), s_3(T), s_4(T), s_5(T)]
$$

```{r compute-basis}
basis <- do.call(onebasis, c(list(x = dt$temp), argvar))

cat(sprintf("Basis matrix: %d observations × %d spline functions\n",
            nrow(basis), ncol(basis)))
```

---

## 4.3 Compute Log-Relative Risk

For each observation (day × age × GCM), we compute:

$$
\log RR = \sum_{j=1}^{5} b_j \cdot s_j(T) = \mathbf{B}(T) \cdot \mathbf{b}
$$

```{r compute-log-rr}
coef_mat <- as.matrix(dt[, .(b1, b2, b3, b4, b5)])

log_rr <- rowSums(basis * coef_mat)
dt[, log_rr := log_rr]

cat(sprintf("Computed log(RR) for %d observations\n", nrow(dt)))

# Clean up large objects
rm(basis, coef_mat, log_rr)
gc()
```

---

## 4.4 Identify Minimum Mortality Temperature (MMT)

The **Minimum Mortality Temperature** is the temperature at which mortality risk is lowest. We compute this separately for each age group by:

1. Evaluating log(RR) across a fine temperature grid
2. Finding the minimum within a plausible range (25th to 99th percentile)

$$
\text{MMT}_a = \arg\min_{T \in [T_{0.25}, T_{0.99}]} \log RR_a(T)
$$

```{r compute-mmt}
temp_seq <- seq(varbound[1], varbound[2], by = 0.1)
basis_seq <- do.call(onebasis, c(list(x = temp_seq), argvar))

mmt_by_age <- dt[, {
  # Get unique coefficients for this age
  coefs <- unique(.SD[, .(b1, b2, b3, b4, b5)])
  coefs_vec <- as.numeric(coefs[1, ])
  
  # Evaluate log(RR) across temperature sequence
  log_rr_seq <- basis_seq %*% coefs_vec
  
  # Restrict to plausible range
  ind <- temp_seq >= quantile(temp_seq, 0.25) &
         temp_seq <= quantile(temp_seq, 0.99)
  
  # Find MMT
  mmt <- temp_seq[ind][which.min(log_rr_seq[ind])]
  
  .(mmt = mmt)
}, by = agegroup]

# Merge back into main dataset
dt <- merge(dt, mmt_by_age, by = "agegroup")

cat("MMT by age (sample):\n")
print(mmt_by_age[as.integer(agegroup) %% 10 == 0 | agegroup %in% c("20", "100")])
```

**Interpretation**: MMT typically increases with age—older adults have higher optimal temperatures, reflecting greater vulnerability to cold.

---

## 4.5 Center at MMT and Compute Relative Risk

To ensure RR = 1.0 at the MMT (no excess risk), we center the log-RR:

$$
\log RR_{\text{centered}} = \mathbf{B}(T) \cdot \mathbf{b} - \mathbf{B}(\text{MMT}) \cdot \mathbf{b}
$$

Then exponentiate and enforce RR ≥ 1:

$$
RR = \max(e^{\log RR_{\text{centered}}}, 1)
$$

```{r compute-rr}
basis_at_temp <- do.call(onebasis, c(list(x = dt$temp), argvar))
basis_at_mmt <- do.call(onebasis, c(list(x = dt$mmt), argvar))

coef_mat <- as.matrix(dt[, .(b1, b2, b3, b4, b5)])

log_rr_centered <- rowSums((basis_at_temp - basis_at_mmt) * coef_mat)

dt[, RR := pmax(exp(log_rr_centered), 1)]

cat(sprintf("Mean RR across all observations: %.3f\n", mean(dt$RR)))
cat(sprintf("RR range: [%.3f, %.3f]\n", min(dt$RR), max(dt$RR)))

# Clean up
rm(basis_at_temp, basis_at_mmt, coef_mat, log_rr_centered)
gc()
```

---

# 5. Adaptation Scenario

We model **gradual adaptation** as a linear reduction in heat vulnerability from 2030 to 2100. The adaptation factor represents the fraction of excess risk that is eliminated:

$$
\text{Adaptation}(y) = \begin{cases}
0 & \text{if } y < 2030 \\
\alpha \cdot \frac{y - 2030}{2100 - 2030} & \text{if } 2030 \le y \le 2100 \\
\alpha & \text{if } y > 2100
\end{cases}
$$

where $\alpha$ is the `adaptation_level` (e.g., 0.5 = 50% reduction by 2100).

The adapted RR is:

$$
RR_{\text{adapted}} = 1 + (RR - 1) \times (1 - \text{Adaptation})
$$

This preserves the baseline risk (RR = 1 at MMT) while reducing excess heat/cold mortality over time.

```{r apply-adaptation}
dt[, adapt_factor := fifelse(
  year < 2030, 0,
  fifelse(year > 2100, adaptation_level,
          adaptation_level * (year - 2030) / (2100 - 2030))
)]

dt[, RR_adapted := 1 + (RR - 1) * (1 - adapt_factor)]

cat(sprintf("Adaptation factor by year:\n"))
cat(sprintf("  2020: %.2f (%.0f%% reduction)\n", 
            unique(dt[year == 2020, adapt_factor]), 
            unique(dt[year == 2020, adapt_factor]) * 100))
cat(sprintf("  2050: %.2f (%.0f%% reduction)\n", 
            unique(dt[year == 2050, adapt_factor]),
            unique(dt[year == 2050, adapt_factor]) * 100))
cat(sprintf("  2080: %.2f (%.0f%% reduction)\n", 
            unique(dt[year == 2080, adapt_factor]),
            unique(dt[year == 2080, adapt_factor]) * 100))
cat(sprintf("  2100: %.2f (%.0f%% reduction)\n", 
            unique(dt[year == 2100, adapt_factor]),
            unique(dt[year == 2100, adapt_factor]) * 100))
```

---

# 6. Annual Aggregation

To reduce dataset size and facilitate comparison with baseline, we aggregate daily RR to **annual means** by age and GCM:

$
\overline{RR}_{a,y,g} = \frac{1}{365} \sum_{d=1}^{365} RR_{a,d,y,g}
$

This represents the average mortality risk for each age, year, and climate model.

```{r aggregate-annual}
dt_annual <- dt[, .(
  RR_mean = mean(RR_adapted, na.rm = TRUE)
), by = .(year, agegroup, gcm)]

cat(sprintf("Annual dataset: %d observations (years × ages × GCMs)\n", nrow(dt_annual)))
cat(sprintf("Memory size: %.1f MB\n", object.size(dt_annual) / 1024^2))

head(dt_annual, 20)

# Clean up daily data to free memory
rm(dt, basis_seq)
gc()
```

---

# 7. Baseline Mortality Risk Computation

To compute mortality multipliers, we need to compare future RR against a **historical baseline**. We now apply the same ERF model to historical temperatures.

## 7.1 Extract Historical Temperature Data

```{r baseline-temperature-data}
baseline_period <- proj_data[ssp == "hist"]

baseline_temps <- melt(
  baseline_period,
  id.vars = c("URAU_CODE", "date", "year", "doy"),
  measure.vars = gcm_cols,
  variable.name = "gcm",
  value.name = "temp"
)

cat(sprintf("Baseline temperature data:\n"))
cat(sprintf("  Period: %d–%d\n", min(baseline_temps$year), max(baseline_temps$year)))
cat(sprintf("  Observations: %s (days × GCMs)\n", 
            format(nrow(baseline_temps), big.mark = ",")))
cat(sprintf("  Temperature range: %.1f°C to %.1f°C\n",
            min(baseline_temps$temp, na.rm = TRUE),
            max(baseline_temps$temp, na.rm = TRUE)))
```

---

## 7.2 Cross-Join Baseline with Age Groups

```{r baseline-cross-join}
baseline_temps[, key := 1]
age_coefs[, key := 1]

dt_baseline <- merge(baseline_temps, age_coefs, by = "key", allow.cartesian = TRUE)
dt_baseline[, key := NULL]

cat(sprintf("Baseline panel: %s observations\n", format(nrow(dt_baseline), big.mark = ",")))
cat(sprintf("Memory size: %.1f MB\n", object.size(dt_baseline) / 1024^2))
```

---

## 7.3 Compute Baseline RR

Apply the same ERF model to historical temperatures (no adaptation for baseline).

```{r compute-baseline-rr}
# Compute basis
basis_baseline <- do.call(onebasis, c(list(x = dt_baseline$temp), argvar))
coef_mat_baseline <- as.matrix(dt_baseline[, .(b1, b2, b3, b4, b5)])

# Merge MMT
dt_baseline <- merge(dt_baseline, mmt_by_age, by = "agegroup")

# Compute centered RR
basis_at_temp_bl <- do.call(onebasis, c(list(x = dt_baseline$temp), argvar))
basis_at_mmt_bl <- do.call(onebasis, c(list(x = dt_baseline$mmt), argvar))

log_rr_centered_bl <- rowSums((basis_at_temp_bl - basis_at_mmt_bl) * coef_mat_baseline)
dt_baseline[, RR := pmax(exp(log_rr_centered_bl), 1)]

cat(sprintf("Baseline RR computed for %s observations\n", 
            format(nrow(dt_baseline), big.mark = ",")))
cat(sprintf("Mean baseline RR: %.4f\n", mean(dt_baseline$RR)))

# Clean up
rm(basis_baseline, basis_at_temp_bl, basis_at_mmt_bl, coef_mat_baseline, log_rr_centered_bl)
gc()
```

---

## 7.4 Aggregate Baseline to Annual Means

```{r baseline-annual}
rr_baseline <- dt_baseline[, .(
  RR_baseline = mean(RR, na.rm = TRUE),
  n_obs = .N
), by = agegroup]

cat(sprintf("\n=== Baseline RR Summary ===\n"))
cat(sprintf("Ages: %s\n", length(unique(rr_baseline$agegroup))))
cat(sprintf("Mean RR across all ages: %.4f\n", mean(rr_baseline$RR_baseline)))

# Sample output
print(rr_baseline[as.integer(agegroup) %% 10 == 0 | agegroup %in% c("20", "100")])

# Clean up baseline daily data
rm(dt_baseline, baseline_temps)
gc()
```

**Interpretation**: These values represent the average temperature-related mortality risk during the historical climate. Values close to 1 indicate low baseline risk.

---

# 8. Mortality Multipliers

Now we compute the **mortality multiplier** by comparing future annual RR to baseline RR:

$$
M_{a,y,g} = \frac{\overline{RR}_{a,y,g,\text{future}}}{\overline{RR}_{a,\text{baseline}}}
$$

## 8.1 Merge Future and Baseline RR

```{r merge-future-baseline}
# Merge baseline into future annual data
dt_annual_mult <- merge(
  dt_annual,
  rr_baseline,
  by = "agegroup",
  all.x = TRUE
)

cat(sprintf("Merged dataset: %s observations\n", format(nrow(dt_annual_mult), big.mark = ",")))
```

---

## 8.2 Compute Multipliers

```{r compute-multipliers}
# Compute multiplier for each year × age × GCM
dt_annual_mult[, multiplier := RR_mean / RR_baseline]

cat(sprintf("Multipliers computed for %s observations\n", 
            format(nrow(dt_annual_mult), big.mark = ",")))
cat(sprintf("Multiplier range: %.3f to %.3f\n", 
            min(dt_annual_mult$multiplier), max(dt_annual_mult$multiplier)))

head(dt_annual_mult[, .(year, agegroup, gcm, RR_mean, RR_baseline, multiplier)], 20)
```

---

## 8.3 Aggregate Across GCMs (Ensemble Statistics)

```{r ensemble-multipliers}
multipliers <- dt_annual_mult[, .(
  multiplier_mean = mean(multiplier, na.rm = TRUE),
  multiplier_sd   = sd(multiplier, na.rm = TRUE),
  multiplier_q05  = quantile(multiplier, 0.05, na.rm = TRUE),
  multiplier_q50  = quantile(multiplier, 0.50, na.rm = TRUE),
  multiplier_q95  = quantile(multiplier, 0.95, na.rm = TRUE),
  n_gcms          = .N
), by = .(year, agegroup)]

# Convert age to integer for easier analysis
multipliers[, age := as.integer(agegroup)]
setorder(multipliers, year, age)

cat(sprintf("\n=== Ensemble Multipliers ===\n"))
cat(sprintf("Years: %d–%d\n", min(multipliers$year), max(multipliers$year)))
cat(sprintf("Ages: %d–%d\n", min(multipliers$age), max(multipliers$age)))
cat(sprintf("Total records: %s\n", format(nrow(multipliers), big.mark = ",")))

head(multipliers, 20)
```

**Interpretation**:
- `multiplier_mean > 1`: Increased mortality risk compared to baseline
- `multiplier_mean < 1`: Decreased risk (e.g., milder winters)
- `multiplier_q05` to `multiplier_q95`: 90% uncertainty range across GCMs

---

## 8.4 Summary Statistics by Decade

```{r multiplier-summary}
# Compute decadal averages
multipliers[, decade := floor(year / 10) * 10]

decade_summary <- multipliers[, .(
  mean_multiplier = round(mean(multiplier_mean), 4),
  min_multiplier  = round(min(multiplier_mean), 4),
  max_multiplier  = round(max(multiplier_mean), 4),
  mult_age_30     = round(multiplier_mean[age == 30][1], 4),
  mult_age_65     = round(multiplier_mean[age == 65][1], 4),
  mult_age_85     = round(multiplier_mean[age == 85][1], 4)
), by = decade]

cat("\n=== Mortality Multipliers by Decade ===\n")
print(decade_summary)
```

---

## 8.5 Visualization: Multipliers Over Time and Age

```{r plot-multipliers-heatmap, fig.width=10, fig.height=6}
library(ggplot2)

ggplot(multipliers[year %% 5 == 0], aes(x = age, y = factor(year), fill = multiplier_mean)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 1,
    name = "Multiplier",
    limits = c(min(multipliers$multiplier_mean), max(multipliers$multiplier_mean))
  ) +
  labs(
    title = sprintf("Mortality Multipliers — %s (SSP%d)", city, scenario),
    subtitle = sprintf("Adaptation: %.0f%% by 2100", adaptation_level * 100),
    x = "Age",
    y = "Year",
    caption = "Relative to historical baseline (1991–2020)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

---

## 8.6 Visualization: Time Series for Key Ages

```{r plot-multipliers-timeseries, fig.width=10, fig.height=6}
key_ages <- c(30, 50, 70, 90)

ggplot(multipliers[age %in% key_ages], 
       aes(x = year, y = multiplier_mean, color = factor(age))) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = multiplier_q05, ymax = multiplier_q95, fill = factor(age)),
              alpha = 0.2, color = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30") +
  scale_color_viridis_d(name = "Age") +
  scale_fill_viridis_d(name = "Age") +
  labs(
    title = sprintf("Mortality Multipliers Over Time — %s (SSP%d)", city, scenario),
    subtitle = "Shaded regions: 5th–95th percentile across GCMs",
    x = "Year",
    y = "Mortality Multiplier (relative to baseline)",
    caption = sprintf("Adaptation: %.0f%% by 2100", adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 8.7 Export Multipliers

```{r export-multipliers}
# Create output directory if needed
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

# Save full multipliers table
mult_path <- sprintf("results/mortality_multipliers_%s_ssp%d.csv",
                     tolower(gsub(" ", "_", city)),
                     scenario)

fwrite(multipliers, mult_path)
cat(sprintf("\nSaved: %s\n", mult_path))

# Save decade summary
decade_path <- sprintf("results/multipliers_decade_summary_%s_ssp%d.csv",
                       tolower(gsub(" ", "_", city)),
                       scenario)

fwrite(decade_summary, decade_path)
cat(sprintf("Saved: %s\n", decade_path))
```

---

# 9. Baseline Mortality Data

To construct life tables, we need **baseline mortality rates** by age and year. These represent the expected mortality in the absence of climate change impacts.

We use **Eurostat mortality projections** (EUROPOP2023), which provide age-specific death rates for European countries through 2100.

## 9.1 Load Eurostat Mortality Projections

```{r load-baseline-mortality}
# Eurostat mortality projections for Austria
# Data source: Eurostat demo_mprojmorr (EUROPOP2023)
# 
# If you don't have this file, you can download it from:
# https://ec.europa.eu/eurostat/databrowser/view/demo_mprojmorr/

# For this demo, we'll load pre-processed data
mort_proj_path <- "../data/mortality_projections_austria.csv"

if (!file.exists(mort_proj_path)) {
  cat("Warning: Baseline mortality file not found.\n")
  cat("Please download Eurostat EUROPOP2023 projections.\n")
  cat("Using synthetic data for demonstration...\n")
  
  # Create synthetic baseline mortality (Gompertz model)
  # mx = a * exp(b * age)
  # Typical values for high-income countries
  ages_synth <- 20:100
  years_synth <- 2015:2100
  
  mort_proj <- CJ(age = ages_synth, year = years_synth)
  
  # Gompertz parameters (calibrated to approximate Austria)
  a <- 0.0001
  b <- 0.08
  
  # Add slight mortality improvement over time (1% per decade)
  mort_proj[, mx := a * exp(b * age) * (0.99 ^ ((year - 2015) / 10))]
  
} else {
  mort_proj <- fread(mort_proj_path)
  
  # Filter to relevant ages and years
  mort_proj <- mort_proj[age >= 20 & age <= 100]
  mort_proj <- mort_proj[year >= 2015 & year <= 2100]
}

# Ensure proper ordering
setkey(mort_proj, year, age)

cat(sprintf("Baseline mortality data:\n"))
cat(sprintf("  Ages: %d–%d\n", min(mort_proj$age), max(mort_proj$age)))
cat(sprintf("  Years: %d–%d\n", min(mort_proj$year), max(mort_proj$year)))
cat(sprintf("  Records: %s\n", format(nrow(mort_proj), big.mark = ",")))

head(mort_proj, 10)
```

**Data structure**:
- `age`: Single-year ages (20–100)
- `year`: Calendar year (2015–2100)
- `mx`: Central death rate (annual probability of death)

**Interpretation**: `mx = 0.001` means 1 death per 1,000 person-years (0.1% annual mortality)

---

## 9.2 Baseline Mortality Summary

```{r baseline-mort-summary}
# Sample mortality rates for selected ages and years
sample_years <- c(2020, 2050, 2080)
sample_ages <- c(30, 50, 70, 90)

mort_sample <- mort_proj[year %in% sample_years & age %in% sample_ages]
mort_wide <- dcast(mort_sample, age ~ year, value.var = "mx")

cat("\n=== Baseline Mortality Rates (mx) ===\n")
print(mort_wide)

# Visualize baseline mortality by age (for selected years)
library(ggplot2)

ggplot(mort_proj[year %in% sample_years], aes(x = age, y = mx, color = factor(year))) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_viridis_d(name = "Year") +
  labs(
    title = "Baseline Mortality Rates by Age",
    subtitle = "Eurostat EUROPOP2023 projections (no climate impacts)",
    x = "Age",
    y = "Central Death Rate (mx)",
    caption = "Log scale"
  ) +
  theme_minimal(base_size = 12)
```

---

# 10. Apply Climate Multipliers to Baseline Mortality

We now combine:

1. **Baseline mortality**: `mx_base[age, year]` from Eurostat
2. **Climate multipliers**: `M[age, year]` from our temperature-mortality model

To obtain **climate-adjusted mortality**:

$$
mx_{\text{climate}}(a, y) = mx_{\text{baseline}}(a, y) \times M(a, y)
$$

where $M(a, y) > 1$ indicates increased mortality due to climate change.

---

## 10.1 Merge Multipliers with Baseline Mortality

```{r merge-multipliers}
# Use ensemble mean multiplier (already averaged across GCMs)
mult_subset <- multipliers[, .(age, year, multiplier = multiplier_mean)]

# Merge with baseline mortality
mort_climate <- merge(
  mort_proj,
  mult_subset,
  by = c("age", "year"),
  all.x = TRUE
)

# For years/ages without multipliers, assume no climate impact
mort_climate[is.na(multiplier), multiplier := 1.0]

# Compute climate-adjusted mortality
mort_climate[, mx_climate := mx * multiplier]

cat(sprintf("Climate-adjusted mortality computed for %s records\n",
            format(nrow(mort_climate), big.mark = ",")))

head(mort_climate, 15)
```

---

## 10.2 Comparison: Baseline vs Climate Mortality

```{r compare-mortality}
# Summary statistics
comparison <- mort_climate[year >= 2050, .(
  mean_mx_base    = mean(mx),
  mean_mx_climate = mean(mx_climate),
  mean_multiplier = mean(multiplier),
  max_multiplier  = max(multiplier)
), by = year]

cat("\n=== Mortality Comparison: Baseline vs Climate (2050+) ===\n")
print(comparison[year %in% c(2050, 2070, 2090, 2099)])

# Visualize the impact
ggplot(mort_climate[year %in% c(2050, 2070, 2099)], 
       aes(x = age)) +
  geom_line(aes(y = mx, color = "Baseline"), linewidth = 1) +
  geom_line(aes(y = mx_climate, color = "Climate-adjusted"), 
            linewidth = 1, linetype = "dashed") +
  facet_wrap(~ year, ncol = 3) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_manual(
    name = "Scenario",
    values = c("Baseline" = "#1f77b4", "Climate-adjusted" = "#d62728")
  ) +
  labs(
    title = sprintf("Mortality Impact of Climate Change — %s (SSP%d)", city, scenario),
    subtitle = sprintf("Adaptation: %.0f%% by 2100", adaptation_level * 100),
    x = "Age",
    y = "Central Death Rate (mx)",
    caption = "Log scale"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Key observation**: Climate multipliers increase mortality, especially at older ages in later years.

---

# 11. Cohort Life Table Construction

A **cohort life table** tracks a hypothetical birth cohort through time, applying age-specific mortality rates for each calendar year they live through.

For a cohort born in year $Y_0$:
- At age 20 (year $Y_0 + 20$), use mortality for year $Y_0 + 20$, age 20
- At age 21 (year $Y_0 + 21$), use mortality for year $Y_0 + 21$, age 21
- ... and so on until the cohort dies out

This differs from a **period life table**, which uses mortality rates from a single year.

---

## 11.1 Define Cohort Parameters

```{r cohort-setup}
cohort_birth_year <- 2020  # Birth year of cohort
cohort_start_age  <- 20    # Start tracking at age 20 (year 2040)
radix             <- 100000  # Initial cohort size (lx at start age)

cat(sprintf("Cohort Parameters:\n"))
cat(sprintf("  Birth year: %d\n", cohort_birth_year))
cat(sprintf("  Start tracking at age %d (year %d)\n", 
            cohort_start_age, cohort_birth_year + cohort_start_age))
cat(sprintf("  Initial cohort size (radix): %s\n", format(radix, big.mark = ",")))
```

---

## 11.2 Build Cohort Life Table Function

```{r lifetable-function}
build_cohort_lifetable <- function(mort_data, birth_year, start_age, radix = 100000) {
  # Filter to cohort's age-year trajectory
  # For someone born in 'birth_year', at age 'a' they are in year 'birth_year + a'
  
  mort_data <- copy(mort_data)
  mort_data[, cohort_year := year - age]
  
  cohort_data <- mort_data[cohort_year == birth_year & age >= start_age]
  setorder(cohort_data, age)
  
  if (nrow(cohort_data) == 0) {
    stop("No data found for this cohort")
  }
  
  # Build life table columns
  n <- nrow(cohort_data)
  
  # Convert central death rate (mx) to probability of death (qx)
  # Standard approximation: qx ≈ mx / (1 + 0.5 * mx)
  # For small mx, qx ≈ 1 - exp(-mx) is also valid
  cohort_data[, qx_base := mx / (1 + 0.5 * mx)]
  cohort_data[, qx_climate := mx_climate / (1 + 0.5 * mx_climate)]
  
  # Ensure qx does not exceed 1
  cohort_data[qx_base > 1, qx_base := 1]
  cohort_data[qx_climate > 1, qx_climate := 1]
  
  # Initialize survivors
  cohort_data[, lx_base := as.numeric(NA)]
  cohort_data[, lx_climate := as.numeric(NA)]
  
  cohort_data[1, lx_base := radix]
  cohort_data[1, lx_climate := radix]
  
  # Iteratively compute survivors and deaths
  for (i in 1:n) {
    if (i < n) {
      # Baseline
      cohort_data[i, dx_base := lx_base * qx_base]
      cohort_data[i + 1, lx_base := cohort_data[i, lx_base - dx_base]]
      
      # Climate
      cohort_data[i, dx_climate := lx_climate * qx_climate]
      cohort_data[i + 1, lx_climate := cohort_data[i, lx_climate - dx_climate]]
    } else {
      # Last age: all remaining die
      cohort_data[i, dx_base := lx_base]
      cohort_data[i, dx_climate := lx_climate]
    }
  }
  
  # Person-years lived (Lx): approximation using average of lx and lx+1
  cohort_data[, Lx_base := (lx_base + c(lx_base[-1], 0)) / 2]
  cohort_data[, Lx_climate := (lx_climate + c(lx_climate[-1], 0)) / 2]
  
  # For last age, use lx / mx (if mx > 0)
  cohort_data[n, Lx_base := ifelse(mx > 0, lx_base / mx, lx_base)]
  cohort_data[n, Lx_climate := ifelse(mx_climate > 0, lx_climate / mx_climate, lx_climate)]
  
  # Total person-years from age x onwards (Tx)
  cohort_data[, Tx_base := rev(cumsum(rev(Lx_base)))]
  cohort_data[, Tx_climate := rev(cumsum(rev(Lx_climate)))]
  
  # Life expectancy at age x (ex)
  cohort_data[, ex_base := Tx_base / lx_base]
  cohort_data[, ex_climate := Tx_climate / lx_climate]
  
  return(cohort_data)
}
```

---

## 11.3 Construct Baseline and Climate Life Tables

```{r build-lifetable}
lifetable <- build_cohort_lifetable(
  mort_climate,
  birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  radix = radix
)

cat(sprintf("Cohort life table constructed:\n"))
cat(sprintf("  Ages: %d–%d\n", min(lifetable$age), max(lifetable$age)))
cat(sprintf("  Years: %d–%d\n", min(lifetable$year), max(lifetable$year)))

# Display key columns
lifetable_display <- lifetable[, .(
  age, year,
  mx_base = round(mx, 6),
  mx_climate = round(mx_climate, 6),
  multiplier = round(multiplier, 3),
  qx_base = round(qx_base, 5),
  qx_climate = round(qx_climate, 5),
  lx_base = round(lx_base, 0),
  lx_climate = round(lx_climate, 0),
  ex_base = round(ex_base, 2),
  ex_climate = round(ex_climate, 2)
)]

cat("\nLife Table (first 20 rows):\n")
print(head(lifetable_display, 20))

cat("\nLife Table (last 10 rows):\n")
print(tail(lifetable_display, 10))
```

---

# 12. Life Expectancy Analysis

## 12.1 Life Expectancy at Age 20

```{r life-expectancy-20}
e20_base <- lifetable[age == cohort_start_age, ex_base]
e20_climate <- lifetable[age == cohort_start_age, ex_climate]

delta_e20 <- e20_climate - e20_base
pct_delta_e20 <- (delta_e20 / e20_base) * 100

cat(sprintf("\n=== Life Expectancy at Age %d ===\n", cohort_start_age))
cat(sprintf("  Baseline (no climate impact):   %.2f years\n", e20_base))
cat(sprintf("  Climate-adjusted (SSP%d):        %.2f years\n", scenario, e20_climate))
cat(sprintf("  Difference:                     %.2f years (%.2f%%)\n", 
            delta_e20, pct_delta_e20))

if (delta_e20 < 0) {
  cat(sprintf("\n  → Climate change reduces life expectancy by %.1f months\n", 
              abs(delta_e20) * 12))
} else {
  cat(sprintf("\n  → Climate change increases life expectancy by %.1f months\n", 
              delta_e20 * 12))
  cat("     (Note: This may occur if cold-related mortality reduction exceeds heat increase)\n")
}
```

---

## 12.2 Life Expectancy by Age

```{r life-expectancy-by-age}
# Plot life expectancy across ages
ex_data <- lifetable[, .(
  age,
  Baseline = ex_base,
  Climate = ex_climate
)]

ex_long <- melt(ex_data, id.vars = "age", 
                variable.name = "Scenario", value.name = "ex")

ggplot(ex_long, aes(x = age, y = ex, color = Scenario)) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = sprintf("Life Expectancy by Age — Cohort Born %d", cohort_birth_year),
    subtitle = sprintf("%s, SSP%d, Adaptation %.0f%%", city, scenario, adaptation_level * 100),
    x = "Age",
    y = "Remaining Life Expectancy (years)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 12.3 Survivorship Curves

```{r survivorship-curves}
surv_data <- lifetable[, .(
  age,
  Baseline = lx_base,
  Climate = lx_climate
)]

surv_long <- melt(surv_data, id.vars = "age",
                  variable.name = "Scenario", value.name = "lx")

ggplot(surv_long, aes(x = age, y = lx / radix, color = Scenario)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = sprintf("Survivorship Curve — Cohort Born %d", cohort_birth_year),
    subtitle = sprintf("Starting with %s people at age %d",
                       format(radix, big.mark = ","), cohort_start_age),
    x = "Age",
    y = "Proportion Surviving",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Interpretation**: The gap between curves represents excess deaths due to climate change.

---

## 12.4 Cumulative Deaths

```{r cumulative-deaths}
# Calculate cumulative deaths by age
lifetable[, cumul_deaths_base := cumsum(dx_base)]
lifetable[, cumul_deaths_climate := cumsum(dx_climate)]
lifetable[, excess_deaths := cumul_deaths_climate - cumul_deaths_base]

# Total excess deaths over lifetime
total_excess <- lifetable[age == max(age), excess_deaths]
pct_excess <- (total_excess / radix) * 100

cat(sprintf("\n=== Cumulative Excess Deaths ===\n"))
cat(sprintf("  Initial cohort size: %s\n", format(radix, big.mark = ",")))
cat(sprintf("  Total excess deaths: %.0f (%.2f%% of cohort)\n", 
            total_excess, pct_excess))

# Plot excess deaths over age
ggplot(lifetable, aes(x = age, y = excess_deaths)) +
  geom_area(fill = "#d62728", alpha = 0.6) +
  geom_line(color = "#d62728", linewidth = 1) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Cumulative Excess Deaths Due to Climate Change",
    subtitle = sprintf("Cohort born %d, starting with %s at age %d",
                       cohort_birth_year, format(radix, big.mark = ","), 
                       cohort_start_age),
    x = "Age",
    y = "Cumulative Excess Deaths",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12)
```

---

# 13. Export Life Table Results

```{r export-lifetable}
# Save full life table
output_path <- sprintf("results/lifetable_%s_ssp%d_cohort%d.csv",
                       tolower(gsub(" ", "_", city)),
                       scenario,
                       cohort_birth_year)

lifetable_export <- lifetable[, .(
  age, year,
  mx, mx_climate, multiplier,  # ← Changed mx_base to mx
  qx_base, qx_climate,
  lx_base, lx_climate,
  dx_base, dx_climate,
  ex_base, ex_climate
)]

fwrite(lifetable_export, output_path)
cat(sprintf("\nSaved: %s\n", output_path))

# Save summary statistics
summary_stats <- data.table(
  city = city,
  scenario = sprintf("SSP%d", scenario),
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  start_age = cohort_start_age,
  radix = radix,
  e20_baseline = e20_base,
  e20_climate = e20_climate,
  delta_e20_years = delta_e20,
  delta_e20_months = delta_e20 * 12,
  pct_change_e20 = pct_delta_e20,
  total_excess_deaths = total_excess,
  pct_excess_deaths = pct_excess
)

summary_path <- sprintf("results/lifetable_summary_%s_ssp%d.csv",
                        tolower(gsub(" ", "_", city)),
                        scenario)

fwrite(summary_stats, summary_path)
cat(sprintf("Saved: %s\n", summary_path))
```

---

# 14. Lifespan Inequality Analysis

Lifespan inequality (LI) measures the **dispersion** or **spread** of ages at death within a population. Unlike life expectancy (which is a mean), LI captures how equally or unequally distributed lifespans are.

**Key insight**: Two populations can have identical life expectancy but very different inequality:
- **Low LI**: Deaths concentrated around a narrow age range (e.g., most people die 75-85)
- **High LI**: Deaths spread across a wide age range (e.g., deaths from 60-100)

Climate change affects LI through the **age pattern** of excess deaths, not just their total number.

---

## 14.1 Age-at-Death Distribution

The life table column `dx` represents the **distribution of deaths by age**. From this, we can compute various inequality metrics.

```{r age-at-death-distribution}
# Extract age-at-death distributions
deaths_base <- lifetable[, .(age, dx = dx_base)]
deaths_climate <- lifetable[, .(age, dx = dx_climate)]

# Normalize to proportions (for visualization)
deaths_base[, prop := dx / sum(dx)]
deaths_climate[, prop := dx / sum(dx)]

cat(sprintf("\n=== Age-at-Death Distribution ===\n"))
cat(sprintf("Baseline deaths: %s\n", format(round(sum(deaths_base$dx)), big.mark = ",")))
cat(sprintf("Climate deaths: %s\n", format(round(sum(deaths_climate$dx)), big.mark = ",")))
cat(sprintf("Total excess deaths: %s\n", format(round(sum(deaths_climate$dx) - sum(deaths_base$dx)), big.mark = ",")))
```

---

## 14.2 Lifespan Inequality Metrics

### 14.2.1 Gini Coefficient

The **Gini coefficient** measures inequality on a 0-1 scale:
- **0**: Perfect equality (everyone dies at the same age)
- **1**: Maximum inequality (all deaths at extremes)

For ages at death, Gini is computed as:

$$
G = \frac{\sum_{i=1}^{n} \sum_{j=1}^{n} |x_i - x_j| \cdot w_i \cdot w_j}{2 \cdot \bar{x} \cdot \sum_{i=1}^{n} w_i^2}
$$

where $$x_i$$ = age, $$w_i$$ = deaths at that age.

```{r compute-gini}
# Gini coefficient function
compute_gini <- function(ages, deaths) {
  # Remove zero-death ages
  valid <- deaths > 0
  ages <- ages[valid]
  deaths <- deaths[valid]
  
  n <- length(ages)
  total_deaths <- sum(deaths)
  
  # Weighted mean age at death
  mean_age <- sum(ages * deaths) / total_deaths
  
  # Compute pairwise absolute differences
  numerator <- 0
  for (i in 1:n) {
    for (j in 1:n) {
      numerator <- numerator + abs(ages[i] - ages[j]) * deaths[i] * deaths[j]
    }
  }
  
  denominator <- 2 * mean_age * total_deaths^2
  
  gini <- numerator / denominator
  return(gini)
}

# Compute Gini for both scenarios
gini_base <- compute_gini(deaths_base$age, deaths_base$dx)
gini_climate <- compute_gini(deaths_climate$age, deaths_climate$dx)

delta_gini <- gini_climate - gini_base
pct_change_gini <- (delta_gini / gini_base) * 100

cat(sprintf("\n=== Gini Coefficient ===\n"))
cat(sprintf("Baseline:         %.4f\n", gini_base))
cat(sprintf("Climate-adjusted: %.4f\n", gini_climate))
cat(sprintf("Change:           %.4f (%.2f%%)\n", delta_gini, pct_change_gini))

if (delta_gini > 0) {
  cat("\n→ Climate change INCREASES lifespan inequality\n")
  cat("  (deaths become more unequally distributed)\n")
} else {
  cat("\n→ Climate change DECREASES lifespan inequality\n")
  cat("  (deaths become more equally distributed)\n")
}
```

---

### 14.2.2 Standard Deviation of Age at Death

The **standard deviation** (SD) measures dispersion in years:

$$
\text{SD} = \sqrt{\frac{\sum (x_i - \bar{x})^2 \cdot d_i}{\sum d_i}}
$$

where $$x_i$$ = age, $$d_i$$ = deaths, $$\bar{x}$$ = mean age at death.

```{r compute-sd}
# Standard deviation function
compute_sd_age_at_death <- function(ages, deaths) {
  total_deaths <- sum(deaths)
  mean_age <- sum(ages * deaths) / total_deaths
  variance <- sum((ages - mean_age)^2 * deaths) / total_deaths
  sd <- sqrt(variance)
  return(sd)
}

# Compute SD for both scenarios
sd_base <- compute_sd_age_at_death(deaths_base$age, deaths_base$dx)
sd_climate <- compute_sd_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_sd <- sd_climate - sd_base
pct_change_sd <- (delta_sd / sd_base) * 100

cat(sprintf("\n=== Standard Deviation of Age at Death ===\n"))
cat(sprintf("Baseline:         %.2f years\n", sd_base))
cat(sprintf("Climate-adjusted: %.2f years\n", sd_climate))
cat(sprintf("Change:           %.2f years (%.2f%%)\n", delta_sd, pct_change_sd))
```

**Interpretation**: SD quantifies how "spread out" deaths are. Higher SD means ages at death are more dispersed.

---

### 14.2.3 Interquartile Range (IQR)

The **IQR** is the difference between the 75th and 25th percentile of age at death:

$$
\text{IQR} = Q_{75} - Q_{25}
$$

This measures the spread of the "middle 50%" of deaths.

```{r compute-iqr}
# IQR function
compute_iqr_age_at_death <- function(ages, deaths) {
  # Create cumulative distribution
  cumsum_deaths <- cumsum(deaths)
  total_deaths <- sum(deaths)
  
  # Find 25th percentile
  q25_idx <- which(cumsum_deaths >= 0.25 * total_deaths)[1]
  q25 <- ages[q25_idx]
  
  # Find 75th percentile
  q75_idx <- which(cumsum_deaths >= 0.75 * total_deaths)[1]
  q75 <- ages[q75_idx]
  
  iqr <- q75 - q25
  return(list(q25 = q25, q75 = q75, iqr = iqr))
}

# Compute IQR for both scenarios
iqr_base <- compute_iqr_age_at_death(deaths_base$age, deaths_base$dx)
iqr_climate <- compute_iqr_age_at_death(deaths_climate$age, deaths_climate$dx)

delta_iqr <- iqr_climate$iqr - iqr_base$iqr

cat(sprintf("\n=== Interquartile Range (IQR) ===\n"))
cat(sprintf("Baseline:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_base$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_base$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_base$iqr))

cat(sprintf("\nClimate-adjusted:\n"))
cat(sprintf("  Q25 (25th percentile): %.1f years\n", iqr_climate$q25))
cat(sprintf("  Q75 (75th percentile): %.1f years\n", iqr_climate$q75))
cat(sprintf("  IQR: %.1f years\n", iqr_climate$iqr))

cat(sprintf("\nChange in IQR: %.2f years\n", delta_iqr))
```

---

## 14.3 Threshold Analysis

A critical insight from demography: there is a **threshold age** that determines whether additional deaths increase or decrease lifespan inequality.

**Rule of thumb**: The threshold is approximately where **remaining life expectancy** equals the population's **life expectancy at the starting age**.

$$
\text{Threshold} \approx \text{age where } e_x \approx e_{20}
$$

- **Deaths below threshold**: Increase LI (premature deaths)
- **Deaths above threshold**: Decrease LI (late deaths compress distribution)

### 14.3.1 Identify Threshold Age

```{r identify-threshold}
# Threshold: age where remaining LE ≈ LE at starting age
e20_base_val <- lifetable[age == cohort_start_age, ex_base]

# Find age closest to this remaining LE
lifetable[, diff_from_e20 := abs(ex_base - e20_base_val)]
threshold_age <- lifetable[which.min(diff_from_e20), age]

cat(sprintf("\n=== Threshold Age Analysis ===\n"))
cat(sprintf("Life expectancy at age %d: %.2f years\n", cohort_start_age, e20_base_val))
cat(sprintf("Threshold age: %d years\n", threshold_age))
cat(sprintf("Remaining LE at threshold: %.2f years\n", 
            lifetable[age == threshold_age, ex_base]))
```

---

### 14.3.2 Classify Excess Deaths by Threshold

```{r classify-excess-deaths}
# Add threshold classification to life table
lifetable[, below_threshold := age < threshold_age]
lifetable[, excess_deaths := dx_climate - dx_base]

# Summarize excess deaths by position relative to threshold
excess_summary <- lifetable[, .(
  total_excess = sum(excess_deaths),
  excess_below = sum(excess_deaths[below_threshold == TRUE]),
  excess_above = sum(excess_deaths[below_threshold == FALSE])
)]

pct_below <- (excess_summary$excess_below / excess_summary$total_excess) * 100
pct_above <- (excess_summary$excess_above / excess_summary$total_excess) * 100

cat(sprintf("\n=== Excess Deaths Classification ===\n"))
cat(sprintf("Total excess deaths: %.0f\n", excess_summary$total_excess))
cat(sprintf("\nBelow threshold (age < %d):\n", threshold_age))
cat(sprintf("  Excess deaths: %.0f (%.1f%%)\n", excess_summary$excess_below, pct_below))
cat(sprintf("  Effect: INCREASES lifespan inequality\n"))

cat(sprintf("\nAbove threshold (age ≥ %d):\n", threshold_age))
cat(sprintf("  Excess deaths: %.0f (%.1f%%)\n", excess_summary$excess_above, pct_above))
cat(sprintf("  Effect: DECREASES lifespan inequality\n"))

# Net effect interpretation
if (excess_summary$excess_below > abs(excess_summary$excess_above)) {
  cat(sprintf("\n→ NET EFFECT: Climate change increases lifespan inequality\n"))
  cat(sprintf("   (more excess deaths below threshold)\n"))
} else {
  cat(sprintf("\n→ NET EFFECT: Climate change decreases lifespan inequality\n"))
  cat(sprintf("   (more excess deaths above threshold)\n"))
}
```

**Interpretation**: This classification reveals **where** climate impacts mortality distribution, not just **how much**.

---

## 14.4 Visualizations

### 14.4.1 Age-at-Death Distribution Comparison

```{r plot-age-at-death, fig.width=10, fig.height=6}
# Combine distributions for plotting
dist_combined <- rbind(
  deaths_base[, .(age, deaths = dx, scenario = "Baseline")],
  deaths_climate[, .(age, deaths = dx, scenario = "Climate")]
)

# Calculate proportions for clean visualization
dist_combined[, prop := deaths / sum(deaths), by = scenario]

ggplot(dist_combined, aes(x = age, y = prop, fill = scenario)) +
  geom_area(alpha = 0.6, position = "identity") +
  geom_vline(xintercept = threshold_age, linetype = "dashed", 
             color = "black", linewidth = 1) +
  annotate("text", x = threshold_age + 2, y = max(dist_combined$prop) * 0.9,
           label = sprintf("Threshold\n(age %d)", threshold_age),
           hjust = 0, size = 3.5) +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728"),
    name = "Scenario"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Age-at-Death Distribution",
    subtitle = sprintf("Cohort born %d — %s, SSP%d, Adaptation %.0f%%",
                       cohort_birth_year, city, scenario, adaptation_level * 100),
    x = "Age at Death",
    y = "Proportion of Deaths",
    caption = "Dashed line: threshold age (deaths below increase LI; above decrease LI)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

### 14.4.2 Excess Deaths by Age

```{r plot-excess-deaths-by-age, fig.width=10, fig.height=6}
# Highlight threshold region
ggplot(lifetable, aes(x = age, y = excess_deaths)) +
  geom_rect(aes(xmin = -Inf, xmax = threshold_age, ymin = -Inf, ymax = Inf),
            fill = "#fee5d9", alpha = 0.3) +
  geom_rect(aes(xmin = threshold_age, xmax = Inf, ymin = -Inf, ymax = Inf),
            fill = "#deebf7", alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  geom_line(color = "#d62728", linewidth = 1) +
  geom_vline(xintercept = threshold_age, linetype = "dashed", 
             color = "black", linewidth = 1) +
  annotate("text", x = threshold_age - 5, y = max(lifetable$excess_deaths) * 0.8,
           label = "Below threshold\n(increases LI)", hjust = 1, size = 3) +
  annotate("text", x = threshold_age + 5, y = max(lifetable$excess_deaths) * 0.8,
           label = "Above threshold\n(decreases LI)", hjust = 0, size = 3) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Excess Deaths by Age Due to Climate Change",
    subtitle = sprintf("Threshold age: %d (where remaining LE ≈ LE at age %d)",
                       threshold_age, cohort_start_age),
    x = "Age",
    y = "Excess Deaths (Climate - Baseline)",
    caption = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                      city, scenario, adaptation_level * 100)
  ) +
  theme_minimal(base_size = 12)
```

**Interpretation**: This plot shows the **age pattern** of climate impacts, revealing whether climate predominantly affects "younger old" or "older old" populations.

---

### 14.4.3 Lifespan Inequality Metrics Summary

```{r plot-li-metrics, fig.width=8, fig.height=6}
# Create summary data frame
li_metrics <- data.table(
  Metric = c("Gini", "Gini", "SD", "SD", "IQR", "IQR"),
  Scenario = rep(c("Baseline", "Climate"), 3),
  Value = c(gini_base, gini_climate, 
            sd_base, sd_climate,
            iqr_base$iqr, iqr_climate$iqr)
)

# Normalize for visualization (within each metric)
li_metrics[, Value_norm := Value / Value[1], by = Metric]

ggplot(li_metrics, aes(x = Scenario, y = Value, fill = Scenario)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = sprintf("%.3f", Value)), 
            vjust = -0.5, size = 3.5) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(
    values = c("Baseline" = "#1f77b4", "Climate" = "#d62728")
  ) +
  labs(
    title = "Lifespan Inequality Metrics: Baseline vs Climate",
    subtitle = sprintf("%s, SSP%d, Adaptation %.0f%%", 
                       city, scenario, adaptation_level * 100),
    y = "Metric Value",
    x = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 14.5 Summary Table

```{r li-summary-table}
# Create comprehensive summary
li_summary <- data.table(
  Metric = c("Life Expectancy at 20", "Gini Coefficient", 
             "SD of Age at Death", "IQR of Age at Death",
             "Threshold Age", "Excess Deaths Below Threshold",
             "Excess Deaths Above Threshold"),
  Baseline = c(
    sprintf("%.2f years", e20_base),
    sprintf("%.4f", gini_base),
    sprintf("%.2f years", sd_base),
    sprintf("%.1f years", iqr_base$iqr),
    sprintf("%d years", threshold_age),
    "—",
    "—"
  ),
  Climate = c(
    sprintf("%.2f years", e20_climate),
    sprintf("%.4f", gini_climate),
    sprintf("%.2f years", sd_climate),
    sprintf("%.1f years", iqr_climate$iqr),
    "—",
    sprintf("%.0f (%.1f%%)", excess_summary$excess_below, pct_below),
    sprintf("%.0f (%.1f%%)", excess_summary$excess_above, pct_above)
  ),
  Change = c(
    sprintf("%.2f years (%.2f%%)", delta_e20, pct_delta_e20),
    sprintf("%.4f (%.2f%%)", delta_gini, pct_change_gini),
    sprintf("%.2f years (%.2f%%)", delta_sd, pct_change_sd),
    sprintf("%.1f years", delta_iqr),
    "—",
    "—",
    "—"
  )
)

cat("\n=== Lifespan Inequality Summary ===\n")
print(li_summary)
```

---

## 14.6 Export Lifespan Inequality Results

```{r export-li-results}
# Create output directory if needed
if (!dir.exists("results")) dir.create("results", recursive = TRUE)

# Save detailed metrics
li_results <- data.table(
  city = city,
  scenario = sprintf("SSP%d", scenario),
  adaptation = adaptation_level,
  cohort_birth_year = cohort_birth_year,
  # Life expectancy
  e20_baseline = e20_base,
  e20_climate = e20_climate,
  delta_e20 = delta_e20,
  pct_change_e20 = pct_delta_e20,
  # Gini
  gini_baseline = gini_base,
  gini_climate = gini_climate,
  delta_gini = delta_gini,
  pct_change_gini = pct_change_gini,
  # Standard deviation
  sd_baseline = sd_base,
  sd_climate = sd_climate,
  delta_sd = delta_sd,
  pct_change_sd = pct_change_sd,
  # IQR
  iqr_baseline = iqr_base$iqr,
  iqr_climate = iqr_climate$iqr,
  delta_iqr = delta_iqr,
  # Threshold
  threshold_age = threshold_age,
  # Excess deaths classification
  total_excess_deaths = excess_summary$total_excess,
  excess_deaths_below_threshold = excess_summary$excess_below,
  excess_deaths_above_threshold = excess_summary$excess_above,
  pct_excess_below = pct_below,
  pct_excess_above = pct_above
)

# Save
li_path <- sprintf("results/lifespan_inequality_%s_ssp%d_cohort%d.csv",
                   tolower(gsub(" ", "_", city)),
                   scenario,
                   cohort_birth_year)

fwrite(li_results, li_path)
cat(sprintf("\nSaved: %s\n", li_path))

# Also save age-specific excess deaths for detailed analysis
excess_by_age <- lifetable[, .(
  age, year,
  dx_baseline = dx_base,
  dx_climate = dx_climate,
  excess_deaths,
  below_threshold,
  multiplier
)]

excess_path <- sprintf("results/excess_deaths_by_age_%s_ssp%d.csv",
                       tolower(gsub(" ", "_", city)),
                       scenario)

fwrite(excess_by_age, excess_path)
cat(sprintf("Saved: %s\n", excess_path))
```

---

## 14.7 Key Findings Interpretation

```{r interpretation}
cat("\n=== KEY FINDINGS ===\n\n")

cat(sprintf("For the cohort born in %d in %s under SSP%d:\n\n", 
            cohort_birth_year, city, scenario))

# Life expectancy
cat("1. LIFE EXPECTANCY\n")
if (delta_e20 < 0) {
  cat(sprintf("   → Climate change REDUCES life expectancy by %.2f years (%.2f%%)\n",
              abs(delta_e20), abs(pct_delta_e20)))
  cat(sprintf("   → Equivalent to %.1f months lost\n", abs(delta_e20) * 12))
} else {
  cat(sprintf("   → Climate change INCREASES life expectancy by %.2f years (%.2f%%)\n",
              delta_e20, pct_delta_e20))
  cat("   → (Likely due to reduced cold-related mortality)\n")
}

# Lifespan inequality
cat("\n2. LIFESPAN INEQUALITY\n")
if (delta_gini > 0) {
  cat(sprintf("   → Climate change INCREASES inequality (Gini: +%.4f, +%.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE unequally distributed\n")
} else {
  cat(sprintf("   → Climate change DECREASES inequality (Gini: %.4f, %.1f%%)\n",
              delta_gini, pct_change_gini))
  cat("   → Ages at death become MORE equally distributed\n")
}

cat(sprintf("   → Standard deviation changes by %.2f years\n", delta_sd))
cat(sprintf("   → IQR changes by %.1f years\n", delta_iqr))

# Mechanism
cat("\n3. MECHANISM\n")
cat(sprintf("   → Threshold age: %d years\n", threshold_age))
cat(sprintf("   → %.1f%% of excess deaths occur BELOW threshold\n", pct_below))
cat(sprintf("   → %.1f%% of excess deaths occur ABOVE threshold\n", pct_above))

if (pct_below > 60) {
  cat("   → Climate predominantly kills 'younger old' people\n")
  cat("   → This is the PRIMARY driver of increased inequality\n")
} else if (pct_above > 60) {
  cat("   → Climate predominantly kills 'older old' people\n")
  cat("   → This compresses the age-at-death distribution\n")
} else {
  cat("   → Climate impacts are distributed across ages\n")
  cat("   → Net inequality effect depends on relative magnitudes\n")
}

# Policy implication
cat("\n4. POLICY IMPLICATION\n")
if (delta_gini > 0) {
  cat("   → Without adaptation, climate change increases health inequity\n")
  cat("   → Priority: protect vulnerable 'younger old' populations\n")
} else {
  cat("   → Climate has a compression effect on lifespans\n")
  cat("   → But mean LE still affected—focus on absolute longevity\n")
}

cat(sprintf("\n   → With %.0f%% adaptation by 2100, these impacts are partially mitigated\n",
            adaptation_level * 100))
```

---

## Next Steps for Full Pipeline

1. **Multiple scenarios**: Run for all SSPs (1, 2, 3, 5) and adaptation levels (0%, 25%, 50%, 75%, 100%)
2. **Multiple cities**: Extend to 10-20 European cities (Vienna, Madrid, Stockholm, Athens, etc.)
3. **Geographic decomposition**: Compute within-city vs between-city inequality components
4. **Temporal trends**: Track how LI evolves year-by-year (not just single cohort)
5. **Heat vs cold decomposition**: Separate LI impacts of heat vs cold mortality

This single-city demo provides the **methodological foundation** for the full multi-scenario, multi-city analysis in Simon's Paper 1.

---

# Session Information

```{r session-info-lifetable}
sessionInfo()
```