---
title: "Heat-Related Mortality and Lifespan Inequality"
subtitle: "Climate Projection Analysis for European Cities"
author: "Simon Project"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This notebook estimates the impact of future temperature on mortality and lifespan inequality for a selected European city under different climate scenarios. The analysis integrates:

1. **Exposure-Response Functions (ERFs)** from Masselot et al., which quantify the relationship between temperature and mortality risk by age group
2. **Climate projections** from multiple General Circulation Models (GCMs) under different Shared Socioeconomic Pathways (SSPs)
3. **Adaptation scenarios** that model potential reductions in heat vulnerability over time

The core output is age-specific relative risk (RR) of mortality for each day and climate model, which can then be used to estimate changes in life expectancy and lifespan inequality.

---

# 1. Configuration

We begin by selecting the analysis parameters:

- **City**: The urban area of interest
- **Scenario**: SSP pathway (1 = SSP1-2.6, 2 = SSP2-4.5, 3 = SSP3-7.0, 5 = SSP5-8.5)
- **Adaptation level**: The fraction of heat-related excess risk eliminated by 2100 (0 = no adaptation, 1 = complete adaptation)

```{r setup}
city <- "Vienna"
city_code <- "AT001C"
scenario <- 2  # SSP2-4.5
adaptation_level <- 0.5  # 50% risk reduction by 2100

cat(sprintf("Analysis Configuration:\n"))
cat(sprintf("  City: %s (%s)\n", city, city_code))
cat(sprintf("  Scenario: SSP%d\n", scenario))
cat(sprintf("  Adaptation by 2100: %.0f%%\n", adaptation_level * 100))
```

---

# 2. Data Loading

## 2.1 Exposure-Response Functions (ERFs)

We load city-specific coefficients from Masselot et al. that define the temperature-mortality relationship. These coefficients parameterize a distributed lag non-linear model (DLNM) with a B-spline basis.

The ERF model is:

$$
\log RR(T) = \sum_{j=1}^{5} b_j \cdot s_j(T)
$$

where $$s_j(T)$$ are B-spline basis functions and $$b_j$$ are fitted coefficients.

```{r load-coefficients, message=FALSE}
source("../R/load_coefficients.R")

coefs <- load_city_coefficients(city_code, coefs_path = "../data/coefs.csv")

# Display city-specific coefficients by age group
head(coefs$city)
```

---

## 2.2 Single-Age Coefficient Interpolation

The original ERF coefficients are grouped into broad age categories:

- 20–44 years
- 45–64 years
- 65–74 years
- 75–84 years
- 85+ years

For fine-grained life table analysis, we interpolate these to **single-year ages** (20–100).

```{r interpolate-ages}
agelabs       <- c("20-44", "45-64", "65-74", "75-84", "85+")
age_midpoints <- c(32.5, 55, 70, 80, 92.5)
age_range     <- 20:100

coefs_single_age <- interpolate_coefs_to_single_age(
  coefs$city, coefs$all,
  agelabs, age_midpoints,
  age_range = age_range,
  city_code = city_code
)

head(coefs_single_age)
```

We now have ERF coefficients for **81 single-year ages** (20–100).

---

## 2.3 Temperature Projections

We load daily temperature projections from multiple GCMs for the selected city and scenario. The data covers:

- Historical period (for calibration)
- Future projections (2015–2100) under SSP scenarios

```{r load-temperature, message=FALSE, warning=FALSE}
source("../R/load_data.R")

parquet_path <- "../data/tmeanproj.gz.parquet"

# Exclude problematic GCMs
gcmexcl <- c("CMCC_CM2_SR5", "TaiESM1")

temp_data <- load_projected_temperatures(
  city_code,
  parquet_path = parquet_path,
  gcmexcl = gcmexcl
)

proj_data <- temp_data$proj_data
gcm_cols  <- temp_data$gcm_cols

cat(sprintf("\nLoaded %d days of temperature data from %d GCMs\n",
            nrow(proj_data), length(gcm_cols)))

head(proj_data)
```

**Data structure**: Each row is one day; columns `tas_*` contain daily mean temperature (°C) from different GCMs.

---

# 3. Data Preparation

## 3.1 Reshape Temperature Data

We convert the wide-format GCM columns into a long format for easier manipulation:

```{r reshape-temperature}
library(data.table)
setDT(proj_data)

temp_long <- proj_data[ssp == scenario]

temp_long <- melt(
  temp_long,
  id.vars = c("URAU_CODE", "date", "year", "doy", "ssp"),
  measure.vars = gcm_cols,
  variable.name = "gcm",
  value.name = "temp"
)

cat(sprintf("Reshaped data: %d observations (days × GCMs)\n", nrow(temp_long)))
head(temp_long)
```

---

## 3.2 Cross-Join with Age Groups

To compute age-specific RR for each day, we create a full panel dataset by cross-joining:

- **Days × GCMs** (temperature observations)
- **Ages** (ERF coefficients)

```{r cross-join-ages}
setDT(coefs_single_age)

# Extract coefficient columns
age_coefs <- coefs_single_age[, .(agegroup, b1, b2, b3, b4, b5)]

# Cartesian join
temp_long[, key := 1]
age_coefs[, key := 1]

dt <- merge(temp_long, age_coefs, by = "key", allow.cartesian = TRUE)
dt[, key := NULL]

cat(sprintf("Full panel: %d observations (days × GCMs × ages)\n", nrow(dt)))
cat(sprintf("Memory size: %.1f MB\n", object.size(dt) / 1024^2))

head(dt)
```

**Warning**: This creates a large dataset (~50 million rows for long time series). Consider aggregating or filtering if memory is limited.

---

# 4. Temperature-Mortality Model Construction

## 4.1 Define B-Spline Basis

We reconstruct the temperature basis function using the same specification as Masselot et al.:

- **Basis**: B-spline (degree 2)
- **Knots**: Placed at the 10th, 75th, and 90th percentiles of historical temperatures
- **Boundary**: Full historical temperature range

This ensures consistency with the original ERF estimation.

```{r define-basis}
library(dlnm)
library(splines)

varfun    <- "bs"
vardegree <- 2
varper    <- c(10, 75, 90)

# Extract historical temperatures for basis calibration
hist_temps <- unlist(
  proj_data[ssp == "hist", ..gcm_cols],
  use.names = FALSE
)
hist_temps <- hist_temps[!is.na(hist_temps)]

# Define knots and boundaries
varknots <- quantile(hist_temps, varper / 100, na.rm = TRUE)
varbound <- range(hist_temps, na.rm = TRUE)

cat(sprintf("Temperature basis:\n"))
cat(sprintf("  Knots: %.1f°C, %.1f°C, %.1f°C\n", varknots[1], varknots[2], varknots[3]))
cat(sprintf("  Boundary: [%.1f°C, %.1f°C]\n", varbound[1], varbound[2]))

# Store basis arguments
argvar <- list(
  fun = varfun,
  degree = vardegree,
  knots = varknots,
  Bound = varbound
)
```

---

## 4.2 Compute Temperature Basis

We evaluate the B-spline basis at each observed temperature:

$$
\mathbf{B}(T) = [s_1(T), s_2(T), s_3(T), s_4(T), s_5(T)]
$$

```{r compute-basis}
basis <- do.call(onebasis, c(list(x = dt$temp), argvar))

cat(sprintf("Basis matrix: %d observations × %d spline functions\n",
            nrow(basis), ncol(basis)))
```

---

## 4.3 Compute Log-Relative Risk

For each observation (day × age × GCM), we compute:

$$
\log RR = \sum_{j=1}^{5} b_j \cdot s_j(T) = \mathbf{B}(T) \cdot \mathbf{b}
$$

```{r compute-log-rr}
coef_mat <- as.matrix(dt[, .(b1, b2, b3, b4, b5)])

log_rr <- rowSums(basis * coef_mat)
dt[, log_rr := log_rr]

cat(sprintf("Computed log(RR) for %d observations\n", nrow(dt)))
```

---

## 4.4 Identify Minimum Mortality Temperature (MMT)

The **Minimum Mortality Temperature** is the temperature at which mortality risk is lowest. We compute this separately for each age group by:

1. Evaluating log(RR) across a fine temperature grid
2. Finding the minimum within a plausible range (25th to 99th percentile)

$$
\text{MMT}_a = \arg\min_{T \in [T_{0.25}, T_{0.99}]} \log RR_a(T)
$$

```{r compute-mmt}
temp_seq <- seq(varbound[1], varbound[2], by = 0.1)
basis_seq <- do.call(onebasis, c(list(x = temp_seq), argvar))

mmt_by_age <- dt[, {
  # Get unique coefficients for this age
  coefs <- unique(.SD[, .(b1, b2, b3, b4, b5)])
  coefs_vec <- as.numeric(coefs[1, ])
  
  # Evaluate log(RR) across temperature sequence
  log_rr_seq <- basis_seq %*% coefs_vec
  
  # Restrict to plausible range
  ind <- temp_seq >= quantile(temp_seq, 0.25) &
         temp_seq <= quantile(temp_seq, 0.99)
  
  # Find MMT
  mmt <- temp_seq[ind][which.min(log_rr_seq[ind])]
  
  .(mmt = mmt)
}, by = agegroup]

# Merge back into main dataset
dt <- merge(dt, mmt_by_age, by = "agegroup")

cat("MMT by age (first 10):\n")
print(head(mmt_by_age, 10))
```

**Interpretation**: MMT typically increases with age (older adults are more vulnerable to cold and less to heat).

---

## 4.5 Center at MMT and Compute Relative Risk

To ensure RR = 1.0 at the MMT (no excess risk), we center the log-RR:

$$
\log RR_{\text{centered}} = \mathbf{B}(T) \cdot \mathbf{b} - \mathbf{B}(\text{MMT}) \cdot \mathbf{b}
$$

Then exponentiate and enforce RR ≥ 1:

$$
RR = \max(e^{\log RR_{\text{centered}}}, 1)
$$

```{r compute-rr}
basis_mmt <- do.call(onebasis, c(list(x = dt$mmt), argvar))

log_rr_centered <- rowSums((basis - basis_mmt) * coef_mat)

dt[, RR := pmax(exp(log_rr_centered), 1)]

cat(sprintf("Mean RR across all observations: %.3f\n", mean(dt$RR)))
cat(sprintf("RR range: [%.3f, %.3f]\n", min(dt$RR), max(dt$RR)))
```

---

# 5. Adaptation Scenario

We model **gradual adaptation** as a linear reduction in heat vulnerability from 2030 to 2100. The adaptation factor represents the fraction of excess risk that is eliminated:

$$
\text{Adaptation}(y) = \begin{cases}
0 & \text{if } y < 2030 \\
\alpha \cdot \frac{y - 2030}{2100 - 2030} & \text{if } 2030 \le y \le 2100 \\
\alpha & \text{if } y > 2100
\end{cases}
$$

where $$\alpha$$ is the `adaptation_level` (e.g., 0.5 = 50% reduction).

The adapted RR is:

$$
RR_{\text{adapted}} = 1 + (RR - 1) \times (1 - \text{Adaptation})
$$

```{r apply-adaptation}
dt[, adapt_factor := fifelse(
  year < 2030, 0,
  fifelse(year > 2100, adaptation_level,
          adaptation_level * (year - 2030) / (2100 - 2030))
)]

dt[, RR_adapted := 1 + (RR - 1) * (1 - adapt_factor)]

cat(sprintf("Adaptation factor in 2030: %.2f\n", unique(dt[year == 2030, adapt_factor])))
cat(sprintf("Adaptation factor in 2065: %.2f\n", unique(dt[year == 2065, adapt_factor])))
cat(sprintf("Adaptation factor in 2100: %.2f\n", unique(dt[year == 2100, adapt_factor])))
```

---

# 6. Results: Annual Relative Risk by Age

To reduce the dataset size and facilitate downstream analysis, we aggregate daily RR to **annual means** by age and GCM:

$$
\overline{RR}_{a,y,g} = \frac{1}{365} \sum_{d=1}^{365} RR_{a,d,y,g}
$$

This represents the average mortality multiplier for each age, year, and climate model.

```{r aggregate-annual}
dt_annual <- dt[, .(
  RR_mean = mean(RR_adapted)
), by = .(year, agegroup, gcm)]

cat(sprintf("Annual dataset: %d observations (years × ages × GCMs)\n", nrow(dt_annual)))
cat(sprintf("Memory size: %.1f MB\n", object.size(dt_annual) / 1024^2))

head(dt_annual, 20)
```

---

# 7. Summary Statistics

```{r summary-stats}
# Overall statistics
cat("Overall Relative Risk Statistics:\n")
cat(sprintf("  Mean RR (adapted): %.4f\n", mean(dt_annual$RR_mean)))
cat(sprintf("  Median RR (adapted): %.4f\n", median(dt_annual$RR_mean)))
cat(sprintf("  Max RR (adapted): %.4f\n", max(dt_annual$RR_mean)))

# By age group (aggregated across years and GCMs)
cat("\nMean RR by Age (sample):\n")
age_summary <- dt_annual[, .(mean_RR = mean(RR_mean)), by = agegroup][order(agegroup)]
print(head(age_summary, 15))

# By year (aggregated across ages and GCMs)
cat("\nMean RR by Year (selected years):\n")
year_summary <- dt_annual[year %in% c(2020, 2040, 2060, 2080, 2100),
                          .(mean_RR = mean(RR_mean)), by = year]
print(year_summary)
```

---

# Next Steps

This analysis produces age-specific relative risks for each year and climate model. To complete the lifespan inequality analysis, the following steps are required:

1. **Compute Attributable Fraction (AF)**:
   $$AF = \frac{RR - 1}{RR}$$

2. **Estimate excess deaths** by applying AF to baseline age-specific mortality rates

3. **Construct modified life tables** incorporating heat-attributable mortality

4. **Calculate life expectancy** ($$e_0$$) and lifespan inequality measures:
   - Gini coefficient
   - Standard deviation of age at death
   - Lifespan disparity ($$e^\dagger$$)

5. **Compare scenarios** (SSP pathways, adaptation levels)

---

# Session Information

```{r session-info}
sessionInfo()
```
