# Simon Life Span Inequality
# =============================

Choose a city, an scenario and an adaptation level for year 2100 (we assume the adaptation rises linearly from 2030 to 2100).

```{r}
city <- "Vienna"
city_code <- "AT001C"
scenario <- 2
adaptation_level <- 0.5
```

## Exposure-Response Functions (Relative Risk Curves)

Load Masselot's ERF for the chosen city.

```{r}
# import function load_city_coefficients from R/load_coefficients.R
source("../R/load_coefficients.R")

coefs <-load_city_coefficients(city_code, coefs_path="../data/coefs.csv")
# contains $city and $all

# Display the first few rows of the city coefficients
head(coefs$city)
```

Original Masselot coefficients are grouped by age. We need to expand them to single ages.

```{r}
agelabs      <- c("20-44", "45-64", "65-74", "75-84", "85+")
age_midpoints <- c(32.5, 55, 70, 80, 92.5)
age_range     <- 20:100

coefs_single_age <- interpolate_coefs_to_single_age(
    coefs$city, coefs$all,
	agelabs, age_midpoints,
	age_range = age_range,
	city_code = city_code
)

# Display the first few rows of the single-age coefficients
head(coefs_single_age)
```

## Temperature Projections

Load the projected temperature data for the city, excluding certain GCMs if needed.

```{r}
source("../R/load_data.R")

parquet_path <- "../data/tmeanproj.gz.parquet"

# GCM exclusions (already removed at prep stage for partitioned data,
# but kept here for fallback / local dev with full parquet)
gcmexcl <- c("CMCC_CM2_SR5", "TaiESM1")

temp_data <- load_projected_temperatures(
    city_code,
    parquet_path = parquet_path,
    gcmexcl = gcmexcl)

proj_data <- temp_data$proj_data
gcm_cols  <- temp_data$gcm_cols

# Display the first few rows of the temperature data
head(proj_data)
```

Currently we have wide GCM columns. Convert to long:


```{r}
library(data.table)

temp_long <- proj_data[ssp == scenario]

temp_long <- melt(
  temp_long,
  id.vars = c("URAU_CODE", "date", "year", "doy", "ssp"),
  measure.vars = gcm_cols,
  variable.name = "gcm",
  value.name = "temp"
)

# Display the first few rows of the long-format temperature data
head(temp_long)
```

Cross-join with ages.

```{r}
setDT(coefs_single_age)

# Keep only needed columns
age_coefs <- coefs_single_age[, .(agegroup, b1, b2, b3, b4, b5)]

# Cross join
temp_long[, key := 1]
age_coefs[, key := 1]

dt <- merge(temp_long, age_coefs, by = "key", allow.cartesian = TRUE)
dt[, key := NULL]

# Display the first few rows of the combined data
head(dt)
```

Recreate the temperature basis.

```{r}
library(dlnm)
library(splines)

varfun    <- "bs"
vardegree <- 2
varper    <- c(10, 75, 90)

# Use historical temps to define knots
hist_temps <- unlist(
  proj_data[ssp == "hist", ..gcm_cols],
  use.names = FALSE
)
hist_temps <- hist_temps[!is.na(hist_temps)]

varknots <- quantile(hist_temps, varper / 100, na.rm = TRUE)
varbound <- range(hist_temps, na.rm = TRUE)

argvar <- list(fun = varfun,
               degree = vardegree,
               knots = varknots,
               Bound = varbound)

# Build basis for each observed temperature
basis <- do.call(onebasis, c(list(x = dt$temp), argvar))
```

Compute log(RR).

```{r}
coef_mat <- as.matrix(dt[, .(b1, b2, b3, b4, b5)])

log_rr <- rowSums(basis * coef_mat)
dt[, log_rr := log_rr]
```

Find MMT per age (for centering).

```{r}
temp_seq <- seq(varbound[1], varbound[2], by = 0.1)
basis_seq <- do.call(onebasis, c(list(x = temp_seq), argvar))

mmt_by_age <- dt[, {
  coefs <- unique(.SD[, .(b1,b2,b3,b4,b5)])
  coefs_vec <- as.numeric(coefs[1,])
  
  log_rr_seq <- basis_seq %*% coefs_vec
  
  ind <- temp_seq >= quantile(temp_seq, 0.25) &
         temp_seq <= quantile(temp_seq, 0.99)
  
  mmt <- temp_seq[ind][which.min(log_rr_seq[ind])]
  
  .(mmt = mmt)
}, by = agegroup]

# Merge back
dt <- merge(dt, mmt_by_age, by = "agegroup")
```

Center at MMT.

```{r}
basis_mmt <- do.call(onebasis, c(list(x = dt$mmt), argvar))

log_rr_centered <- rowSums((basis - basis_mmt) * coef_mat)

dt[, RR := pmax(exp(log_rr_centered), 1)]
```

Apply adaptation.

```{r}
dt[, adapt_factor :=
      fifelse(year < 2030, 0,
      fifelse(year > 2100, adaptation_level,
              adaptation_level * (year - 2030) / (2100 - 2030)))]

dt[, RR_adapted := 1 + (RR - 1) * (1 - adapt_factor)]
```

Average annual impacts.

```{r}
dt_annual <- dt[, .(
  RR_mean = mean(RR_adapted)
), by = .(year, agegroup, gcm)]

# Display the first few rows of the annual impacts
head(dt_annual)
```